### 2021-12-28 LaTeX/tarjan.tex
Slides for presentation + tikz fig

\documentclass[a4 paper, 12pt]{article}
\usepackage{amsmath, amsthm, amsfonts, amssymb, mathrsfs}
\usepackage{tikzit}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}
\usepackage[ruled,vlined, linesnumbered]{algorithm2e}

\input{style.tikzstyles}

%__________________________________________

\hypersetup{
    colorlinks=true,
    linktoc=false,
    linkcolor=blue
}

%__________________________________________
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,datatype,fun,abbreviation,definition,proof,lemma,theorem,corollary},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=fixed,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}

%__________________________________________

\def\NN{$\mathcal{N}~$}
\def\GG{$\mathcal{G}~$}
\def\VV{$\mathcal{V}~$}
\def\EE{$\mathcal{E}~$}

\renewcommand\qedsymbol{$\blacksquare$}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}[lemma]
\newtheorem{remark}{Remark}

%__________________________________________

% Node styles
\tikzstyle{white node}=[fill=none, draw=black, shape=circle, tikzit fill=white, tikzit draw=black]
\tikzstyle{red node}=[fill={rgb,255: red,232; green,118; blue,188}, draw=black, shape=circle, tikzit fill={rgb,255: red,232; green,118; blue,188}, tikzit draw=black]
\tikzstyle{blue node}=[fill={rgb,255: red,179; green,191; blue,255}, draw=black, shape=circle, tikzit fill={rgb,255: red,179; green,191; blue,255}, tikzit draw=black]
\tikzstyle{green node}=[fill={rgb,255: red,201; green,255; blue,189}, draw=black, shape=circle, tikzit fill={rgb,255: red,201; green,255; blue,189}, tikzit draw=black]

\tikzstyle{arrow}=[>]
\tikzstyle{new style 0}=[fill=none, draw=red, shape=circle, tikzit fill=none]
\tikzstyle{arrow}=[<--]
\tikzstyle{new edge style 2}=[- - -]

% Edge styles
\tikzstyle{right arrow}=[->]
\tikzstyle{right red arrow}=[->, draw=red, tikzit draw=red]
\tikzstyle{left arrow}=[<-]
\tikzstyle{left red arrow}=[<-, draw=red, tikzit draw=red]
\tikzstyle{right dashed arrow}=[->, dashed=1, dash pattern=on 1mm off 1mm, draw=red]
\tikzstyle{left dashed arrow}=[<-, dashed=1, dash pattern=on 1mm off 1mm, draw=red]
\tikzstyle{blue fill}=[-, fill={rgb,255: red,179; green,191; blue,255}, draw=none, tikzit draw=none, tikzit fill={rgb,255: red,213; green,207; blue,255}]
\tikzstyle{green fill}=[-, fill={rgb,255: red,201; green,255; blue,189}, draw=none, tikzit draw=none, tikzit fill={rgb,255: red,213; green,255; blue,208}]
\tikzstyle{red fill}=[-, fill={rgb,255: red,232; green,118; blue,188}, draw=none, tikzit draw=none, tikzit fill={rgb,255: red,232; green,118; blue,118}]

%-----------------------------------------------------------------

\begin{document}
\newgeometry{margin=2cm}
\begin{titlepage}
    \begin{figure}[!h]
        \centering
        \includegraphics[height = .1\textwidth]{/Users/vincent/Documents/MinesNancy/logoartem.png}
        \hspace{1cm}
        \includegraphics[height = .1\textwidth]{img/logoloria.jpg}
        \hspace{1cm}
        \includegraphics[height = .1\textwidth]{img/logoUL.png}
    \end{figure}
    \vspace{3cm}

    \begin{center}
%        \huge{Strongly connected components algorithms : parallelization and proofs}
        \huge{Formal methods and assisted proofs: application to strongly connected components algorithms}
    \end{center}
    \vspace{1cm}
    \begin{center}
        \large{Département Informatique --- Parcours Recherche \\ Tuteur : Stephan Merz}
    \end{center}
    \vspace{1 cm}
    \begin{center}
        Vincent TRÉLAT
    \end{center}
    \vspace{2 cm}
    \begin{center}
        \textit{\today}
    \end{center}
    \vspace{2 cm}
    \begin{center}
        ***
    \end{center}

\end{titlepage}
\restoregeometry
\tableofcontents
\pagebreak

\section{Models and representation}
\subsection{Nodes}
% Je pense que cette description n'a de sens que dans le contexte d'un algorithme particulier. Dans le document final il faudra commencer par décrire le problème indépendamment de l'algorithme utilisé pour le résoudre, d'autant plus qu'il y a deux algorithmes différents.

Vertices of a graph can be represented as nodes.
A node \NN is a simple data structure composed of an index, a boolean value telling if it has already been visited and two integer values \textit{num} and \textit{lowlink} whose role will be explained later.\\
In the following, the aforesaid attributes will be referred to through the following notation:\\
Let \NN be a node. The attributes of \NN can be accessed via \texttt{N.index}, \texttt{N.visited}, \texttt{N.num} and \texttt{N.lowlink}. This notation will be applied to any object that lends itself to it.

\subsection{Graphs and their representation}
A graph \GG is the data $(\mathcal{V}, \mathcal{E})$ where:
\begin{itemize}
    \item $\mathcal{V}$ is a set of vertices
    \item $\mathcal{E} \subseteq \{(x,y) \in V^2\}$ is a set of edges\footnote{Note the use of the couple $(x, y)$ and not the pair $\{x, y\}$ that makes the graph directed.}.
\end{itemize}
Vertices will often be called nodes and edges will be represented through adjency lists for each node.\\
Let us give an example. Let \GG = $(\mathcal{V}, \mathcal{E})$ be the graph represented on figure \ref{fig:graph}.
\begin{figure}[!h]
    \ctikzfig{graph}
    \caption{Representation of \GG\label{fig:graph}}
\end{figure}
Thus, \VV = $\{0,1,2,3,4,5,6,7\}$ and
\begin{flushleft}
    ~~~~~~~~\EE = $\{(0,0),(0,1),(0,2),(0,3),$\\
    $~~~~~~~~~~~~~~~(1,4),(1,7),$\\
    $~~~~~~~~~~~~~~~(3,0),(3,1),(3,2),(3,5),$\\
    $~~~~~~~~~~~~~~~(4,3),(4,6),$\\
    $~~~~~~~~~~~~~~~(5,6),$\\
    $~~~~~~~~~~~~~~~(6,3),$\\
    $~~~~~~~~~~~~~~~(7,6)\}$
\end{flushleft}

This representation being somewhat long, adjency lists can be used instead and therefore it gives:
\begin{center}
    \texttt{G.adjency = [[0,1,2,3],[4,7],[0,1,2,5],[3,6],[6],[3],[6]]}
\end{center}
Thus, for all $i \in \{0,\ldots,7\}$, \texttt{G.adjency[i]} is the list of nodes to which node $i$ is connected -- \textit{i.e.} there is an directed edge from node $i$ to every node of \texttt{G.adjency[i]}.
\pagebreak

\section{Formalisation}
\subsection{Strongly connected components}
\subsubsection{Directed graphs}
\begin{definition}
    For two vertices $x$ and $y$ of \VV, the relation "has an edge to" is noted "$\Rightarrow$" such that
    \begin{equation*}
        (x, y) \in \mathcal{E} \Longleftrightarrow x \Rightarrow y
    \end{equation*}
    By extension, if there is a path from $x$ to $y$ with more than one edge, the same notation is kept for the sake of simplicity.
    The reflexive and transitive closure of the relation $\Rightarrow$ is noted $\Rightarrow^*$.
\end{definition}

\begin{definition}
    Let \GG = (\VV, \EE) be an directed graph.
$\mathcal{C} \subseteq \mathcal{V}$ is a strongly connected component of \GG if:
\begin{gather*}
    \forall x, y \in \mathcal{C}, (x \Rightarrow y) \wedge (y \Rightarrow x)
\end{gather*}
\textit{i.e.} there is a path between every $x$ and $y$ in $\mathcal{C}$.\\

\end{definition}

\subsubsection{Examples}
\begin{figure}[!h]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \ctikzfig{example2}
        \subcaption{Strongly connected graph}
    \end{subfigure}
    \begin{subfigure}[t]{.49\textwidth}
        \ctikzfig{example3}
        \subcaption{Not strongly connected graph}
    \end{subfigure}
    \caption{Basic example of what is a small SCC}
\end{figure}
\begin{figure}[!h]
    \ctikzfig{exampleSCC}
    \caption{Example of a graph where each colored set of node is a maximal set of SCC\label{fig:exampleSCC}}
\end{figure}

\begin{figure}[!h]
    \ctikzfig{exampleSCCreduced}
    \caption{Reduced visualization of the graph represented if figure \ref{fig:exampleSCC}}
\end{figure}

\subsection{Order of traversal and backtracking edges}
\subsubsection{DFS and \textit{num} value}
Tarjan's SCC algorithm basically lies on a depth-first search. The figure \ref{fig:dfs} shows an example of a DFS traversal on a simple directed graph.
\begin{figure}[!h]
    \ctikzfig{dfs}
    \caption{Example of a DFS\label{fig:dfs}}
\end{figure}

The previously mentioned figure also displays in red the \texttt{num} value which represents the order in which the nodes are visited in the graph during the DFS.

\subsubsection{Backtracking edges}
\begin{definition}
    Given a graph \GG and an order of traversal in this graph, \textit{i.e.} each node of \GG has a unique value $\texttt{num} \in [0,~|\mathcal{V}|]~$ and two nodes $u$ and $v$, there is a backtracking edge from $v$ to $u$ if:
    \begin{equation*}
        \left\{
            \begin{array}{l}
                \texttt{u.num} < \texttt{v.num}\\
                v~\leadsto~u,~\textit{i.e. }~(v, u) \in \mathcal{E}\\
            \end{array}
        \right.
    \end{equation*}
    In this case, the backtracking edge from $v$ to $u$ is represented by $v \hookrightarrow u$.\\      
\end{definition}

\subsection{Lowlink value}
\subsubsection{Definition}
Informally, the \texttt{lowlink} value of a node represents the \texttt{num} value of the attachment node of their SCC, \textit{i.e.} the \texttt{num} value of the entrance node in the corresponding SCC.\\

A more formal definition would be the following:
\begin{definition}
    Let $u$ be a node.
    \begin{equation*}
        \texttt{u.lowlink} = \min \{\texttt{w.num}~|~\exists~v\in \mathcal{V},~u \Rightarrow v \hookrightarrow w\}
    \end{equation*}
\end{definition}

\subsubsection{Example}
Let \GG be the graph given in fig. \ref{fig:exlowlink}. The order of traversal of the graph is given by the value \textit{num} for each node of \GG. The \textit{lowlink} value is also displayed.

\begin{figure}[!h]
    \ctikzfig{lowlink}
    \caption{A DFS was performed through \GG from $x_1$ and next to each node is represented the couple of value (\textit{num}, \textit{lowlink})\label{fig:exlowlink}}
\end{figure}

Now, backtracking edges can be highlighted w.r.t. the order of traversal. In fig. \ref{fig:exlowlink2}, they are represented as red dashed arrows.\\
Knowing the backtracking edges, all \textit{lowlink} values can be computed\footnote{In fact, they are refreshed during the DFS.}.\\
Let us take $x_8$ as an example: its \textit{lowlink} is equal to 2, which actually means that $x_6$\footnote{Because $\texttt{x6.num} = 2$} is its anchor -- or attachment node -- in their SCC, namely the green one. Indeed, $x_8$ is alone in its equivalence class\footnote{for the relation $\Rightarrow$}, and from all nodes linked by one of the backtracking edges of $x_8$, $x_6$ has the minimum value \textit{num}. Likewise, $x_3$ is its own attachment node in the green SCC since it is the first node visited when performing the DFS.

\begin{figure}[!h]
    \ctikzfig{lowlink2}
    \caption{Same graph as in fig. \ref{fig:exlowlink} whose backtracking edges have been represented with red dashed arrows and SCCs have been highlighted\label{fig:exlowlink2}}
\end{figure}

Then, SCCs can be easily found, namely $\{x_1\}$ and $\{x_2,x_3,x_4,x_5,x_6,x_7,x_8, x_9\}$, as shown in fig. \ref{fig:exlowlink2}.

\pagebreak
\section{Tarjan's algorithm}
Tarjan's algorithm is an efficient on-the-fly SCC computing algorithm. It basically perfoms a DFS while updating the \textit{num} and \textit{lowlink} values. All nodes are stored in a stack during the traversal until a backtracking edge is found. In this case, the \textit{lowlinks} are computed and all nodes are unstacked and saved in a SCC until a node verifying the equality between its \textit{num} and its \textit{lowlink} -- which has to occur -- is found. Then, the DFS goes on. The whole process is written in the following algorithm \ref{alg:tarjan}.\\
It can be shown that every node and edge are visited only once so the algorithm can achieve a linear complexity, \textit{i.e.} $\mathcal{O}(|\mathcal{V}|+|\mathcal{E}|)$.

\subsection{Description of Tarjan's algorithm}
\begin{algorithm}[H]\label{alg:tarjan}
    \SetAlgoLined
    \KwData{A graph \GG = (\VV, \EE)}
    \KwResult{A partition \texttt{SCCs} of \VV where each element of \texttt{SCCs} is a SCC of \GG}
    Initialize an empty stack \texttt{R}\;
    Initialize an empty set \texttt{SCCs}\;
    Let $\texttt{num} := 0$\;
    \ForAll{$v \in \mathcal{V}$}{
        \If{\texttt{v.num} is unefined}{
            SCC(v)\;
        }
    }
    \SetKwProg{Function}{function}{}{}
    \Function{SCC: $v \in \mathcal{V} \rightarrow \texttt{None}$}{
        \texttt{v.num = num}\;
        \texttt{v.lowlink = num}\;
        increment \texttt{num}\;
        Push $v$ in \texttt{R}\;
        \texttt{v.onStack = true}\;
        \ForAll{$w \in \text{POST}(v)$}{
            \If{\texttt{w.num} is undefined}{
                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.lowlink})$\;
            }
            \ElseIf{\texttt{w.onStack}}{
                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.num})$\;
            }
        }
        \If{\texttt{v.lowlink = v.num}}{
            Initialize an empty set \texttt{currentSCC}\;
            \Repeat{$\texttt{v.num} \neq \texttt{w.num}$}{
                Let $w := \texttt{R.pop()}$\;
                \texttt{w.onStack = false}\;
                $\texttt{currentSCC} = \texttt{currentSCC} \cup \{w\}$\;
            }
            $\texttt{SCCs} = \texttt{SCCs} \cup \texttt{currentSCC}$\;
        }
    }
    \caption{Tarjan's algorithm}
\end{algorithm}
\pagebreak

\section{A sequential set-based algorithm}
\subsection{Formalisation}
\begin{definition}[SCC mapping]
    In the following algorithm, the SCCs are progressively tracked in a collection of disjoint sets through a map $\mathcal{S} : \mathcal{V} \longrightarrow \mathcal{P}(\mathcal{V})$, where $\mathcal{P}(\mathcal{V})$ is the powerset of \VV, s.t. the following invariant is maintained:
    \begin{equation}\label{invariant}
        \forall v, w \in \mathcal{V}, w\in \mathcal{S}(v) \Longleftrightarrow \mathcal{S}(v) = \mathcal{S}(w)
    \end{equation}
\end{definition}

\begin{remark}
    In particular, $\forall v \in \mathcal{V}, v \in \mathcal{S}(v)$
\end{remark}

\begin{definition}[SCC union]
    Let \textsc{Unite} be the function taking as parameters a map $\mathcal{S}$ as defined previously and two vertices $u$ and $v$ of $\mathcal{V}$ such that $\textsc{Unite}(\mathcal{S}, u, v)$ merges the two mapped sets $\mathcal{S}(u)$ and $\mathcal{S}(v)$ and maintains the invariant (\ref{invariant}) by updating the function $\mathcal{S}$.
\end{definition}
Let us give an example:\\
Let $\mathcal{V} = \{u,v,w\}$ such that there is the following mapping: $\mathcal{S}(u) = \{u\}$ and $\mathcal{S}(v) = \mathcal{S}(w) = \{v,w\}$.\\
Then, $\textsc{Unite}(\mathcal{S}, u, v) = \mathcal{S}(u) = \mathcal{S}(v) = \mathcal{S}(w) = \{u,v,w\}$.

\begin{definition}[Successors set for a node]
    Let $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ and $v \in \mathcal{V}$. The set of successors of $v$ in \GG is \textsc{Post}($v$) such that:
    \begin{equation*}
        \forall w\in \textsc{Post}(v), (v,w)\in \mathcal{E}
    \end{equation*}
\end{definition}

\subsection{The algorithm}
\begin{algorithm}[H]\label{alg:seqsetbased}
    \SetAlgoLined
    \KwData{A graph \GG = (\VV, \EE), a starting node $v_0$\;}
    \KwResult{A partition \texttt{SCCs} of \VV where each element of \texttt{SCCs} is a maximal set of strongly connected components of \GG\;}
    Initialize an empty set \texttt{DEAD}\;
    Initialize an empty set \texttt{VISITED}\;
    Initialize an empty stack \texttt{R}\;
    setBased($v_0$)\;
    \SetKwProg{Function}{function}{}{}
    \Function{setBased: $v \in \mathcal{V} \rightarrow \texttt{None}$}{
        $\texttt{VISITED} := \texttt{VISITED} \cup \{v\}$\;
        \texttt{R.push($v$)}\;
        \ForEach{$w\in \texttt{POST(v)}$}{
            \If{$w\in \texttt{DEAD}$}{
                continue\;
            }
            \ElseIf{$w \notin \texttt{VISITED}$}{
                setBased($w$)\;
            }
            \Else{
                \While{$\mathcal{S}(v) \neq \mathcal{S}(w)$}{
                    $r := \texttt{R.pop()}$\;
                    $\texttt{UNITE}(\mathcal{S}, r, \texttt{R.top()})$\;
                }
            }
        }
        \If{$v = \texttt{R.top()}$}{
            \textbf{report SCC} $\mathcal{S}(v)$\;
            $\texttt{DEAD} := \texttt{DEAD} \cup \mathcal{S}(v)$\;
            \texttt{R.pop()}\;
        }
    }

    \caption{Sequential set-based SCC algorithm}
\end{algorithm}
\subsection{Correctness}
\begin{lemma}(First invariant)\label{lemma:disjointness}
    \begin{equation*}
        \forall x, y \in \texttt{R}, x\neq y \implies \mathcal{S}(x) \cap \mathcal{S}(y) = \varnothing
    \end{equation*}
    Note the misuse of the set notation $x, y \in \texttt{R}$ which just means that $x$ and $y$ are in the stack \texttt{R}.
\end{lemma}

\begin{proof}
    Let $x \in \mathcal{V}$ be the following node to be visited during the execution of the algorithm \ref{alg:seqsetbased}: $x$ is pushed in \texttt{R}. Let $y \in \textsc{Post}(x)$. There are two cases:\\
    \begin{itemize}
        \item y has not been visited yet, \textit{i.e.} $y \notin \textsc{Visited}$. Thus, a DFS-like traversal is performed from $y$, so $y$ is pushed in \texttt{R} and $\mathcal{S}(y) = \{y\}$ because $y$ is alone in its equivalence class for the moment since it has not been visited yet.\\
        Therefore, $\mathcal{S}(x) \cap \mathcal{S}(y) = \varnothing$.

        \item y has already been visited, \textit{i.e.} $y \in \textsc{Visited}$. Then, $y$ was already pushed in \texttt{R} before $x$. Let $(x_i)_{1\leq i \leq n}$ be the first nodes of the stack s.t. $x_0 = x$ and $x_n = y$.\\
        In order to avoid writing $\texttt{R} = [\ldots, y, \ldots, x]$, let us define $\widetilde{R}$ the stack containing the first $n$ nodes in \texttt{R}, s.t. $\widetilde{R} = [y, \ldots, x] = [x_n, \ldots, x_0]$.\\
        Let us consider the worst case, \textit{i.e.} when
        \begin{equation*}
            \forall~1 \leq i \leq n,~\mathcal{S}(x_i) = \{x_i\}
        \end{equation*}
        So, the while loop has to go down to $y$ because all partial SCCs are disjoint. As the length of the stack \texttt{R} is bounded by $|\mathcal{V}|$, the algorithm terminates.\\
        $x_0$ is first unstacked and both $\mathcal{S}(x_0)$ and $\mathcal{S}(\texttt{R.top()}) = \mathcal{S}(x_1)$ are then united. The current state of $\mathcal{S}$ and $\widetilde{R}$ is:
        \begin{equation*}
            \left\{
                \begin{array}{l}
                    \mathcal{S} = \{\{x_0, x_1\}, \{x_2\}, \ldots, \{x_n\}, \ldots\}\\
                    \widetilde{R} = [x_n, \ldots, x_1]\\
                \end{array}
            \right.
        \end{equation*}
        Then, $x_1$ is unstacked and $\mathcal{S}(x_1)$ and $\mathcal{S}(x_2)$ are then united, so that:
        \begin{equation*}
            \left\{
                    \begin{array}{l}
                        \mathcal{S} = \{\{x_0, x_1, x_2\}, \{x_3\}, \ldots, \{x_n\}, \ldots\}\\
                        \widetilde{R} = [x_n, \ldots, x_2]\\
                    \end{array}
                \right.
        \end{equation*}
        Finally (by induction), $\mathcal{S} = \{x_0, \ldots, x_n\}$ and $\widetilde{R} = [y]$, \textit{i.e.} $\mathcal{S}(x)=\mathcal{S}(y)$. It is important to notice that $x=x_0, x_1, \ldots, x_{n-1}$ are no longer in the stack, so this operation kept the invariant true.
    \end{itemize}
\end{proof}

\begin{lemma}
    \begin{equation*}
        \biguplus_{v \in \texttt{R}} \mathcal{S}(v) = \textsc{Live} := \textsc{Visited} \setminus \textsc{Dead}
    \end{equation*}
\end{lemma}

\begin{proof}
    The disjointness of all on-stack partial SCCs is given by lemma \ref{lemma:disjointness}. Nodes from $\textsc{Visited} \setminus \textsc{Dead}$ are in \texttt{R} because they are being processed. So, $\textsc{Live} \subseteq \texttt{R}$.\\
    By L.6-7 of algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.\\
    L.9-10 ensure that no dead node is pushed in \texttt{R}.\\
    L.24-25 keep the invariant by unstacking dead nodes from \texttt{R}, so $\texttt{R} \cap \textsc{Dead} = \varnothing$. Thus, $\texttt{R} = \textsc{Visited} \setminus \textsc{Dead} = \textsc{Live}$.
\end{proof}

\begin{corollary}\label{cor:cor1}
    \begin{equation*}
        \forall v \in \textsc{Live}, \exists!~r \in \texttt{R} \cap \mathcal{S}(v), \mathcal{S}(v) = \mathcal{S}(r)
    \end{equation*}
\end{corollary}
\begin{proof}
    Let $v \in \textsc{Live} = \displaystyle{\biguplus_{v \in \texttt{R}} \mathcal{S}(v)}$. $v$ is in a unique partial SCC $\mathscr{S} := \mathcal{S}(v)$. Because of lemma \ref{lemma:disjointness}, there cannot exist $x \neq y \in \texttt{R}$ s.t. $\mathcal{S}(x) = \mathcal{S}(y) = \mathscr{S}$. Thus, there exists a unique $x \in \texttt{R}$ s.t. $\mathcal{S}(x) = \mathscr{S}$ (and $x\in \texttt{R} \cap \mathscr{S}$).
\end{proof}

\begin{corollary}\label{cor:cor2}
    \begin{equation*}
        \forall v \in \mathcal{V}, \forall w \in \textsc{Post}(v), w \in \textsc{Live} \implies \exists w' \in \texttt{R}, \mathcal{S}(w') = \mathcal{S}(w)
    \end{equation*}
\end{corollary}

\begin{proof}
    Holds because of corollary \ref{cor:cor1}.
\end{proof}

\begin{remark}
    In the algorithm \ref{alg:seqsetbased}, this property is held by L.16-18. These lines also illustrate how the algorithm "reads" the SCCs. Corollary \ref{cor:cor2} shows that when the mapped representatives of the top two nodes of \texttt{R} are united (until $\mathcal{S}(w') = \mathcal{S}(v) = \mathcal{S}(w)$ since $w'$ has a path to $v$), then all united components are in the same SCC.
\end{remark}

\begin{remark}\label{rem:proof}
    Because \texttt{R} only contains exactly one representative for each partial SCC (corollary \ref{cor:cor1}), after each step of the main loop -- \textit{i.e.} the DFS -- every partial SCC is actually maximal in the current set of visited nodes.
\end{remark}

\begin{theorem}
    The sequential algorithm \ref{alg:seqsetbased} is correct, \textit{i.e.} it returns a set of maximal SCCs.
\end{theorem}
\begin{proof}
    Holds by remark \ref{rem:proof}.
\end{proof}

\pagebreak

\subsection{Isabelle proof}
cf Proof

%-----------------------------------------------------------------
\pagebreak

\begin{thebibliography}{99}

\bibitem{} R. Chen, C. Cohen, J.-J. Lévy, S. Merz, L. Théry, \emph{Formal Proofs of Tarjan’s Strongly Connected 2 Components Algorithm in Why3, Coq and Isabelle}, 2019

\bibitem{} V. Bloemen, A. Laarman, J. van de Pol, \emph{Multi-Core On-The-Fly SCC Decomposition}, 2016

\bibitem{} V. Bloemen, \emph{Strong Connectivity and Shortest Paths for Checking Models}, 2019

\end{thebibliography}

\end{document}

### 2022-01-07 LaTeX/tarjan.tex
Slides updates, figures added
Updates

@@
     linktoc=false,
-    linkcolor=blue
+    linkcolor=blue,
+    citecolor=blue
 }
@@
                 \texttt{u.num} < \texttt{v.num}\\
-                v~\leadsto~u,~\textit{i.e. }~(v, u) \in \mathcal{E}\\
+                v~\Rightarrow~u,~\textit{i.e. }~(v, u) \in \mathcal{E}\\
             \end{array}
@@
 \section{Tarjan's algorithm}
-Tarjan's algorithm is an efficient on-the-fly SCC computing algorithm.
+Tarjan's algorithm is an efficient on-the-fly SCC computing algorithm \cite{tarjanMerz}.
 It basically perfoms a DFS while updating the \textit{num} and \textit{lowlink} values.
@@
 \subsection{The algorithm}
+See \cite{articleSCC} for the original paper.\\
 \begin{algorithm}[H]\label{alg:seqsetbased}
@@
 \subsection{Isabelle proof}
-cf Proof
-
+Not finished yet! See my oral presentation on January 14, 2022.
 %-----------------------------------------------------------------
@@
 
-\bibitem{} R. Chen, C. Cohen, J.-J. Lévy, S. Merz, L. Théry, \emph{Formal Proofs of Tarjan’s Strongly Connected 2 Components Algorithm in Why3, Coq and Isabelle}, 2019
+\bibitem{tarjanMerz} R. Chen, C. Cohen, J.-J. Lévy, S. Merz, L. Théry, \emph{Formal Proofs of Tarjan’s Strongly Connected 2 Components Algorithm in Why3, Coq and Isabelle}, 2019
 
-\bibitem{} V. Bloemen, A. Laarman, J. van de Pol, \emph{Multi-Core On-The-Fly SCC Decomposition}, 2016
+\bibitem{thesisBloemen} V. Bloemen, A. Laarman, J. van de Pol, \emph{Multi-Core On-The-Fly SCC Decomposition}, 2016
 
-\bibitem{} V. Bloemen, \emph{Strong Connectivity and Shortest Paths for Checking Models}, 2019
+\bibitem{articleSCC} V. Bloemen, \emph{Strong Connectivity and Shortest Paths for Checking Models}, 2019
 

### 2022-02-25 LaTeX/report.tex
Updates

--- 2022-01-07 LaTeX/tarjan.tex
@@
 \tableofcontents
+\pagebreak
+
+\section{Introduction}
+\subsection{Formal methods}
+Formal methods are a field of computer science related to mathematical logic and reasoning.
+The whole purpose of the discipline is to ensure by a logical proof that a given algorithm is not only correct on its domain of definition, but also to find -- or define -- that domain.
+Formal methods find applications in a variety of fields, both concrete, such as the railway industry or self-driving cars, and abstract, such as computational architecture.\\
+Altough a formal proof lies first on paper, the real formalisation starts when proofs are mechanised in a proof assistant.
+
+\subsection{Isabelle (HOL)}
+\begin{quote}
+    Isabelle is a generic proof assistant. It allows mathematical formulas to be expressed in a formal language and provides tools for proving those formulas in a logical calculus.
+\end{quote}
+\begin{flushright}
+    \href{https://isabelle.in.tum.de/}{isabelle.in.tum.de}
+\end{flushright}
+Isabelle is a really powerful low-level proof assistant coming with a higher order logic (HOL) proving environment making the proofs easily readible and comprehensible without adding any abstract overlay.
+The term "assistant" designates the fact that Isabelle has numerous tools allowing various automations in the proofs such as a theorem seeker or an automatic solver.
+
+\subsection{Isabelle by example}
+The following example is a good introduction to the use of Isabelle.
+
 \pagebreak
@@
 \begin{remark}
-    In particular, $\forall v \in \mathcal{V}, v \in \mathcal{S}(v)$
+    In particular, $\forall v \in \mathcal{V}, v \in \mathcal{S}(v)$.
 \end{remark}

### 2022-03-24 LaTeX/report.tex
Updates

@@
 %__________________________________________
+\newcommand{\ldb}{\mathopen{\ooalign{\makebox[.4em][l]{$\lbrack$}\cr\makebox[.4em][r]{$\lbrack$}\cr}}}                                                       
+\newcommand{\rdb}{\mathclose{\ooalign{\makebox[.4em][l]{$\rbrack$}\cr\makebox[.4em][r]{$\rbrack$}\cr}}}
+
+%__________________________________________
 
@@
 %__________________________________________
-\definecolor{isarblue}{HTML}{006699}
-\definecolor{isargreen}{HTML}{009966}
+% Colors definition
+\definecolor{isa_red}{RGB}{255, 58, 71}
+\definecolor{isa_blue}{RGB}{0, 103, 158}
+\definecolor{isa_green}{RGB}{0, 157, 97}
+\definecolor{isa_dark_green}{RGB}{0,131, 0}
+\definecolor{isa_purple}{RGB}{174, 5, 238}
+\definecolor{isa_dark_blue}{RGB}{26, 0, 253}
+
+% Isabelle keywords
+\newcommand{\apply}{{\color{isa_red}{apply}}}
+\newcommand{\done}{{\color{isa_red}{done}}}
+\newcommand{\datatype}{{\color{isa_blue}{datatype}}}
+\newcommand{\inductive}{{\color{isa_blue}{inductive}}}
+\newcommand{\abbreviation}{{\color{isa_blue}{abbreviation}}}
+\newcommand{\thm}{{\color{isa_blue}{theorem}}}
+\newcommand{\lm}{{\color{isa_blue}{lemma}}}
+\newcommand{\fun}{{\color{isa_blue}{fun}}}
+\renewcommand{\locale}{{\color{isa_blue}{locale}}}
+\newcommand{\where}{{\color{isa_green}{where}}}
+\renewcommand{\and}{{\color{isa_green}{and}}}
+\newcommand{\fixes}{{\color{isa_green}{fixes}}}
+\newcommand{\assumes}{{\color{isa_green}{assumes}}}
+\newcommand{\shows}{{\color{isa_green}{shows}}}
+\newcommand{\generic}[1]{{\color{isa_purple}{\textquotesingle#1}}}
+\newcommand{\isa}[1]{\small\texttt{\\\noindent#1}}
+\newcommand{\blue}[1]{{\color{isa_dark_blue}{#1}}}
+\newcommand{\green}[1]{{\color{isa_dark_green}{#1}}}
+\newcommand{\env}[1]{$(\!|$#1$|\!)$}
+
+\theoremstyle{definition}
+\newtheorem{isabelle}{Isabelle}
+
 \lstdefinelanguage{isabelle}{%
@@
 \restoregeometry
+\pagebreak
+~
+\pagebreak
+
 \tableofcontents
@@
 \section{Introduction}
+\subsection{Academic context}
+This research work was carried out as part of my curriculum at the french \href{https://mines-nancy.univ-lorraine.fr}{École des Mines de Nancy}.
+All documents such as codes or source papers are available on a \href{https://github.com/VTrelat/Tarjan}{GitHub repository}.
+
 \subsection{Formal methods}
@@
 \subsection{Strongly connected components}
-\subsubsection{Directed graphs}
+\subsubsection{Directed graphs}\label{sec:directedgraphs}
 \begin{definition}
@@
     \KwResult{A partition \texttt{SCCs} of \VV where each element of \texttt{SCCs} is a maximal set of strongly connected components of \GG\;}
-    Initialize an empty set \texttt{DEAD}\;
+    Initialize an empty set \texttt{EXPLORED}\;
     Initialize an empty set \texttt{VISITED}\;
@@
         \ForEach{$w\in \texttt{POST(v)}$}{
-            \If{$w\in \texttt{DEAD}$}{
+            \If{$w\in \texttt{EXPLORED}$}{
                 continue\;
@@
             \textbf{report SCC} $\mathcal{S}(v)$\;
-            $\texttt{DEAD} := \texttt{DEAD} \cup \mathcal{S}(v)$\;
+            $\texttt{EXPLORED} := \texttt{EXPLORED} \cup \mathcal{S}(v)$\;
             \texttt{R.pop()}\;
@@
 \end{algorithm}
-\subsection{Correctness}
+
+\pagebreak
+
+\subsection{Informal proof}
+Note that this proof is said informal only because it is not formally automated. Both logical and mathematical arguments developped below are absolutely relevant.
+
 \begin{lemma}(First invariant)\label{lemma:disjointness}
@@
     \begin{equation*}
-        \biguplus_{v \in \texttt{R}} \mathcal{S}(v) = \textsc{Live} := \textsc{Visited} \setminus \textsc{Dead}
+        \biguplus_{v \in \texttt{R}} \mathcal{S}(v) = \textsc{Live} := \textsc{Visited} \setminus \textsc{Explored}
     \end{equation*}
@@
     The disjointness of all on-stack partial SCCs is given by lemma \ref{lemma:disjointness}.
-Nodes from $\textsc{Visited} \setminus \textsc{Dead}$ are in \texttt{R} because they are being processed.
+Nodes from $\textsc{Visited} \setminus \textsc{Eexplored}$ are in \texttt{R} because they are being processed.
 So, $\textsc{Live} \subseteq \texttt{R}$.\\
     By L.6-7 of algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.\\
-    L.9-10 ensure that no dead node is pushed in \texttt{R}.\\
-    L.24-25 keep the invariant by unstacking dead nodes from \texttt{R}, so $\texttt{R} \cap \textsc{Dead} = \varnothing$.
-Thus, $\texttt{R} = \textsc{Visited} \setminus \textsc{Dead} = \textsc{Live}$.
+    L.9-10 ensure that no explored node is pushed in \texttt{R}.\\
+    L.24-25 keep the invariant by unstacking explored nodes from \texttt{R}, so $\texttt{R} \cap \textsc{Explored} = \varnothing$.
+Thus, $\texttt{R} = \textsc{Visited} \setminus \textsc{Explored} = \textsc{Live}$.
 \end{proof}
@@
 
-\subsection{Isabelle proof}
-Not finished yet! See my oral presentation on January 14, 2022.
-%-----------------------------------------------------------------
+\subsection{Formal proof}
+Since the informal proof seems to be consistant, the formal -- automated -- proof can be written in Isabelle (HOL) based on the basis of the reasonning developped above.
+
+\subsubsection{Environment setup}
+The first definitions should be the different structures used in the algorithm.
+In particular, a record containing all the sets needed and described in the pseudo-code of algorithm \ref{alg:seqsetbased} :
+
+\isa{
+    {\color{isa_blue}record} \generic{v} env =\\
+    $~~~\mathcal{S}$ :: "\generic{v} $\Rightarrow$ \generic{v} set"\\
+    $~~~$explored :: "\generic{v} set"\\
+    $~~~$visited :: "\generic{v} set"\\
+    $~~~$sccs :: "\generic{v} set set"\\
+    $~~~$stack :: "\generic{v} list"\\
+}
+
+The following lines define a graph structure and some useful natural relations :
+
+\isa{
+    {\locale} graph =\\
+    $~~~$\fixes{} {\color{isa_dark_blue} vertices} :: "\generic{v} set" \and{} {\color{isa_dark_blue} successors} :: "\generic{v} $\Rightarrow$ \generic{v} set"\\
+    $~~~$\assumes{} vfin: "finite {\color{isa_dark_blue}vertices}"\\
+    $~~~$\and{} sclosed: "$\forall {\text{{\color{isa_dark_green}x}}} \in$ {\color{isa_dark_blue}vertices}.
+{\color{isa_dark_blue}successors} {\color{isa_dark_green}x} $\subseteq$ {\color{isa_dark_blue}vertices}"\\
+}
+\isa{
+    \abbreviation{} edge \where\\
+    $~~~$"{\color{isa_dark_blue}edge} {\color{isa_dark_green}x y} $\equiv$ {\color{isa_dark_green}y} $\in$ {\color{isa_dark_blue}successors} {\color{isa_dark_green}x}"
+}
+
+\isa{
+    \inductive{} reachable \where\\
+    $~~~$reachable\_refl[iff]: "{\color{isa_dark_blue}reachable} {\color{isa_dark_green}x x}"\\
+    $~{\text{|}}$ reachable\_succ[elim]: "$\ldb$edge {\color{isa_dark_green}x y};  {\color{isa_dark_blue}reachable} {\color{isa_dark_green}y z}$\rdb \Longrightarrow$ {\color{isa_dark_blue}reachable} {\color{isa_dark_green}x z}"\\
+}
+
+Those two relations are \texttt{edge}, which simply translates the property for two nodes of being linked by an edge and \texttt{reachable}, which is the binary relation $\Rightarrow^*$ defined in section \ref{sec:directedgraphs}.\\
+In order to be able to use those relations in the proofs later, it is essential to prove a list of lemmas, namely all the different natural properties that Isabelle cannot deduce\footnote{That is an abuse of language.
+The idea is for example that for the moment, there is no formal link between \texttt{edge} and \texttt{reachable}.
+The goal is to formalize it so Isabelle is logically able to both use and simplify some results in the proofs.} from nothing\footnote{There is actually a theorem fetcher that is particularly useful to find a basic set of lemmas.}.
+For instance, the following lemmas are essential.
+
+\isa{
+\lm{} reachable\_edge: "edge {\color{isa_dark_blue}x y} $\Longrightarrow$ reachable {\color{isa_dark_blue}x y}"\\
+$~~~~~${\color{isa_blue}by} auto\\
+}
+Mathematical writing: $\forall x, \forall y, x \Rightarrow y \Longrightarrow x \Rightarrow^* y$
+
+\isa{
+\lm{} succ\_reachable:\\
+$~~~~~$\assumes{} "reachable {\color{isa_dark_blue}x y}" \and{} "edge {\color{isa_dark_blue}y z}"\\
+$~~~~~$\shows "reachable {\color{isa_dark_blue}x z}"\\
+$~~~~~${\color{isa_blue}using} assms {\color{isa_blue}by} induct auto\\
+}
+Mathematical writing: $\forall x, \forall y, \forall z, (x \Rightarrow^* y \wedge y \Rightarrow z) \Longrightarrow x \Rightarrow^* z$
+
+\isa{
+\lm{} reachable\_trans:\\
+$~~~~~$\assumes{} y: "reachable \blue{x y}" \and{} z: "reachable \blue{y z}"\\
+$~~~~~$\shows{} "reachable \blue{x z}"\\
+$~~~~~${\color{isa_blue}using} assms {\color{isa_blue}by} induct auto\\
+}
+Mathematical writing: $\forall x, \forall y, \forall z, (x \Rightarrow^* y \wedge y \Rightarrow^* z) \Longrightarrow x \Rightarrow^* z$\\
+
+As the formal proofs will enventually deal with strongly connected components, it is also essential to formally define SCCs.
+For the purpose of the proof, the property of being a SCC is called \texttt{sub\_scc} and being a \textit{maximal} SCC is called \texttt{is\_scc} :
+
+\isa{
+{\color{isa_blue}definition} is\_subscc \where\\
+$~~~~~$"is\_subscc S $\equiv \forall$ x $\in$ S. $\forall$ y $\in$ S. reachable x y"\\
+}
+Mathematical writing: A set $S$ is a SCC if $\forall x \in S, \forall y \in S, x \Rightarrow^* y$
+
+\isa{
+{\color{isa_blue}definition} is\_scc \where\\
+$~~~~~$"is\_scc S $\equiv$ S $\neq$ \{\} $\wedge$ is\_subscc S\\
+$~~~~~\wedge$ ($\forall$ S'. S $\subseteq$ S' $\wedge$ is\_subscc S' $\longrightarrow$ S' = S)"\\
+}
+Mathematical writing: A non-empty SCC $S$ is maximal if for all SCC $S'$, $S \subseteq S' \Longrightarrow S'=S$\\
+
+Once again, there are some lemmas to prove, such as telling Isabelle when an element can be added to a SCC, or that two vertices that are reachable from each other are in the same SCC, or that two SCCs having a common element are identical.
+
+\subsubsection{Implementation of the algorithm}
+
+Now that the environment is set up, the actual algorithm -- seen as a function -- can be implemented.\\
+Since Isabelle does not support loops, the implementation will be split into two mutually recursive functions.
+The main function is called \texttt{dfs} and takes its name after the Depth First Search algorithm because the algorithm \ref{alg:seqsetbased} roughly consists in a deep traversal of a graph.
+The second function is called \texttt{dfss} and represents the \textit{while} loop of the algorithm \ref{alg:seqsetbased}.
+The two functions are mutually recursive because they recursively call each other.
+In particular, \texttt{dfss} will call either itself or \texttt{dfs}, depending on the case.
+
+\isa{
+    {\color{isa_blue}{function}} dfs :: "\generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \and{}\\
+    $~~~~~~~~~~$dfss:: "\generic{v} $\Rightarrow$ \generic{v} set $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where\\
+    "{\color{isa_dark_blue}dfs} {\color{isa_dark_green}v e} =\\
+    $~~~~~$({\color{isa_blue}{let}} {\color{isa_dark_green}e1} = {\color{isa_dark_green}e}\env{visited := visited e $\cup$ \{{\color{isa_dark_green}v}\}, stack := ({\color{isa_dark_green}v} \# stack {\color{isa_dark_green}e})};\\
+    $~~~~~~~~~~${\color{isa_dark_green}e'} = {\color{isa_dark_blue}dfss} {\color{isa_dark_green}v} (successors {\color{isa_dark_green}v}) {\color{isa_dark_green}e1}\\
+    $~~~~~${\color{isa_blue}{in if}} {\color{isa_dark_green}v} = hd(stack {\color{isa_dark_green}e'})\\
+    $~~~~~~~~~~${\color{isa_blue}{then}} {\color{isa_dark_green}e'}\env{sccs:=sccs {\color{isa_dark_green}e'} $\cup$ {$\mathcal{S}$ {\color{isa_dark_green}e' v}}, explored:=explored {\color{isa_dark_green}e'} $\cup$ ($\mathcal{S}$ {\color{isa_dark_green}e' v}), stack:=tl(stack {\color{isa_dark_green}e'})}\\
+    $~~~~~~~~~~${\color{isa_blue}else} {\color{isa_dark_green}e'})"\\
+    | "{\color{isa_dark_blue}dfss} {\color{isa_dark_green}v vs e} =\\
+    $~~~~~$({\color{isa_blue}if} {\color{isa_dark_green}vs} = \{\} {\color{isa_blue}then} {\color{isa_dark_green}e}\\
+    $~~~~~${\color{isa_blue}else} ({\color{isa_blue}let} {\color{isa_dark_green}w} = SOME {\color{isa_dark_green}x}. {\color{isa_dark_green}x} $\in$ {\color{isa_dark_green}vs}\\
+    $~~~~~~~~~~${\color{isa_blue}in} ({\color{isa_blue}let} {\color{isa_dark_green}e'} = ({\color{isa_blue}if} {\color{isa_dark_green}w} $\in$ explored {\color{isa_dark_green}e} {\color{isa_blue}then} {\color{isa_dark_green}e}\\
+    $~~~~~~~~~~~~~~~~~~~${\color{isa_blue}else if} {\color{isa_dark_green}w} $\notin$ visited {\color{isa_dark_green}e} {\color{isa_blue}then} {\color{isa_dark_blue}dfs} {\color{isa_dark_green}w e}\\
+    $~~~~~~~~~~~~~~~~~~~${\color{isa_blue}else} unite {\color{isa_dark_green}v w e})\\
+    $~~~~~~~~~~~~~~${\color{isa_blue}in} {\color{isa_dark_blue}dfss} {\color{isa_dark_green}v} ({\color{isa_dark_green}v} - \{{\color{isa_dark_green}w}\}) {\color{isa_dark_green}e'})))"\\
+    $~~~${\color{isa_blue}by} pat\_completeness (force+)\\
+}
+
+The two last keywords require explanations as well : \texttt{pat\_completeness} stands for \textit{pattern completeness} and ensures that there is no missing patterns.
+The keyword \texttt{force} is used\footnote{Here, \texttt{auto} cannot terminate because of the mutual recursion.} to help Isabelle know -- by proving it -- that both \texttt{dfs} and \texttt{dfss} are actually functions and that those functions are well defined with respect to the usual logical and mathematical meaning.
+
 \pagebreak

### 2022-03-31 LaTeX/report.tex
Report updates

@@
 
-Once again, there are some lemmas to prove, such as telling Isabelle when an element can be added to a SCC, or that two vertices that are reachable from each other are in the same SCC, or that two SCCs having a common element are identical.
+Once again, there are some lemmas to prove, such as telling Isabelle when an element can be added to a SCC, or that two vertices that are reachable from each other are in the same SCC, or that two SCCs having a common element are identical.\\
+
+\subsubsection{Ordering relation}
+In the proof, a ordering relation\footnote{In fact, a total order is being defined on stacks.} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack.
+Let $x$ and $y$ be two nodes and $R$ be a stack.
+Informally, $x$ precedes $y$ in $R$ if $x$ was pushed in $R$ before $y$ (see figure \ref{fig:stackorder}).
+
+\begin{figure}[!h]
+    \ctikzfig{stackorder}
+    \caption{The ordering relation on stacks\label{fig:stackorder}}
+\end{figure}
+
+\begin{definition}[Ordering relation]
+    Let $x$ and $y$ be two nodes and $xs$ be a stack.
+    \begin{equation*}
+        x \preceq y~\text{in}~xs \equiv \exists~h,~\exists~r,~(xs = h @ [x] @ r) \wedge (y \in [x]@r)
+    \end{equation*}
+\end{definition}
+
+The idea is to later use the following property: if $x \preceq y~\text{in}~xs$, then $x \Rightarrow^* y$.\\
+It is defined in Isabelle as follows:\\
+\isa{
+    {\color{isa_blue}definition} precedes ("\_ $\preceq$ \_ in \_" [100,100,100] 39) \where\\
+$~~~~~$"{\color{isa_dark_green}x} $\preceq$ {\color{isa_dark_green}y} in {\color{isa_dark_green}xs} $\equiv$ $\exists${\color{isa_dark_green}h} {\color{isa_dark_green}r}.
+{\color{isa_dark_green}xs} = {\color{isa_dark_green}h} @ ({\color{isa_dark_green}x} \# {\color{isa_dark_green}r}) $\wedge$ {\color{isa_dark_green}y} $\in$ set ({\color{isa_dark_green}x} \# {\color{isa_dark_green}r})"\\
+}
+
+All the different properties (\textit{i.e.} lemmas) which follow this definition in the Isabelle implementation are detailed in the natural mathematical writing below:\\
+
+Let $x, y, z$ be three nodes, and let $xs, ys, zs$ be three lists of nodes representing stacks.\\
+By abuse of language, if an element is on a stack, it is in the set of elements contained in the stack so the following statement can be written: $x$ is on $xs \Longleftrightarrow x \in xs$.
+However, $xs$ in not seen as the set representing $xs$ since an element may occur several times in a stack.\\
+The operator $@$ denotes the concatenation and operates on two lists: $[x_0, \dots, x_n] @ [y_0, \dots, y_m] = [x_1, \dots, x_n, y_1, \dots, y_m]$.\\
+
+\begin{flushleft}
+    (i)
+$x \preceq y~\text{in}~xs \Longrightarrow (x \in xs)\wedge(y \in xs)$
+\end{flushleft}
+\begin{flushleft}
+    (ii)
+$y \in [x@ xs \Longrightarrow (x \in xs)\wedge(y \in xs)$
+\end{flushleft}
+
+\begin{flushleft}
+    (iii)
+    $x \neq z \Longrightarrow (x \preceq y~\text{in}~([z] @ zs) \Longrightarrow x \preceq y ~\text{in}~zs)$
+\end{flushleft}
+
+\begin{flushleft}
+    (iv)
+    $(y \preceq x~\text{in}~([x] @ xs)) \wedge (x \notin xs) \Longrightarrow (x = y)$
+\end{flushleft}
+
+\begin{flushleft}
+    (v)
+    $y \in (ys @ [x]) \Longrightarrow y \preceq x~\text{in}~(ys @ [x] @ xs)$
+\end{flushleft}
+
+\begin{flushleft}
+    (vi)
+    $(x \preceq x~\text{in}~xs) = (x \in xs)$
+\end{flushleft}
+
+\begin{flushleft}
+    (vii)
+    $x \preceq y~\text{in}~xs \Longrightarrow x \preceq y~\text{in}~(ys @ xs)$
+\end{flushleft}
+
+\begin{flushleft}
+    (viii)
+    $x \notin ys \Longrightarrow (x \preceq y~\text{in}~(ys @ xs) \Longleftrightarrow x \preceq y~\text{in}~xs)$
+\end{flushleft}
+
+\begin{flushleft}
+    (ix)
+    $x \preceq y~\text{in}~xs \Longrightarrow x \preceq y~\text{in}~(xs @ ys)$
+\end{flushleft}
+
+\begin{flushleft}
+    (x)
+    $y \notin ys \Longrightarrow x \preceq y~\text{in}~(xs @ ys) \Longleftrightarrow x \preceq y~\text{in}~xs$
+\end{flushleft}
+
+\begin{flushleft}
+    (xi)
+    $(x \preceq y~\text{in}~xs) \wedge (y \preceq z~\text{in}~xs) \wedge \underset{\text{all elements on $xs$ are distinct}}{\underbrace{(\forall~0 \leq i < j \leq \text{length}(xs), xs[i] \neq xs[j])}} \Longrightarrow x \preceq z~\text{in}~xs$
+\end{flushleft}
+
+% continue with precedes_append_right_iff
 

### 2022-04-08 LaTeX/report.tex
comments on report and small fix in Isabelle theory

@@
 \input{style.tikzstyles}
+
+% editorial comments in the text or in marginal notes
+% 1st argument: initials of the person making the comment,
+% 2nd argument: comment to insert
+\long\def\ednote#1#2{\par\noindent\framebox{\begin{minipage}{0.99\linewidth}\linespread{.7}\footnotesize #1: #2\end{minipage}}\par}
+\newcommand{\edmargin}[2]{\marginpar{\raggedright\linespread{.7}\tiny #1: #2}}
 
@@
 \pagebreak
-~
+
 \pagebreak
@@
 \subsection{Academic context}
-This research work was carried out as part of my curriculum at the french \href{https://mines-nancy.univ-lorraine.fr}{École des Mines de Nancy}.
+This research work was carried out as part of my curriculum at the french\edmargin{sm}{French} \href{https://mines-nancy.univ-lorraine.fr}{École des Mines de Nancy}.
 All documents such as codes or source papers are available on a \href{https://github.com/VTrelat/Tarjan}{GitHub repository}.
@@
 Formal methods find applications in a variety of fields, both concrete, such as the railway industry or self-driving cars, and abstract, such as computational architecture.\\
-Altough a formal proof lies first on paper, the real formalisation starts when proofs are mechanised in a proof assistant.
+Although a formal proof lies first on paper, the real formalisation starts when proofs are mechanised in a proof assistant.
+
+\ednote{sm}{%
+  First, a stylistic remark: do not end lines in a \LaTeX\ document using \texttt{\textbackslash\textbackslash} (except in \texttt{tabular} environments or similar).
+Use a blank line to start a new paragraph: the style will define the appearance of a paragraph break.
+
+  More importantly, formal methods should not be equated with theorem proving.
+They are really about giving precise, mathematical definitions to computer science concepts.
+Although the purpose of giving such definitions is to enable formal verification, many techniques besides theorem proving, such as model-based testing, run-time monitoring, model checking etc.\ are used.
+}
 
@@
 \end{flushright}
-Isabelle is a really powerful low-level proof assistant coming with a higher order logic (HOL) proving environment making the proofs easily readible and comprehensible without adding any abstract overlay.
-The term "assistant" designates the fact that Isabelle has numerous tools allowing various automations in the proofs such as a theorem seeker or an automatic solver.
+Isabelle is a really powerful low-level\edmargin{sm}{What does ``low-level'' mean?} proof assistant coming with a higher order logic (HOL) proving environment making the proofs easily readable and comprehensible without adding any abstract overlay.\edmargin{sm}{I don't understand what that means.} The term ``assistant'' designates the fact that Isabelle has numerous tools allowing various automations in the proofs such as a theorem seeker or an automatic solver.
+
+\ednote{sm}{%
+  What is a ``theorem seeker''?
+In fact, ``assistant'' refers to the fact that the machine checks the proof provided by the user, in contrast to automatic theorem proving where the machine finds the proof itself.
+The tools for automation are intended to help the user write the proof at a conveniently high level, without needing to work at the level of a logical calculus, for example.
+}
 
@@
 
+\ednote{sm}{%
+  First, there is a break of notation (\NN vs.\ \texttt{N}).
+Also, the ``dot'' notation for records is not used in Isabelle.
+Finally, as you say, the presentation corresponds to an algorithm.
+I'd first introduce graphs mathematically and refer to data structures only with respect to a particular algorithm.
+}
+
 \subsection{Graphs and their representation}
@@
     \item $\mathcal{V}$ is a set of vertices
-    \item $\mathcal{E} \subseteq \{(x,y) \in V^2\}$ is a set of edges\footnote{Note the use of the couple $(x, y)$ and not the pair $\{x, y\}$ that makes the graph directed.}.
+    \item $\mathcal{E} \subseteq \{(x,y) \in V^2\}$ is a set of edges\footnote{Note the use of the couple $(x, y)$ and not the pair $\{x, y\}$ that makes the graph directed.}.\edmargin{sm}{couple $\rightarrow$ (ordered) pair, pair $\rightarrow$ set}
 \end{itemize}
-Vertices will often be called nodes and edges will be represented through adjency lists for each node.\\
+Vertices will often be called nodes and edges will be represented through adjency lists\edmargin{sm}{Again, that's an algorithmic concept.} for each node.\\
 Let us give an example. Let \GG = $(\mathcal{V}, \mathcal{E})$ be the graph represented on figure \ref{fig:graph}.
@@
 Thus, \VV = $\{0,1,2,3,4,5,6,7\}$ and
-\begin{flushleft}
+\begin{flushleft}\edmargin{sm}{Use numbers, not letters, in the figure so that the correspondence becomes clearer.}
     ~~~~~~~~\EE = $\{(0,0),(0,1),(0,2),(0,3),$\\
@@
 
-This representation being somewhat long, adjency lists can be used instead and therefore it gives:
+This representation being somewhat long, adjacency lists can be used instead and therefore it gives:
 \begin{center}
-    \texttt{G.adjency = [[0,1,2,3],[4,7],[0,1,2,5],[3,6],[6],[3],[6]]}
+    \texttt{G.adjacency = [[0,1,2,3],[4,7],[0,1,2,5],[3,6],[6],[3],[6]]}
 \end{center}
-Thus, for all $i \in \{0,\ldots,7\}$, \texttt{G.adjency[i]} is the list of nodes to which node $i$ is connected -- \textit{i.e.} there is an directed edge from node $i$ to every node of \texttt{G.adjency[i]}.
+Thus, for all $i \in \{0,\ldots,7\}$, \texttt{G.adjacency[i]} is the list of nodes to which node $i$ is connected -- \textit{i.e.} there is a directed edge from node $i$ to every node of \texttt{G.adjacency[i]}.
 \pagebreak
@@
 \subsubsection{Directed graphs}\label{sec:directedgraphs}
+
+\ednote{sm}{%
+  I suggest moving this material to section 2: all of this is plain math. You can then show how these definitions are written in Isabelle.
+}
 \begin{definition}
-    For two vertices $x$ and $y$ of \VV, the relation "has an edge to" is noted "$\Rightarrow$" such that
+    For two vertices $x$ and $y$ of \VV, the relation ``has an edge to'' is noted ``$\Rightarrow$'' such that
     \begin{equation*}
-        (x, y) \in \mathcal{E} \Longleftrightarrow x \Rightarrow y
+        (x, y) \in \mathcal{E} \quad\Longleftrightarrow\quad x \Rightarrow y
     \end{equation*}
-    By extension, if there is a path from $x$ to $y$ with more than one edge, the same notation is kept for the sake of simplicity.
+    By extension, if there is a path from $x$ to $y$ with more than one edge, the same notation is kept for the sake of simplicity.\edmargin{sm}{The two sentences seem to contradict each other, and in order to avoid confusion I suggest writing systematically $\Rightarrow^*$ in order to denote reachability.}
     The reflexive and transitive closure of the relation $\Rightarrow$ is noted $\Rightarrow^*$.
@@
         \ctikzfig{example2}
+        \ednote{sm}{The notion of connectedness has not been introduced, and what's the relationship to SCCs?}
         \subcaption{Strongly connected graph}
@@
     \ctikzfig{exampleSCC}
+\ednote{sm}{%
+  ``Set of SCC'' is confusing: an SCC is a set, and you are not talking about sets of sets here.
+}
     \caption{Example of a graph where each colored set of node is a maximal set of SCC\label{fig:exampleSCC}}
@@
     \ctikzfig{exampleSCCreduced}
-    \caption{Reduced visualization of the graph represented if figure \ref{fig:exampleSCC}}
+    \caption{Reduced visualization of the graph represented in figure \ref{fig:exampleSCC}}
 \end{figure}
@@
 \subsubsection{DFS and \textit{num} value}
-Tarjan's SCC algorithm basically lies on a depth-first search. The figure \ref{fig:dfs} shows an example of a DFS traversal on a simple directed graph.
+Tarjan's SCC algorithm\edmargin{sm}{not mentioned yet} basically lies on a depth-first search. The figure \ref{fig:dfs} shows an example of a DFS traversal on a simple directed graph.
 \begin{figure}[!h]
@@
     \end{equation*}
-    In this case, the backtracking edge from $v$ to $u$ is represented by $v \hookrightarrow u$.\\      
+    In this case, the backtracking edge from $v$ to $u$ is represented by $v \hookrightarrow u$.\edmargin{sm}{add example for the graph of Fig.5}\\      
 \end{definition}
@@
 
-A more formal definition would be the following:
+A more formal definition would be\edmargin{sm}{is?} the following:
 \begin{definition}
@@
 \subsubsection{Example}
-Let \GG be the graph given in fig. \ref{fig:exlowlink}. The order of traversal of the graph is given by the value \textit{num} for each node of \GG. The \textit{lowlink} value is also displayed.
+Let \GG be the graph given in fig.
+\ref{fig:exlowlink}.\edmargin{sm}{``Figure'' vs.\ ``figure'' vs.\ ``Fig.'': usually capitalized, make it uniform.
+And why not use the same graph as before?} The order of traversal of the graph is given by the value \textit{num} for each node of \GG.
+The \textit{lowlink} value is also displayed.
 
@@
     \ctikzfig{lowlink}
+    \ednote{sm}{couple of value $\rightarrow$ pair of values}
     \caption{A DFS was performed through \GG from $x_1$ and next to each node is represented the couple of value (\textit{num}, \textit{lowlink})\label{fig:exlowlink}}
@@
 Knowing the backtracking edges, all \textit{lowlink} values can be computed\footnote{In fact, they are refreshed during the DFS.}.\\
-Let us take $x_8$ as an example: its \textit{lowlink} is equal to 2, which actually means that $x_6$\footnote{Because $\texttt{x6.num} = 2$} is its anchor -- or attachment node -- in their SCC, namely the green one.
-Indeed, $x_8$ is alone in its equivalence class\footnote{for the relation $\Rightarrow$}, and from all nodes linked by one of the backtracking edges of $x_8$, $x_6$ has the minimum value \textit{num}.
+Let us take $x_8$ as an example: its \textit{lowlink} is equal to 2, which actually means that $x_6$\footnote{Because $\texttt{x6.num} = 2$} is its anchor -- or attachment node -- in their\edmargin{sm}{reference?} SCC, namely the green one.
+Indeed, $x_8$ is alone in its equivalence class\footnote{for the relation $\Rightarrow$},\edmargin{sm}{I don't understand the notion of equivalence, e.g.
+we have $x_8 \Rightarrow^* x_9 \Rightarrow^* x_8$.} and from all nodes linked by one of the backtracking edges of $x_8$, $x_6$ has the minimum value \textit{num}.
 Likewise, $x_3$ is its own attachment node in the green SCC since it is the first node visited when performing the DFS.
@@
 \section{Tarjan's algorithm}
+\ednote{sm}{%
+  Why present Tarjan's algorithm, which is not considered further here?
+Perhaps because it is the best known algorithm for computing SCCs?
+But then it should be compared with the set-based algorithm that this project is about?
+}
 Tarjan's algorithm is an efficient on-the-fly SCC computing algorithm \cite{tarjanMerz}.
@@
     \ForAll{$v \in \mathcal{V}$}{
-        \If{\texttt{v.num} is unefined}{
+        \If{\texttt{v.num} is undefined}{
             SCC(v)\;
@@
         \texttt{v.onStack = true}\;
-        \ForAll{$w \in \text{POST}(v)$}{
+        \ForAll{$w \in \text{POST}(v)$\quad\textcolor{red}{sm: POST undefined?}}{
             \If{\texttt{w.num} is undefined}{
-                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.lowlink})$\;
+                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.lowlink})$ \quad\textcolor{red}{w.lowlink undefined?}\;
             }
@@
             Initialize an empty set \texttt{currentSCC}\;
-            \Repeat{$\texttt{v.num} \neq \texttt{w.num}$}{
+            \Repeat{$\texttt{v.num} \neq \texttt{w.num}$\quad\textcolor{red}{sm: w.lowlink?}}{
                 Let $w := \texttt{R.pop()}$\;
@@
 \begin{definition}[SCC mapping]
-    In the following algorithm, the SCCs are progressively tracked in a collection of disjoint sets through a map $\mathcal{S} : \mathcal{V} \longrightarrow \mathcal{P}(\mathcal{V})$, where $\mathcal{P}(\mathcal{V})$ is the powerset of \VV, s.t.
+    In the following algorithm, the SCCs are progressively tracked in a collection of disjoint\edmargin{sm}{We may have $\mathcal{S}(v) = \mathcal{S}(w)$ for $v \neq w$?!} sets through a map $\mathcal{S} : \mathcal{V} \longrightarrow \mathcal{P}(\mathcal{V})$, where $\mathcal{P}(\mathcal{V})$ is the powerset of \VV, s.t.
 the following invariant is maintained:
@@
 \end{algorithm}
+\ednote{sm}{%
+  The algorithm should also explain how $\mathcal{S}$ is initialized.
+}
 
@@
 \subsection{Informal proof}
-Note that this proof is said informal only because it is not formally automated. Both logical and mathematical arguments developped below are absolutely relevant.
+Note that this proof is said informal only because it is not formally automated.\edmargin{sm}{checked by a mechanized proof assistant?} Both logical and mathematical arguments developed below are absolutely relevant.
 
@@
     Let $x \in \mathcal{V}$ be the following node to be visited during the execution of the algorithm \ref{alg:seqsetbased}: $x$ is pushed in \texttt{R}.
-Let $y \in \textsc{Post}(x)$.
-There are two cases:\\
+Let $y \in \textsc{Post}(x)$.\edmargin{sm}{The lemma assumes $y \in \texttt{R}$, is $y \in \textsc{Post}(x)$ an additional assumption?
+Why is it justified?} There are two cases:\\
     \begin{itemize}
@@
 Thus, a DFS-like traversal is performed from $y$, so $y$ is pushed in \texttt{R} and $\mathcal{S}(y) = \{y\}$ because $y$ is alone in its equivalence class for the moment since it has not been visited yet.\\
-        Therefore, $\mathcal{S}(x) \cap \mathcal{S}(y) = \varnothing$.
+        Therefore, $\mathcal{S}(x) \cap \mathcal{S}(y) = \varnothing$.\edmargin{sm}{I don't see an argument why $y \notin \mathcal{S}(x)$?}
 
@@
 Let $(x_i)_{1\leq i \leq n}$ be the first nodes of the stack s.t.
-$x_0 = x$ and $x_n = y$.\\
+$x_0 = x$ and $x_n = y$.\edmargin{sm}{Should explain the order of the stack.}\\
         In order to avoid writing $\texttt{R} = [\ldots, y, \ldots, x]$, let us define $\widetilde{R}$ the stack containing the first $n$ nodes in \texttt{R}, s.t.
 $\widetilde{R} = [y, \ldots, x] = [x_n, \ldots, x_0]$.\\
-        Let us consider the worst case, \textit{i.e.} when
+        Let us consider the worst case, \textit{i.e.} when\edmargin{sm}{Why is this the worst case?}
         \begin{equation*}
@@
         So, the while loop has to go down to $y$ because all partial SCCs are disjoint. As the length of the stack \texttt{R} is bounded by $|\mathcal{V}|$, the algorithm terminates.\\
-        $x_0$ is first unstacked and both $\mathcal{S}(x_0)$ and $\mathcal{S}(\texttt{R.top()}) = \mathcal{S}(x_1)$ are then united. The current state of $\mathcal{S}$ and $\widetilde{R}$ is:
+        $x_0$ is first unstacked and both $\mathcal{S}(x_0)$ and $\mathcal{S}(\texttt{R.top()}) = \mathcal{S}(x_1)$ are then united.
+The current state of $\mathcal{S}$ and $\widetilde{R}$ is:\edmargin{sm}{Ambiguity between $\mathcal{S}$ as a mapping and as a set of equivalence classes.}
         \begin{equation*}
@@
     The disjointness of all on-stack partial SCCs is given by lemma \ref{lemma:disjointness}.
-Nodes from $\textsc{Visited} \setminus \textsc{Eexplored}$ are in \texttt{R} because they are being processed.
+Nodes from $\textsc{Visited} \setminus \textsc{Explored}$ are in \texttt{R} because they are being processed.
 So, $\textsc{Live} \subseteq \texttt{R}$.\\
@@
 \begin{proof}
-    Let $v \in \textsc{Live} = \displaystyle{\biguplus_{v \in \texttt{R}} \mathcal{S}(v)}$.
-$v$ is in a unique partial SCC $\mathscr{S} := \mathcal{S}(v)$.
+    Let $v \in \textsc{Live} = \displaystyle{\biguplus_{v \in \texttt{R}} \mathcal{S}(v)}$.\edmargin{sm}{Avoid using $v$ as a free variable and a bound variable in the same formula.
+Rather use $r$ for the bound variable, as in the lemma.} $v$ is in a unique partial SCC $\mathscr{S} := \mathcal{S}(v)$.
 Because of lemma \ref{lemma:disjointness}, there cannot exist $x \neq y \in \texttt{R}$ s.t.
@@
 \begin{proof}
-    Holds because of corollary \ref{cor:cor1}.
+    Holds because of corollary \ref{cor:cor1}.\edmargin{sm}{In fact, just a weakening of Cor.1, why state it separately?}
 \end{proof}
@@
 \begin{remark}
-    In the algorithm \ref{alg:seqsetbased}, this property is held by L.16-18.
-These lines also illustrate how the algorithm "reads" the SCCs.
+    In the algorithm \ref{alg:seqsetbased}, this property is held\edmargin{sm}{maintained?} by L.16-18.
+These lines also illustrate how the algorithm ``reads'' the SCCs.
 Corollary \ref{cor:cor2} shows that when the mapped representatives of the top two nodes of \texttt{R} are united (until $\mathcal{S}(w') = \mathcal{S}(v) = \mathcal{S}(w)$ since $w'$ has a path to $v$), then all united components are in the same SCC.
@@
 \begin{remark}\label{rem:proof}
-    Because \texttt{R} only contains exactly one representative for each partial SCC (corollary \ref{cor:cor1}), after each step of the main loop -- \textit{i.e.} the DFS -- every partial SCC is actually maximal in the current set of visited nodes.
+    Because \texttt{R} only contains exactly one representative for each partial SCC \edmargin{sm}{I think the fact that this representative is the lowest one in the stack also plays a role here?} (corollary \ref{cor:cor1}), after each step of the main loop -- \textit{i.e.} the DFS -- every partial SCC is actually maximal in the current set of visited nodes.
 \end{remark}
@@
 \subsection{Formal proof}
-Since the informal proof seems to be consistant, the formal -- automated -- proof can be written in Isabelle (HOL) based on the basis of the reasonning developped above.
+Since the informal proof seems to be consistant,\edmargin{sm}{consistent, or rather convincing?} the formal -- automated -- proof\edmargin{sm}{be careful: it is checked automatically, but found by the user!} can be written in Isabelle (HOL) based on the basis of the reasoning developed above.
 
@@
 The first definitions should be the different structures used in the algorithm.
-In particular, a record containing all the sets needed and described in the pseudo-code of algorithm \ref{alg:seqsetbased} :
+In particular, a record containing all the sets needed and described in the pseudo-code of algorithm \ref{alg:seqsetbased} :\edmargin{sm}{I'd explain what 'v stands for.
+Also, somewhat inconsistent spacing.}
 
@@
 
-The following lines define a graph structure and some useful natural relations :
+The following lines define a graph structure and some useful natural relations :\edmargin{sm}{Is there an advantage to introducing the graph via ``successors'' rather than a relation as in the informal presentation?}
 
@@
 
-Those two relations are \texttt{edge}, which simply translates the property for two nodes of being linked by an edge and \texttt{reachable}, which is the binary relation $\Rightarrow^*$ defined in section \ref{sec:directedgraphs}.\\
+Those two relations are \texttt{edge}, which simply translates the property for two nodes of being linked by an edge and \texttt{reachable},\edmargin{sm}{I think it's worthwhile to explain ``inductive''.} which is the binary relation $\Rightarrow^*$ defined in section \ref{sec:directedgraphs}.\\
 In order to be able to use those relations in the proofs later, it is essential to prove a list of lemmas, namely all the different natural properties that Isabelle cannot deduce\footnote{That is an abuse of language.
@@
 The goal is to formalize it so Isabelle is logically able to both use and simplify some results in the proofs.} from nothing\footnote{There is actually a theorem fetcher that is particularly useful to find a basic set of lemmas.}.
-For instance, the following lemmas are essential.
+For instance, the following lemmas are essential.\edmargin{sm}{Given that the first lemma is proved by auto, is it really useful?}
 
@@
 \subsubsection{Ordering relation}
-In the proof, a ordering relation\footnote{In fact, a total order is being defined on stacks.} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack.
+In the proof, a ordering\edmargin{sm}{precedence?
+In order to avoid the impression that one orders stacks?} relation\footnote{In fact, a total order is being defined on stacks.} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack.
 Let $x$ and $y$ be two nodes and $R$ be a stack.
-Informally, $x$ precedes $y$ in $R$ if $x$ was pushed in $R$ before $y$ (see figure \ref{fig:stackorder}).
+Informally, $x$ precedes $y$ in $R$ if $x$ was pushed in $R$ before $y$ (see figure \ref{fig:stackorder}).\edmargin{sm}{The formal definition is the opposite of what's shown in the figure.}
 
@@
 
-The idea is to later use the following property: if $x \preceq y~\text{in}~xs$, then $x \Rightarrow^* y$.\\
+The idea is to later use the following property: if $x \preceq y~\text{in}~xs$, then $x \Rightarrow^* y$.\edmargin{sm}{Probably the other way around.}\\
 It is defined in Isabelle as follows:\\
@@
 However, $xs$ in not seen as the set representing $xs$ since an element may occur several times in a stack.\\
-The operator $@$ denotes the concatenation and operates on two lists: $[x_0, \dots, x_n] @ [y_0, \dots, y_m] = [x_1, \dots, x_n, y_1, \dots, y_m]$.\\
+The operator $@$ denotes the concatenation and operates on two lists: $[x_0, \dots, x_n] @ [y_0, \dots, y_m] = [x_1, \dots, x_n, y_1, \dots, y_m]$.\edmargin{sm}{$x_0$, $y_0$}\\
 
@@
 $y \in [x@ xs \Longrightarrow (x \in xs)\wedge(y \in xs)$
+\edmargin{sm}{I don't understand, $y=x \lor y \in xs$? (And ``$[x]$''.)}
 \end{flushleft}
@@
     (xi)
-    $(x \preceq y~\text{in}~xs) \wedge (y \preceq z~\text{in}~xs) \wedge \underset{\text{all elements on $xs$ are distinct}}{\underbrace{(\forall~0 \leq i < j \leq \text{length}(xs), xs[i] \neq xs[j])}} \Longrightarrow x \preceq z~\text{in}~xs$
+    $(x \preceq y~\text{in}~xs) \wedge (y \preceq z~\text{in}~xs) \wedge \underset{\text{all elements of $xs$ are distinct}}{\underbrace{(\forall~0 \leq i < j \leq \text{length}(xs), xs[i] \neq xs[j])}} \Longrightarrow x \preceq z~\text{in}~xs$
 \end{flushleft}
+
+\ednote{sm}{%
+  Perhaps it would be enough to just say that one builds a library of lemmas and display a few of them?
+}
 
@@
 
+\ednote{sm}{Should also show the Isabelle representation of unite.}
+
 The two last keywords require explanations as well : \texttt{pat\_completeness} stands for \textit{pattern completeness} and ensures that there is no missing patterns.
-The keyword \texttt{force} is used\footnote{Here, \texttt{auto} cannot terminate because of the mutual recursion.} to help Isabelle know -- by proving it -- that both \texttt{dfs} and \texttt{dfss} are actually functions and that those functions are well defined with respect to the usual logical and mathematical meaning.
+The keyword \texttt{force} is used\footnote{Here, \texttt{auto} cannot terminate because of the mutual recursion.}\edmargin{sm}{Not sure if mutual recursion is a problem here: force is more aggressive in instantiation and seems to find the right instance.} to help Isabelle know -- by proving it -- that both \texttt{dfs} and \texttt{dfss} are actually functions and that those functions are well defined with respect to the usual logical and mathematical meaning.
 
@@
 
-\bibitem{tarjanMerz} R. Chen, C. Cohen, J.-J. Lévy, S. Merz, L. Théry, \emph{Formal Proofs of Tarjan’s Strongly Connected 2 Components Algorithm in Why3, Coq and Isabelle}, 2019
+\bibitem{tarjanMerz} R. Chen, C. Cohen, J.-J. Lévy, S. Merz, L. Théry, \emph{Formal Proofs of Tarjan’s Strongly Connected Components Algorithm in Why3, Coq and Isabelle}, 2019
 

### 2022-04-24 LaTeX/report.tex
Updates in the report and in the proof

@@
 %        \huge{Strongly connected components algorithms : parallelization and proofs}
-        \huge{Formal methods and assisted proofs: application to strongly connected components algorithms}
+%        \huge{Formal methods and assisted proofs: application to strongly connected components algorithms}
+        \huge{Formal verification of an algorithm for computing strongly connected components}
     \end{center}
@@
 \subsection{Academic context}
-This research work was carried out as part of my curriculum at the french\edmargin{sm}{French} \href{https://mines-nancy.univ-lorraine.fr}{École des Mines de Nancy}.
+This research work was carried out as part of my curriculum at the French \href{https://mines-nancy.univ-lorraine.fr}{École des Mines de Nancy}.
 All documents such as codes or source papers are available on a \href{https://github.com/VTrelat/Tarjan}{GitHub repository}.
@@
 The whole purpose of the discipline is to ensure by a logical proof that a given algorithm is not only correct on its domain of definition, but also to find -- or define -- that domain.
-Formal methods find applications in a variety of fields, both concrete, such as the railway industry or self-driving cars, and abstract, such as computational architecture.\\
+Formal methods find applications in a variety of fields, both concrete, such as the railway industry or self-driving cars, and abstract, such as computational architecture.
+\BlankLine
 Although a formal proof lies first on paper, the real formalisation starts when proofs are mechanised in a proof assistant.
@@
 \end{flushright}
-Isabelle is a really powerful low-level\edmargin{sm}{What does ``low-level'' mean?} proof assistant coming with a higher order logic (HOL) proving environment making the proofs easily readable and comprehensible without adding any abstract overlay.\edmargin{sm}{I don't understand what that means.} The term ``assistant'' designates the fact that Isabelle has numerous tools allowing various automations in the proofs such as a theorem seeker or an automatic solver.
-
-\ednote{sm}{%
-  What is a ``theorem seeker''?
+Isabelle is a really powerful proof assistant coming with a higher order logic (HOL) proving environment.
+Isabelle proofs are written in the Isar (``intelligible semi-automated reasoning'') language that is designed to make proofs readable and comprehensible for a mathematically inclined reader, with minimal overhead introduced by the formalism.
 In fact, ``assistant'' refers to the fact that the machine checks the proof provided by the user, in contrast to automatic theorem proving where the machine finds the proof itself.
 The tools for automation are intended to help the user write the proof at a conveniently high level, without needing to work at the level of a logical calculus, for example.
-}
 
 \subsection{Isabelle by example}
-The following example is a good introduction to the use of Isabelle.
+The following example is a good introduction to the use of Isabelle.\edmargin{sm}{I'd explain what 'v stands for.}
 
-\pagebreak
-
-\section{Models and representation}
-\subsection{Nodes}
-% Je pense que cette description n'a de sens que dans le contexte d'un algorithme particulier.
-Dans le document final il faudra commencer par décrire le problème indépendamment de l'algorithme utilisé pour le résoudre, d'autant plus qu'il y a deux algorithmes différents.
-
-Vertices of a graph can be represented as nodes.
-A node \NN is a simple data structure composed of an index, a boolean value telling if it has already been visited and two integer values \textit{num} and \textit{lowlink} whose role will be explained later.\\
-In the following, the aforesaid attributes will be referred to through the following notation:\\
-Let \NN be a node.
-The attributes of \NN can be accessed via \texttt{N.index}, \texttt{N.visited}, \texttt{N.num} and \texttt{N.lowlink}.
-This notation will be applied to any object that lends itself to it.
-
-\ednote{sm}{%
-  First, there is a break of notation (\NN vs.\ \texttt{N}).
-Also, the ``dot'' notation for records is not used in Isabelle.
-Finally, as you say, the presentation corresponds to an algorithm.
-I'd first introduce graphs mathematically and refer to data structures only with respect to a particular algorithm.
-}
-
-\subsection{Graphs and their representation}
-A graph \GG is the data $(\mathcal{V}, \mathcal{E})$ where:
-\begin{itemize}
-    \item $\mathcal{V}$ is a set of vertices
-    \item $\mathcal{E} \subseteq \{(x,y) \in V^2\}$ is a set of edges\footnote{Note the use of the couple $(x, y)$ and not the pair $\{x, y\}$ that makes the graph directed.}.\edmargin{sm}{couple $\rightarrow$ (ordered) pair, pair $\rightarrow$ set}
-\end{itemize}
-Vertices will often be called nodes and edges will be represented through adjency lists\edmargin{sm}{Again, that's an algorithmic concept.} for each node.\\
-Let us give an example. Let \GG = $(\mathcal{V}, \mathcal{E})$ be the graph represented on figure \ref{fig:graph}.
-\begin{figure}[!h]
-    \ctikzfig{graph}
-    \caption{Representation of \GG\label{fig:graph}}
-\end{figure}
-Thus, \VV = $\{0,1,2,3,4,5,6,7\}$ and
-\begin{flushleft}\edmargin{sm}{Use numbers, not letters, in the figure so that the correspondence becomes clearer.}
-    ~~~~~~~~\EE = $\{(0,0),(0,1),(0,2),(0,3),$\\
-    $~~~~~~~~~~~~~~~(1,4),(1,7),$\\
-    $~~~~~~~~~~~~~~~(3,0),(3,1),(3,2),(3,5),$\\
-    $~~~~~~~~~~~~~~~(4,3),(4,6),$\\
-    $~~~~~~~~~~~~~~~(5,6),$\\
-    $~~~~~~~~~~~~~~~(6,3),$\\
-    $~~~~~~~~~~~~~~~(7,6)\}$
-\end{flushleft}
-
-This representation being somewhat long, adjacency lists can be used instead and therefore it gives:
-\begin{center}
-    \texttt{G.adjacency = [[0,1,2,3],[4,7],[0,1,2,5],[3,6],[6],[3],[6]]}
-\end{center}
-Thus, for all $i \in \{0,\ldots,7\}$, \texttt{G.adjacency[i]} is the list of nodes to which node $i$ is connected -- \textit{i.e.} there is a directed edge from node $i$ to every node of \texttt{G.adjacency[i]}.
 \pagebreak
@@
 
-\ednote{sm}{%
-  I suggest moving this material to section 2: all of this is plain math. You can then show how these definitions are written in Isabelle.
-}
-\begin{definition}
-    For two vertices $x$ and $y$ of \VV, the relation ``has an edge to'' is noted ``$\Rightarrow$'' such that
-    \begin{equation*}
-        (x, y) \in \mathcal{E} \quad\Longleftrightarrow\quad x \Rightarrow y
-    \end{equation*}
-    By extension, if there is a path from $x$ to $y$ with more than one edge, the same notation is kept for the sake of simplicity.\edmargin{sm}{The two sentences seem to contradict each other, and in order to avoid confusion I suggest writing systematically $\Rightarrow^*$ in order to denote reachability.}
-    The reflexive and transitive closure of the relation $\Rightarrow$ is noted $\Rightarrow^*$.
+\begin{definition}[Reachability]
+    For two vertices $x$ and $y$ of \VV, the reachability relation is noted ``$\Rightarrow^*$'' such that $x \Rightarrow^* y$ iff $x$ can reach $y$ in \GG.
 \end{definition}
 
-\begin{definition}
+\begin{remark}
+    The relation $\Rightarrow^*$ is in fact the transitive closure of the binary relation $\Rightarrow$ defining edges in a graph.
+\end{remark}
+
+\begin{definition}[SCC]
     Let \GG = (\VV, \EE) be an directed graph.
-$\mathcal{C} \subseteq \mathcal{V}$ is a strongly connected component of \GG if:
-\begin{gather*}
-    \forall x, y \in \mathcal{C}, (x \Rightarrow y) \wedge (y \Rightarrow x)
-\end{gather*}
-\textit{i.e.} there is a path between every $x$ and $y$ in $\mathcal{C}$.\\
+$\mathcal{C} \subseteq \mathcal{V}$ is a strongly connected component (SCC) of \GG if:
+\begin{equation*}
+    \forall x, y \in \mathcal{C}, (x \Rightarrow^* y) \wedge (y \Rightarrow^* x)
+\end{equation*}
+\textit{i.e.} there is a path between every $x$ and $y$ in $\mathcal{C}$.
+\BlankLine
+$\mathcal{C}$ is maximal, or $\mathcal{C}$ is a maximal SCC of \GG if there is no other SCC containing $\mathcal{C}$, \textit{i.e.} if:
+\begin{equation*}
+    \forall \mathcal{X}, (\mathcal{C} \subseteq \mathcal{X}) \wedge (\forall x, y \in \mathcal{X}, (x \Rightarrow^* y) \wedge (y \Rightarrow^* x)) \Longrightarrow \mathcal{C} = \mathcal{X}
+\end{equation*}
+\end{definition}
 
+\begin{definition}(Strong connectedness)
+Let \GG = (\VV, \EE) be a directed graph. \GG is strongly connected if \VV is a SCC.
 \end{definition}
@@
         \ctikzfig{example2}
-        \ednote{sm}{The notion of connectedness has not been introduced, and what's the relationship to SCCs?}
         \subcaption{Strongly connected graph}
@@
     \ctikzfig{exampleSCC}
-\ednote{sm}{%
-  ``Set of SCC'' is confusing: an SCC is a set, and you are not talking about sets of sets here.
-}
-    \caption{Example of a graph where each colored set of node is a maximal set of SCC\label{fig:exampleSCC}}
+    \caption{Example of a graph where each colored set of node is a -- maximal -- SCC\label{fig:exampleSCC}}
 \end{figure}
@@
 
-\subsection{Order of traversal and backtracking edges}
-\subsubsection{DFS and \textit{num} value}
-Tarjan's SCC algorithm\edmargin{sm}{not mentioned yet} basically lies on a depth-first search. The figure \ref{fig:dfs} shows an example of a DFS traversal on a simple directed graph.
-\begin{figure}[!h]
-    \ctikzfig{dfs}
-    \caption{Example of a DFS\label{fig:dfs}}
-\end{figure}
-
-The previously mentioned figure also displays in red the \texttt{num} value which represents the order in which the nodes are visited in the graph during the DFS.
-
-\subsubsection{Backtracking edges}
-\begin{definition}
-    Given a graph \GG and an order of traversal in this graph, \textit{i.e.} each node of \GG has a unique value $\texttt{num} \in [0,~|\mathcal{V}|]~$ and two nodes $u$ and $v$, there is a backtracking edge from $v$ to $u$ if:
-    \begin{equation*}
-        \left\{
-            \begin{array}{l}
-                \texttt{u.num} < \texttt{v.num}\\
-                v~\Rightarrow~u,~\textit{i.e. }~(v, u) \in \mathcal{E}\\
-            \end{array}
-        \right.
-    \end{equation*}
-    In this case, the backtracking edge from $v$ to $u$ is represented by $v \hookrightarrow u$.\edmargin{sm}{add example for the graph of Fig.5}\\      
-\end{definition}
-
-\subsection{Lowlink value}
-\subsubsection{Definition}
-Informally, the \texttt{lowlink} value of a node represents the \texttt{num} value of the attachment node of their SCC, \textit{i.e.} the \texttt{num} value of the entrance node in the corresponding SCC.\\
-
-A more formal definition would be\edmargin{sm}{is?} the following:
-\begin{definition}
-    Let $u$ be a node.
-    \begin{equation*}
-        \texttt{u.lowlink} = \min \{\texttt{w.num}~|~\exists~v\in \mathcal{V},~u \Rightarrow v \hookrightarrow w\}
-    \end{equation*}
-\end{definition}
-
-\subsubsection{Example}
-Let \GG be the graph given in fig.
-\ref{fig:exlowlink}.\edmargin{sm}{``Figure'' vs.\ ``figure'' vs.\ ``Fig.'': usually capitalized, make it uniform.
-And why not use the same graph as before?} The order of traversal of the graph is given by the value \textit{num} for each node of \GG.
-The \textit{lowlink} value is also displayed.
-
-\begin{figure}[!h]
-    \ctikzfig{lowlink}
-    \ednote{sm}{couple of value $\rightarrow$ pair of values}
-    \caption{A DFS was performed through \GG from $x_1$ and next to each node is represented the couple of value (\textit{num}, \textit{lowlink})\label{fig:exlowlink}}
-\end{figure}
-
-Now, backtracking edges can be highlighted w.r.t. the order of traversal. In fig. \ref{fig:exlowlink2}, they are represented as red dashed arrows.\\
-Knowing the backtracking edges, all \textit{lowlink} values can be computed\footnote{In fact, they are refreshed during the DFS.}.\\
-Let us take $x_8$ as an example: its \textit{lowlink} is equal to 2, which actually means that $x_6$\footnote{Because $\texttt{x6.num} = 2$} is its anchor -- or attachment node -- in their\edmargin{sm}{reference?} SCC, namely the green one.
-Indeed, $x_8$ is alone in its equivalence class\footnote{for the relation $\Rightarrow$},\edmargin{sm}{I don't understand the notion of equivalence, e.g.
-we have $x_8 \Rightarrow^* x_9 \Rightarrow^* x_8$.} and from all nodes linked by one of the backtracking edges of $x_8$, $x_6$ has the minimum value \textit{num}.
-Likewise, $x_3$ is its own attachment node in the green SCC since it is the first node visited when performing the DFS.
-
-\begin{figure}[!h]
-    \ctikzfig{lowlink2}
-    \caption{Same graph as in fig. \ref{fig:exlowlink} whose backtracking edges have been represented with red dashed arrows and SCCs have been highlighted\label{fig:exlowlink2}}
-\end{figure}
-
-Then, SCCs can be easily found, namely $\{x_1\}$ and $\{x_2,x_3,x_4,x_5,x_6,x_7,x_8, x_9\}$, as shown in fig. \ref{fig:exlowlink2}.
-
-\pagebreak
-\section{Tarjan's algorithm}
-\ednote{sm}{%
-  Why present Tarjan's algorithm, which is not considered further here?
-Perhaps because it is the best known algorithm for computing SCCs?
-But then it should be compared with the set-based algorithm that this project is about?
-}
-Tarjan's algorithm is an efficient on-the-fly SCC computing algorithm \cite{tarjanMerz}.
-It basically perfoms a DFS while updating the \textit{num} and \textit{lowlink} values.
-All nodes are stored in a stack during the traversal until a backtracking edge is found.
-In this case, the \textit{lowlinks} are computed and all nodes are unstacked and saved in a SCC until a node verifying the equality between its \textit{num} and its \textit{lowlink} -- which has to occur -- is found.
-Then, the DFS goes on.
-The whole process is written in the following algorithm \ref{alg:tarjan}.\\
-It can be shown that every node and edge are visited only once so the algorithm can achieve a linear complexity, \textit{i.e.} $\mathcal{O}(|\mathcal{V}|+|\mathcal{E}|)$.
-
-\subsection{Description of Tarjan's algorithm}
-\begin{algorithm}[H]\label{alg:tarjan}
-    \SetAlgoLined
-    \KwData{A graph \GG = (\VV, \EE)}
-    \KwResult{A partition \texttt{SCCs} of \VV where each element of \texttt{SCCs} is a SCC of \GG}
-    Initialize an empty stack \texttt{R}\;
-    Initialize an empty set \texttt{SCCs}\;
-    Let $\texttt{num} := 0$\;
-    \ForAll{$v \in \mathcal{V}$}{
-        \If{\texttt{v.num} is undefined}{
-            SCC(v)\;
-        }
-    }
-    \SetKwProg{Function}{function}{}{}
-    \Function{SCC: $v \in \mathcal{V} \rightarrow \texttt{None}$}{
-        \texttt{v.num = num}\;
-        \texttt{v.lowlink = num}\;
-        increment \texttt{num}\;
-        Push $v$ in \texttt{R}\;
-        \texttt{v.onStack = true}\;
-        \ForAll{$w \in \text{POST}(v)$\quad\textcolor{red}{sm: POST undefined?}}{
-            \If{\texttt{w.num} is undefined}{
-                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.lowlink})$ \quad\textcolor{red}{w.lowlink undefined?}\;
-            }
-            \ElseIf{\texttt{w.onStack}}{
-                \texttt{v.lowlink} = $\min(\texttt{v.lowlink, w.num})$\;
-            }
-        }
-        \If{\texttt{v.lowlink = v.num}}{
-            Initialize an empty set \texttt{currentSCC}\;
-            \Repeat{$\texttt{v.num} \neq \texttt{w.num}$\quad\textcolor{red}{sm: w.lowlink?}}{
-                Let $w := \texttt{R.pop()}$\;
-                \texttt{w.onStack = false}\;
-                $\texttt{currentSCC} = \texttt{currentSCC} \cup \{w\}$\;
-            }
-            $\texttt{SCCs} = \texttt{SCCs} \cup \texttt{currentSCC}$\;
-        }
-    }
-    \caption{Tarjan's algorithm}
-\end{algorithm}
 \pagebreak
@@
 \subsection{Informal proof}
-Note that this proof is said informal only because it is not formally automated.\edmargin{sm}{checked by a mechanized proof assistant?} Both logical and mathematical arguments developed below are absolutely relevant.
+Note that this proof is said informal only because it is not checked by a mechanized proof assistant. Both logical and mathematical arguments developed below are absolutely relevant.
 
@@
     \begin{equation*}
-        \biguplus_{v \in \texttt{R}} \mathcal{S}(v) = \textsc{Live} := \textsc{Visited} \setminus \textsc{Explored}
+        \biguplus_{r \in \texttt{R}} \mathcal{S}(r) = \textsc{Live} := \textsc{Visited} \setminus \textsc{Explored}
     \end{equation*}
@@
 
-\begin{corollary}\label{cor:cor1}
+\begin{corollary}[Strong version]\label{cor:cor1}
     \begin{equation*}
@@
 \begin{proof}
-    Let $v \in \textsc{Live} = \displaystyle{\biguplus_{v \in \texttt{R}} \mathcal{S}(v)}$.\edmargin{sm}{Avoid using $v$ as a free variable and a bound variable in the same formula.
-Rather use $r$ for the bound variable, as in the lemma.} $v$ is in a unique partial SCC $\mathscr{S} := \mathcal{S}(v)$.
+    Let $v \in \textsc{Live} = \displaystyle{\biguplus_{r \in \texttt{R}} \mathcal{S}(r)}$.
+$v$ is in a unique partial SCC $\mathscr{S} := \mathcal{S}(v)$.
 Because of lemma \ref{lemma:disjointness}, there cannot exist $x \neq y \in \texttt{R}$ s.t.
@@
 
-\begin{corollary}\label{cor:cor2}
+\begin{corollary}[Weak version]\label{cor:cor2}
     \begin{equation*}
@@
 \begin{proof}
-    Holds because of corollary \ref{cor:cor1}.\edmargin{sm}{In fact, just a weakening of Cor.1, why state it separately?}
+    Holds because of corollary \ref{cor:cor1}.
 \end{proof}
@@
 \begin{remark}
-    In the algorithm \ref{alg:seqsetbased}, this property is held\edmargin{sm}{maintained?} by L.16-18.
+    In the algorithm \ref{alg:seqsetbased}, this property is maintained by L.16-18.
 These lines also illustrate how the algorithm ``reads'' the SCCs.
@@
 \subsection{Formal proof}
-Since the informal proof seems to be consistant,\edmargin{sm}{consistent, or rather convincing?} the formal -- automated -- proof\edmargin{sm}{be careful: it is checked automatically, but found by the user!} can be written in Isabelle (HOL) based on the basis of the reasoning developed above.
+Since the informal proof seems to be convincing, the formal -- checked automatically -- proof can be written in Isabelle (HOL) based on the basis of the reasoning developed above.
 
@@
 The first definitions should be the different structures used in the algorithm.
-In particular, a record containing all the sets needed and described in the pseudo-code of algorithm \ref{alg:seqsetbased} :\edmargin{sm}{I'd explain what 'v stands for.
-Also, somewhat inconsistent spacing.}
+In particular, a record containing all the sets needed and described in the pseudo-code of algorithm \ref{alg:seqsetbased}.
+The environment has a generic type parameter, which is used to represent the type of the nodes in the graph (often integers):
 
@@
 
-The following lines define a graph structure and some useful natural relations :\edmargin{sm}{Is there an advantage to introducing the graph via ``successors'' rather than a relation as in the informal presentation?}
+The following lines define a graph structure and some useful natural relations:
 
@@
 }
+
+The use of \texttt{successors} instead of an adjacency matrix, for instance, is a consequence of the fact that the algorithm is only concerned with the topological ordering of the nodes.
+For instance, nodes can represent integers, logical propositions or sets of states in a proving system for example.
+
+\subsubsection{Reachability}
+Now that graphs are defined, the reachability can be defined. Defining an edge is simply some rewriting of being a successor of one node.
+
 \isa{
@@
 }
+\BlankLine
+\BlankLine
+\BlankLine
+\BlankLine
+
+\noindent
+Regarding the reachability binary relation, a choice has to be made since there are several ways to define it.
+In particular, there are two possible keywords, {\color{isa_blue}\texttt{inductive}} and {\color{isa_blue}\texttt{fun}}, respectively for an inductive or recursive definition.
+If both definitions are valid, the inductive one is kept for the following reasons.
+Although a recursive definition allows one to do some rewriting in the middle of terms, a recursive definition expresses both the positive and negative information\footnote{In this case, the positive information designates the fact of being reachable and the negative information designates the fact of not being reachable.} whereas the inductive one only expresses the positive information directly.
+Therefore, with an inductive definition, the negative information has to be proved.
+One would be right to argue that it would be more convenient to be able to tell without proving it that two nodes are not reachable from each other, but this does not interest us for the following.
+Another important point is that there is no datatype for a recursive definition, especially in this case with the transitive closure of the $\Rightarrow^*$ relation.
+Thus, the choice of the inductive definition is not a choice of simplicity but of necessity.
+Lastly, Isabelle will generate a simple inductive rule for the proofs split into the reflexive case, which stands in the definition, and the transitive case, which has to be proved.
 
@@
 
-Those two relations are \texttt{edge}, which simply translates the property for two nodes of being linked by an edge and \texttt{reachable},\edmargin{sm}{I think it's worthwhile to explain ``inductive''.} which is the binary relation $\Rightarrow^*$ defined in section \ref{sec:directedgraphs}.\\
 In order to be able to use those relations in the proofs later, it is essential to prove a list of lemmas, namely all the different natural properties that Isabelle cannot deduce\footnote{That is an abuse of language.
@@
 The goal is to formalize it so Isabelle is logically able to both use and simplify some results in the proofs.} from nothing\footnote{There is actually a theorem fetcher that is particularly useful to find a basic set of lemmas.}.
-For instance, the following lemmas are essential.\edmargin{sm}{Given that the first lemma is proved by auto, is it really useful?}
-
-\isa{
-\lm{} reachable\_edge: "edge {\color{isa_dark_blue}x y} $\Longrightarrow$ reachable {\color{isa_dark_blue}x y}"\\
-$~~~~~${\color{isa_blue}by} auto\\
-}
-Mathematical writing: $\forall x, \forall y, x \Rightarrow y \Longrightarrow x \Rightarrow^* y$
+For instance, the following lemmas are essential.
 
@@
 
-\subsubsection{Ordering relation}
-In the proof, a ordering\edmargin{sm}{precedence?
-In order to avoid the impression that one orders stacks?} relation\footnote{In fact, a total order is being defined on stacks.} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack.
+\subsubsection{Equivalence realtion and graph partition}
+
+\isa{
+{\color{isa_blue}function} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
+$~~~~~$"{\color{isa_dark_blue}unite} {\color{isa_dark_green}v w e} =\\
+$~~~~~~~~~~$({\color{isa_blue}if} ($\mathcal{S}$ {\color{isa_dark_green}e v} = $\mathcal{S}$ {\color{isa_dark_green}e w}) {\color{isa_blue}then} {\color{isa_dark_green}e}\\
+$~~~~~~~~~~${\color{isa_blue}else let} {\color{isa_dark_green}r} = hd(stack {\color{isa_dark_green}e});\\
+$~~~~~~~~~~~~~~~~~~~${\color{isa_dark_green}r'}= hd(tl(stack {\color{isa_dark_green}e}));\\
+$~~~~~~~~~~~~~~~~~~~${\color{isa_dark_green}joined} = $\mathcal{S}$ {\color{isa_dark_green}e r} $\cup~\mathcal{S}$ {\color{isa_dark_green}e r'};\\
+$~~~~~~~~~~~~~~~~~~~${\color{isa_dark_green}e'}= {\color{isa_dark_green}e'} \env{stack := tl(stack {\color{isa_dark_green}e}), $\mathcal{S}$ := ($\lambda$n.
+{\color{isa_blue}if} n $\in$ {\color{isa_dark_green}joined} {\color{isa_blue}then} {\color{isa_dark_green}joined} {\color{isa_blue}else} $\mathcal{S}$ {\color{isa_dark_green}e n})}\\
+$~~~~~~~~~~${\color{isa_blue}in} {\color{isa_dark_blue}unite} {\color{isa_dark_green}v w e'})"\\
+$~~~~~${\color{isa_blue}by} pat\_completeness auto
+}
+
+\subsubsection{Ordering relation}\label{sec:orderingrelation}
+In the proof, a precedence relation\footnote{In fact, a total order is being defined on stacks.} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack.
 Let $x$ and $y$ be two nodes and $R$ be a stack.
-Informally, $x$ precedes $y$ in $R$ if $x$ was pushed in $R$ before $y$ (see figure \ref{fig:stackorder}).\edmargin{sm}{The formal definition is the opposite of what's shown in the figure.}
+Informally, $x$ precedes $y$ in $R$ if $y$ was pushed in $R$ before $x$ (see \textsc{Figure} \ref{fig:stackorder}).
 
@@
 
-The idea is to later use the following property: if $x \preceq y~\text{in}~xs$, then $x \Rightarrow^* y$.\edmargin{sm}{Probably the other way around.}\\
+The idea is to later use the following property: if $x \preceq y~\text{in}~xs$, then $y \Rightarrow^* x$.\\
 It is defined in Isabelle as follows:\\
@@
 
-All the different properties (\textit{i.e.} lemmas) which follow this definition in the Isabelle implementation are detailed in the natural mathematical writing below:\\
+All the different properties (\textit{i.e.} lemmas) which follow this definition in the Isabelle implementation are detailed in the natural mathematical writing in \nameref{appendix} .\\
 
+\subsubsection{Implementation of the algorithm}
+
+Now that the environment is set up, the actual algorithm -- seen as a function -- can be implemented.\\
+Since Isabelle does not support loops, the implementation will be split into two mutually recursive functions.
+The main function is called \texttt{dfs} and takes its name after the Depth First Search algorithm because the algorithm \ref{alg:seqsetbased} roughly consists in a deep traversal of a graph.
+The second function is called \texttt{dfss} and represents the \textit{while} loop of the algorithm \ref{alg:seqsetbased}.
+The two functions are mutually recursive because they recursively call each other.
+In particular, \texttt{dfss} will call either itself or \texttt{dfs}, depending on the case.
+
+\isa{
+    {\color{isa_blue}{function}} dfs :: "\generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \and{}\\
+    $~~~~~~~~~~$dfss:: "\generic{v} $\Rightarrow$ \generic{v} set $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where\\
+    "{\color{isa_dark_blue}dfs} {\color{isa_dark_green}v e} =\\
+    $~~~~~$({\color{isa_blue}{let}} {\color{isa_dark_green}e1} = {\color{isa_dark_green}e}\env{visited := visited e $\cup$ \{{\color{isa_dark_green}v}\}, stack := ({\color{isa_dark_green}v} \# stack {\color{isa_dark_green}e})};\\
+    $~~~~~~~~~~${\color{isa_dark_green}e'} = {\color{isa_dark_blue}dfss} {\color{isa_dark_green}v} (successors {\color{isa_dark_green}v}) {\color{isa_dark_green}e1}\\
+    $~~~~~${\color{isa_blue}{in if}} {\color{isa_dark_green}v} = hd(stack {\color{isa_dark_green}e'})\\
+    $~~~~~~~~~~${\color{isa_blue}{then}} {\color{isa_dark_green}e'}\env{sccs:=sccs {\color{isa_dark_green}e'} $\cup$ {$\mathcal{S}$ {\color{isa_dark_green}e' v}}, explored:=explored {\color{isa_dark_green}e'} $\cup$ ($\mathcal{S}$ {\color{isa_dark_green}e' v}), stack:=tl(stack {\color{isa_dark_green}e'})}\\
+    $~~~~~~~~~~${\color{isa_blue}else} {\color{isa_dark_green}e'})"\\
+    | "{\color{isa_dark_blue}dfss} {\color{isa_dark_green}v vs e} =\\
+    $~~~~~$({\color{isa_blue}if} {\color{isa_dark_green}vs} = \{\} {\color{isa_blue}then} {\color{isa_dark_green}e}\\
+    $~~~~~${\color{isa_blue}else} ({\color{isa_blue}let} {\color{isa_dark_green}w} = SOME {\color{isa_dark_green}x}. {\color{isa_dark_green}x} $\in$ {\color{isa_dark_green}vs}\\
+    $~~~~~~~~~~${\color{isa_blue}in} ({\color{isa_blue}let} {\color{isa_dark_green}e'} = ({\color{isa_blue}if} {\color{isa_dark_green}w} $\in$ explored {\color{isa_dark_green}e} {\color{isa_blue}then} {\color{isa_dark_green}e}\\
+    $~~~~~~~~~~~~~~~~~~~${\color{isa_blue}else if} {\color{isa_dark_green}w} $\notin$ visited {\color{isa_dark_green}e} {\color{isa_blue}then} {\color{isa_dark_blue}dfs} {\color{isa_dark_green}w e}\\
+    $~~~~~~~~~~~~~~~~~~~${\color{isa_blue}else} unite {\color{isa_dark_green}v w e})\\
+    $~~~~~~~~~~~~~~${\color{isa_blue}in} {\color{isa_dark_blue}dfss} {\color{isa_dark_green}v} ({\color{isa_dark_green}v} - \{{\color{isa_dark_green}w}\}) {\color{isa_dark_green}e'})))"\\
+    $~~~${\color{isa_blue}by} pat\_completeness (force+)\\
+}
+
+The two last keywords require explanations as well : \texttt{pat\_completeness} stands for \textit{pattern completeness} and ensures that there is no missing patterns.
+The keyword \texttt{force} is used\footnote{\texttt{force} is more aggressive in instantiation and seems to find the right instance.} to help Isabelle know -- by proving it -- that both \texttt{dfs} and \texttt{dfss} are actually functions and that those functions are well defined with respect to the usual logical and mathematical meaning.
+
+\pagebreak
+
+\section{Appendix}\label{appendix}
+\subsection{Some lemmas}
+Those lemmas refer to the precedence relation introduced in \textsc{Section} \ref{sec:orderingrelation}.
+\BlankLine
 Let $x, y, z$ be three nodes, and let $xs, ys, zs$ be three lists of nodes representing stacks.\\
@@
 However, $xs$ in not seen as the set representing $xs$ since an element may occur several times in a stack.\\
-The operator $@$ denotes the concatenation and operates on two lists: $[x_0, \dots, x_n] @ [y_0, \dots, y_m] = [x_1, \dots, x_n, y_1, \dots, y_m]$.\edmargin{sm}{$x_0$, $y_0$}\\
+The operator $@$ denotes the concatenation and operates on two lists: $[x_0, \dots, x_n] @ [y_0, \dots, y_m] = [x_0, \dots, x_n, y_0, \dots, y_m]$.\\
 
@@
     (ii)
-$y \in [x@ xs \Longrightarrow (x \in xs)\wedge(y \in xs)$
-\edmargin{sm}{I don't understand, $y=x \lor y \in xs$? (And ``$[x]$''.)}
+$y \in [x] @ xs \Longrightarrow x \preceq y~\text{in}~([x] @ xs)$
 \end{flushleft}
@@
 \begin{flushleft}
-    (xi)
+    (xi)(transitivity)
     $(x \preceq y~\text{in}~xs) \wedge (y \preceq z~\text{in}~xs) \wedge \underset{\text{all elements of $xs$ are distinct}}{\underbrace{(\forall~0 \leq i < j \leq \text{length}(xs), xs[i] \neq xs[j])}} \Longrightarrow x \preceq z~\text{in}~xs$
@@
 
-\ednote{sm}{%
-  Perhaps it would be enough to just say that one builds a library of lemmas and display a few of them?
-}
+\begin{flushleft}
+    (xi)(antisymmetry)
+    $(x \preceq y~\text{in}~xs) \wedge (y \preceq x~\text{in}~xs) \wedge \underset{\text{all elements of $xs$ are distinct}}{\underbrace{(\forall~0 \leq i < j \leq \text{length}(xs), xs[i] \neq xs[j])}} \Longrightarrow x = y$
+\end{flushleft}
 
 % continue with precedes_append_right_iff
-
-\subsubsection{Implementation of the algorithm}
-
-Now that the environment is set up, the actual algorithm -- seen as a function -- can be implemented.\\
-Since Isabelle does not support loops, the implementation will be split into two mutually recursive functions.
-The main function is called \texttt{dfs} and takes its name after the Depth First Search algorithm because the algorithm \ref{alg:seqsetbased} roughly consists in a deep traversal of a graph.
-The second function is called \texttt{dfss} and represents the \textit{while} loop of the algorithm \ref{alg:seqsetbased}.
-The two functions are mutually recursive because they recursively call each other.
-In particular, \texttt{dfss} will call either itself or \texttt{dfs}, depending on the case.
-
-\isa{
-    {\color{isa_blue}{function}} dfs :: "\generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \and{}\\
-    $~~~~~~~~~~$dfss:: "\generic{v} $\Rightarrow$ \generic{v} set $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where\\
-    "{\color{isa_dark_blue}dfs} {\color{isa_dark_green}v e} =\\
-    $~~~~~$({\color{isa_blue}{let}} {\color{isa_dark_green}e1} = {\color{isa_dark_green}e}\env{visited := visited e $\cup$ \{{\color{isa_dark_green}v}\}, stack := ({\color{isa_dark_green}v} \# stack {\color{isa_dark_green}e})};\\
-    $~~~~~~~~~~${\color{isa_dark_green}e'} = {\color{isa_dark_blue}dfss} {\color{isa_dark_green}v} (successors {\color{isa_dark_green}v}) {\color{isa_dark_green}e1}\\
-    $~~~~~${\color{isa_blue}{in if}} {\color{isa_dark_green}v} = hd(stack {\color{isa_dark_green}e'})\\
-    $~~~~~~~~~~${\color{isa_blue}{then}} {\color{isa_dark_green}e'}\env{sccs:=sccs {\color{isa_dark_green}e'} $\cup$ {$\mathcal{S}$ {\color{isa_dark_green}e' v}}, explored:=explored {\color{isa_dark_green}e'} $\cup$ ($\mathcal{S}$ {\color{isa_dark_green}e' v}), stack:=tl(stack {\color{isa_dark_green}e'})}\\
-    $~~~~~~~~~~${\color{isa_blue}else} {\color{isa_dark_green}e'})"\\
-    | "{\color{isa_dark_blue}dfss} {\color{isa_dark_green}v vs e} =\\
-    $~~~~~$({\color{isa_blue}if} {\color{isa_dark_green}vs} = \{\} {\color{isa_blue}then} {\color{isa_dark_green}e}\\
-    $~~~~~${\color{isa_blue}else} ({\color{isa_blue}let} {\color{isa_dark_green}w} = SOME {\color{isa_dark_green}x}. {\color{isa_dark_green}x} $\in$ {\color{isa_dark_green}vs}\\
-    $~~~~~~~~~~${\color{isa_blue}in} ({\color{isa_blue}let} {\color{isa_dark_green}e'} = ({\color{isa_blue}if} {\color{isa_dark_green}w} $\in$ explored {\color{isa_dark_green}e} {\color{isa_blue}then} {\color{isa_dark_green}e}\\
-    $~~~~~~~~~~~~~~~~~~~${\color{isa_blue}else if} {\color{isa_dark_green}w} $\notin$ visited {\color{isa_dark_green}e} {\color{isa_blue}then} {\color{isa_dark_blue}dfs} {\color{isa_dark_green}w e}\\
-    $~~~~~~~~~~~~~~~~~~~${\color{isa_blue}else} unite {\color{isa_dark_green}v w e})\\
-    $~~~~~~~~~~~~~~${\color{isa_blue}in} {\color{isa_dark_blue}dfss} {\color{isa_dark_green}v} ({\color{isa_dark_green}v} - \{{\color{isa_dark_green}w}\}) {\color{isa_dark_green}e'})))"\\
-    $~~~${\color{isa_blue}by} pat\_completeness (force+)\\
-}
-
-\ednote{sm}{Should also show the Isabelle representation of unite.}
-
-The two last keywords require explanations as well : \texttt{pat\_completeness} stands for \textit{pattern completeness} and ensures that there is no missing patterns.
-The keyword \texttt{force} is used\footnote{Here, \texttt{auto} cannot terminate because of the mutual recursion.}\edmargin{sm}{Not sure if mutual recursion is a problem here: force is more aggressive in instantiation and seems to find the right instance.} to help Isabelle know -- by proving it -- that both \texttt{dfs} and \texttt{dfss} are actually functions and that those functions are well defined with respect to the usual logical and mathematical meaning.
 

### 2022-05-26 LaTeX/report.tex
Updates on the report

@@
 % Isabelle keywords
+\newcommand{\stackprec}[3]{\texttt{#1}~\preceq~\texttt{#2}~\text{in}~\texttt{#3}}
 \newcommand{\apply}{{\color{isa_red}{apply}}}
@@
 \newcommand{\blue}[1]{{\color{isa_dark_blue}{#1}}}
+\newcommand{\bblue}[1]{{\color{isa_blue}{#1}}}
 \newcommand{\green}[1]{{\color{isa_dark_green}{#1}}}
@@
 
-\subsection{Isabelle by example}
-The following example is a good introduction to the use of Isabelle.\edmargin{sm}{I'd explain what 'v stands for.}
-
 \pagebreak
@@
 \begin{definition}[SCC mapping]
-    In the following algorithm, the SCCs are progressively tracked in a collection of disjoint\edmargin{sm}{We may have $\mathcal{S}(v) = \mathcal{S}(w)$ for $v \neq w$?!} sets through a map $\mathcal{S} : \mathcal{V} \longrightarrow \mathcal{P}(\mathcal{V})$, where $\mathcal{P}(\mathcal{V})$ is the powerset of \VV, s.t.
+    In the following algorithm, the SCCs are progressively tracked in a collection of disjoint sets through a map $\mathcal{S} : \mathcal{V} \longrightarrow \mathcal{P}(\mathcal{V})$, where $\mathcal{P}(\mathcal{V})$ is the powerset of \VV, s.t.
 the following invariant is maintained:
-    \begin{equation}\label{invariant}
+    \begin{equation}\label{invariant:Sdisjoint}
         \forall v, w \in \mathcal{V}, w\in \mathcal{S}(v) \Longleftrightarrow \mathcal{S}(v) = \mathcal{S}(w)
@@
 \end{definition}
+
+\begin{remark}
+    In the following, the same notation $\mathcal{S}$ will be used to denote both the function defined above and the induced equivalence relation\footnote{For the relation $(x, y) \mapsto x \Rightarrow^* y ~\land~ y \Rightarrow^* x $} since $\mathcal{S}$ associates to each node its class of equivalence.
+\end{remark}
 
@@
 \begin{definition}[SCC union]
-    Let \textsc{Unite} be the function taking as parameters a map $\mathcal{S}$ as defined previously and two vertices $u$ and $v$ of $\mathcal{V}$ such that $\textsc{Unite}(\mathcal{S}, u, v)$ merges the two mapped sets $\mathcal{S}(u)$ and $\mathcal{S}(v)$ and maintains the invariant (\ref{invariant}) by updating the function $\mathcal{S}$.
+    Let \textsc{Unite} be the function taking as parameters a map $\mathcal{S}$ as defined previously and two vertices $u$ and $v$ of $\mathcal{V}$ such that $\textsc{Unite}(\mathcal{S}, u, v)$ merges the two mapped sets $\mathcal{S}(u)$ and $\mathcal{S}(v)$ and maintains the invariant (\ref{invariant:Sdisjoint}) by updating $\mathcal{S}$.
 \end{definition}
@@
 \begin{proof}
-    Let $x \in \mathcal{V}$ be the following node to be visited during the execution of the algorithm \ref{alg:seqsetbased}: $x$ is pushed in \texttt{R}.
-Let $y \in \textsc{Post}(x)$.\edmargin{sm}{The lemma assumes $y \in \texttt{R}$, is $y \in \textsc{Post}(x)$ an additional assumption?
-Why is it justified?} There are two cases:\\
-    \begin{itemize}
-        \item y has not been visited yet, \textit{i.e.} $y \notin \textsc{Visited}$.
-Thus, a DFS-like traversal is performed from $y$, so $y$ is pushed in \texttt{R} and $\mathcal{S}(y) = \{y\}$ because $y$ is alone in its equivalence class for the moment since it has not been visited yet.\\
-        Therefore, $\mathcal{S}(x) \cap \mathcal{S}(y) = \varnothing$.\edmargin{sm}{I don't see an argument why $y \notin \mathcal{S}(x)$?}
+    Let $x, y \in \texttt{R}$, $x \neq y$. We can assume w.l.o.g that $\stackprec{y}{x}{R}$.
+    If $y \notin \mathcal{S}(x)$, then $\mathcal{S}(x)$ is disjoint from $\mathcal{S}(y)$.
+    \BlankLine
+    Otherwise, let us assume that $y \in \mathcal{S}(x)$, \textit{i.e.} $\mathcal{S}(x) = \mathcal{S}(y)$.
+Then, the \textit{while}\footnote{See \nameref{alg:seqsetbased}} loop will unstack all nodes of \texttt{R} until the successor of $x$ in \texttt{R} is reached and add them with \textsc{Unite} to $\mathcal{S}(x)$, so that the only representative of $\mathcal{S}(x)$ on \texttt{R} is $x$.
+\end{proof}
 
-        \item y has already been visited, \textit{i.e.} $y \in \textsc{Visited}$.
-Then, $y$ was already pushed in \texttt{R} before $x$.
-Let $(x_i)_{1\leq i \leq n}$ be the first nodes of the stack s.t.
-$x_0 = x$ and $x_n = y$.\edmargin{sm}{Should explain the order of the stack.}\\
-        In order to avoid writing $\texttt{R} = [\ldots, y, \ldots, x]$, let us define $\widetilde{R}$ the stack containing the first $n$ nodes in \texttt{R}, s.t.
-$\widetilde{R} = [y, \ldots, x] = [x_n, \ldots, x_0]$.\\
-        Let us consider the worst case, \textit{i.e.} when\edmargin{sm}{Why is this the worst case?}
-        \begin{equation*}
-            \forall~1 \leq i \leq n,~\mathcal{S}(x_i) = \{x_i\}
-        \end{equation*}
-        So, the while loop has to go down to $y$ because all partial SCCs are disjoint. As the length of the stack \texttt{R} is bounded by $|\mathcal{V}|$, the algorithm terminates.\\
-        $x_0$ is first unstacked and both $\mathcal{S}(x_0)$ and $\mathcal{S}(\texttt{R.top()}) = \mathcal{S}(x_1)$ are then united.
-The current state of $\mathcal{S}$ and $\widetilde{R}$ is:\edmargin{sm}{Ambiguity between $\mathcal{S}$ as a mapping and as a set of equivalence classes.}
-        \begin{equation*}
-            \left\{
-                \begin{array}{l}
-                    \mathcal{S} = \{\{x_0, x_1\}, \{x_2\}, \ldots, \{x_n\}, \ldots\}\\
-                    \widetilde{R} = [x_n, \ldots, x_1]\\
-                \end{array}
-            \right.
-        \end{equation*}
-        Then, $x_1$ is unstacked and $\mathcal{S}(x_1)$ and $\mathcal{S}(x_2)$ are then united, so that:
-        \begin{equation*}
-            \left\{
-                    \begin{array}{l}
-                        \mathcal{S} = \{\{x_0, x_1, x_2\}, \{x_3\}, \ldots, \{x_n\}, \ldots\}\\
-                        \widetilde{R} = [x_n, \ldots, x_2]\\
-                    \end{array}
-                \right.
-        \end{equation*}
-        Finally (by induction), $\mathcal{S} = \{x_0, \ldots, x_n\}$ and $\widetilde{R} = [y]$, \textit{i.e.} $\mathcal{S}(x)=\mathcal{S}(y)$.
-It is important to notice that $x=x_0, x_1, \ldots, x_{n-1}$ are no longer in the stack, so this operation kept the invariant true.
-    \end{itemize}
-\end{proof}
+\begin{remark}\label{rem:rootSCC}
+    It is worth noting that the representative of a partial — with respect to the final result — SCC is the earliest node of this SCC to be visited regarding the graph traversal.
+This ensures that a representative on the stack is in fact the root of its SCC.
+\end{remark}
 
@@
 So, $\textsc{Live} \subseteq \texttt{R}$.\\
-    By L.6-7 of algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.\\
+    By L.6-7 of Algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.\\
     L.9-10 ensure that no explored node is pushed in \texttt{R}.\\
@@
 \begin{remark}\label{rem:proof}
-    Because \texttt{R} only contains exactly one representative for each partial SCC \edmargin{sm}{I think the fact that this representative is the lowest one in the stack also plays a role here?} (corollary \ref{cor:cor1}), after each step of the main loop -- \textit{i.e.} the DFS -- every partial SCC is actually maximal in the current set of visited nodes.
+    Because \texttt{R} only contains exactly one representative for each partial SCC (corollary \ref{cor:cor1} and remark \ref{rem:rootSCC}), after each step of the main loop -- \textit{i.e.} the DFS -- every partial SCC is actually maximal in the current set of visited nodes.
 \end{remark}
@@
 
-\subsection{Formal proof}
+\subsection{Prerequisites for the formal proof}
 Since the informal proof seems to be convincing, the formal -- checked automatically -- proof can be written in Isabelle (HOL) based on the basis of the reasoning developed above.
 
-\subsubsection{Environment setup}
+\subsubsection{Environment setup}\label{envdef}
 The first definitions should be the different structures used in the algorithm.
@@
 
-\subsubsection{Equivalence realtion and graph partition}
+\subsubsection{Equivalence relation and graph partition}
+In the algorithm \ref{alg:seqsetbased}, the SCCs are progressively tracked in \texttt{sccs} and the equivalence relation $\mathcal{S}$ is updated with the \textsc{Unite} function.
+In Isabelle, a first recursive definition was written as follows:
 
@@
     {\color{isa_blue}function} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
-$~~~~~$"{\color{isa_dark_blue}unite} {\color{isa_dark_green}v w e} =\\
-$~~~~~~~~~~$({\color{isa_blue}if} ($\mathcal{S}$ {\color{isa_dark_green}e v} = $\mathcal{S}$ {\color{isa_dark_green}e w}) {\color{isa_blue}then} {\color{isa_dark_green}e}\\
-$~~~~~~~~~~${\color{isa_blue}else let} {\color{isa_dark_green}r} = hd(stack {\color{isa_dark_green}e});\\
-$~~~~~~~~~~~~~~~~~~~${\color{isa_dark_green}r'}= hd(tl(stack {\color{isa_dark_green}e}));\\
-$~~~~~~~~~~~~~~~~~~~${\color{isa_dark_green}joined} = $\mathcal{S}$ {\color{isa_dark_green}e r} $\cup~\mathcal{S}$ {\color{isa_dark_green}e r'};\\
-$~~~~~~~~~~~~~~~~~~~${\color{isa_dark_green}e'}= {\color{isa_dark_green}e'} \env{stack := tl(stack {\color{isa_dark_green}e}), $\mathcal{S}$ := ($\lambda$n.
-{\color{isa_blue}if} n $\in$ {\color{isa_dark_green}joined} {\color{isa_blue}then} {\color{isa_dark_green}joined} {\color{isa_blue}else} $\mathcal{S}$ {\color{isa_dark_green}e n})}\\
-$~~~~~~~~~~${\color{isa_blue}in} {\color{isa_dark_blue}unite} {\color{isa_dark_green}v w e'})"\\
-$~~~~~${\color{isa_blue}by} pat\_completeness auto
+    "\blue{unite} \green{v w e} =\\
+    $~~~$({\color{isa_blue}if} ($\mathcal{S}$ \green{e v} = $\mathcal{S}$ \green{e w}) \bblue{then} \green{e}\\
+    $~~~$\bblue{else let} \green{r} = hd(stack \green{e});\\
+    $~~~~~~~~~~~~$\green{r'}= hd(tl(stack \green{e}));\\
+    $~~~~~~~~~~~~$\green{joined} = $\mathcal{S}$ \green{e r} $\cup$ $\mathcal{S}$ \green{e r};\\
+    $~~~~~~~~~~~~$\green{e'}= \green{e}\env{\\
+    $~~~~~~~~~~~~~~~$stack := tl(stack \green{e}),\\
+    $~~~~~~~~~~~~~~~\mathcal{S}$ := ($\lambda$ n. \bblue{if} n $\in$ \green{joined} \bblue{then} \green{joined} \bblue{else} $\mathcal{S}$ \green{e n})\\
+    $~~~~~~~~~~~~$}\\
+    $~~~$\bblue{in} \blue{unite} \green{v w e'})"\\
+    $~~~$\bblue{by} pat\_completeness auto
 }
+
+\BlankLine
+
+However, this definition makes the proofs too difficult due to the recursion. An imperative version was therefore written:
+
+\isa{
+    \bblue{definition} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
+    $~~~$"\blue{unite} \green{v w e} $\equiv$\\
+    $~~~~~~$\bblue{let} \green{pfx} = takeWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
+    $~~~~~~~~~~$\green{sfx} = dropWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
+    $~~~~~~~~~~$\green{cc} = $\bigcup$ \{$\mathcal{S}$ \green{e x} $|$ \green{x}. \green{x} $\in$ set \green{pfx} $\cup$ \{hd \green{sfx}\}\}\\
+    $~~~~~~$\bblue{in}  \green{e}\env{$\mathcal{S}$ := $\lambda$\green{x}. \bblue{if} \green{x} $\in$ \green{xx} \bblue{then} \green{cc} \bblue{else} $\mathcal{S}$ \green{e x}, stack := \green{sfx}}"
+}
+
+\BlankLine
+\BlankLine
+
+The idea of this definition is to create a partition of \texttt{stack e = pfx @ sfx} such that \texttt{pfx} contains the nodes which are to be merged into \texttt{$\mathcal{S}$ e w} and \texttt{sfx} contains the root of \texttt{$\mathcal{S}$ e w} followed by the rest of the stack.
+Then, \texttt{cc} -- which stands for \textit{connected component} -- contains all the nodes which are equivalent to \texttt{w} in the sub-graph currently explored.
+The function \texttt{takeWhile} applied to a boolean function \texttt{P}\footnote{\texttt{P :: \generic{a} $\Rightarrow$ bool}} seen as a property and a list \texttt{xs} returns the elements of \texttt{xs} which satisfy \texttt{P} and stops to the first element not satisfying P.
+The function \texttt{dropWhile} is the opposite of \texttt{takeWhile}.
+Both the imperative and recursive versions of \texttt{unite} are equivalent.
 
@@
 
-All the different properties (\textit{i.e.} lemmas) which follow this definition in the Isabelle implementation are detailed in the natural mathematical writing in \nameref{appendix} .\\
+All the different properties (\textit{i.e.} lemmas) which follow this definition in the Isabelle implementation are detailed in the natural mathematical writing in \nameref{appendix}.
+The right part of the notation represents the orders of priority for each operand since $\preceq$ is an infix operator.
 
@@
 
-Now that the environment is set up, the actual algorithm -- seen as a function -- can be implemented.\\
+Now that the environment is set up, the actual algorithm -- seen as a function -- can be implemented.
 Since Isabelle does not support loops, the implementation will be split into two mutually recursive functions.
@@
     $~~~~~~~~~~~~~~${\color{isa_blue}in} {\color{isa_dark_blue}dfss} {\color{isa_dark_green}v} ({\color{isa_dark_green}v} - \{{\color{isa_dark_green}w}\}) {\color{isa_dark_green}e'})))"\\
-    $~~~${\color{isa_blue}by} pat\_completeness (force+)\\
+    $~~~${\color{isa_blue}by} pat\_completeness (force\bblue{+})\\
 }
@@
 The keyword \texttt{force} is used\footnote{\texttt{force} is more aggressive in instantiation and seems to find the right instance.} to help Isabelle know -- by proving it -- that both \texttt{dfs} and \texttt{dfss} are actually functions and that those functions are well defined with respect to the usual logical and mathematical meaning.
+
+\pagebreak
+
+\subsection{Formal proof}
+\subsubsection{General scheme}
+As the algorithm is composed of two mutually recursive functions, the correctness of the algorithm is proved by induction on the environment structure (cf \ref{envdef}).
+Since both \texttt{dfs} and \texttt{dfss} are quite complex, the proof is split into several parts.
+The idea is to prove for each function that its execution given some pre-conditions on the input environment implies some post-conditions on the output environment.
+Then, it has to be made for the mutually recursive calls as well, so that given the same pre-conditions on one function, the pre-conditions on the other function are also satisfied.
+Finally, it has to be proved that if the pre-conditions are satisfied for one function, and if the pre-conditions implie the post-conditions on the other function, then the post-conditions are also satisfied for the first function.
+
+\subsubsection{Well-formedness of the environment}
+The whole proof relies on one big invariant regarding the environment structure.
+It defines the fact for an environment to be well-formed.
+This invariant is a conjunction of several properties and is defined as follows:
+
+\isa{
+    \bblue{definition} wf\_env \where{}\\
+    $~~~$"\blue{wf\_env} \green{e} $\equiv$\\
+    $~~~~~~$distinct (stack \green{e})\\
+    $~~~~\wedge$ set (stack \green{e}) $\subseteq$ visited \green{e}\\
+    $~~~~\wedge$ explored \green{e} $\subseteq$ visited \green{e}\\
+    $~~~~\wedge$ explored \green{e} $\cap$ set (stack \green{e}) = \{\}\\
+    $~~~~\wedge$ ($\forall$ \green{v} \green{w}. \green{w} $\in$ $\mathcal{S}$ \green{e} \green{v} $\longleftrightarrow$ ($\mathcal{S}$ \green{e} \green{v} = $\mathcal{S}$ \green{e} \green{w}))\\
+    $~~~~\wedge$ ($\forall$\green{v} $\in$ set (stack \green{e}).$\forall$ \green{w} $\in$ set (stack \green{e}).\green{v} $\neq$ \green{w} $\longrightarrow$ $\mathcal{S}$ \green{e} \green{v} $\cap$ $\mathcal{S}$ \green{e} \green{w} = \{\})\\
+    $~~~~\wedge$ ($\forall$ \green{v}. \green{v} $\notin$ visited \green{e} $\longrightarrow$ $\mathcal{S}$ \green{e} \green{v} = \{\green{v}\})\\
+    $~~~~\wedge$ $\bigcup$ \{$\mathcal{S}$ \green{e} \green{v} $|$ \green{v}. \green{v} $\in$ set (stack \green{e})\} = visited \green{e} - explored \green{e}\\
+    $~~~~\wedge$ ($\forall$ \green{x} \green{y}. \green{x} $\preceq$ \green{y} in stack \green{e} $\longrightarrow$ reachable \green{y} \green{x})\\
+    $~~~~\wedge$ ($\forall$ \green{x}. is\_subscc ($\mathcal{S}$ \green{e} \green{x}))\\
+    $~~~~\wedge$ ($\forall$ \green{x} $\in$ explored \green{e}. $\forall$ \green{y}. reachable \green{x} \green{y} $\longrightarrow$ \green{y} $\in$ explored \green{e})\\
+    $~~~~\wedge$ ($\forall$ \green{S} $\in$ sccs \green{e}. is\_scc \green{S})"
+}
+
+\BlankLine
+\BlankLine
+
+Let us take a closer look to this invariant, taken in the same order as the definition above:
+\begin{itemize}
+    \item[$\bullet$] First, the stack is a set of distinct elements.
+    \item[$\bullet$] All elements of the stack are visited.
+    \item[$\bullet$] The set of explored nodes is a subset of the set of visited nodes.
+    \item[$\bullet$] Explored nodes cannot be in the stack.
+    \item[$\bullet$] The three next properties are about the equivalence relation $\mathcal{S}$.
+    \item[$\bullet$] The union of the sets of equivalent nodes in the stack is equal to the set of visited nodes minus the set of explored nodes.
+    \item[$\bullet$] A node in the stack can reach all nodes before it in the stack (\textit{i.e.} pushed after)
+    \item[$\bullet$] $\mathcal{S}$ represents a set of strongly connected components (not maximal).
+    \item[$\bullet$] For all explored nodes, the sub-graph induced by their successors is totally explored
+    \item[$\bullet$] \texttt{sccs} is a set of maximal SCCs
+\end{itemize}
+\noindent
+These properties are natural and most of them are easy to prove.
+
+\BlankLine
+\BlankLine
+
+It is also useful to induce a notion of monotonicity on the environments during the execution of the algorithm.
+This is define as follows through the definition of an ordering relation on environments:
+
+\isa{
+\bblue{definition} sub\_env \where\\
+"\blue{sub\_env} \green{e e'} $\equiv$\\
+$~~~~$visited \green{e} $\subseteq$ visited \green{e'}\\
+$~~$$\wedge$ explored \green{e} $\subseteq$ explored \green{e'}\\
+$~~$$\wedge$ ($\forall$ \green{v}. $\mathcal{S}$ \green{e v} $\subseteq$ $\mathcal{S}$ \green{e' v})\\
+$~~$$\wedge$ ($\bigcup$\{$\mathcal{S}$ \green{e v} $|$ \green{v}.
+\green{v} $\in$ set (stack \green{e})\}) $\subseteq$ ($\bigcup$\{$\mathcal{S}$ \green{e' v} $|$ \green{v}.
+\green{v} $\in$ set (stack \green{e'})\})"
+}
+
+\BlankLine
+\BlankLine
+
+\subsubsection{\texttt{dfs} pre- and post-conditions}
+The pre-conditions of \texttt{dfs} are rather simple.
+The environment must be well-formed and the node must not be visited.
+There is also a condition on the reachability of the nodes in the stack and the node on which the function is called, but once again this condition is rather natural to consider since \texttt{dfs} is performing a DFS graph traversal:
+
+\isa{
+\bblue{definition} pre\_dfs \where\\
+$~~$"\blue{pre\_dfs} \green{v e} $\equiv$\\
+$~~~~$wf\_env \green{e}\\
+$~~$$\wedge$ \green{v} $\notin$ visited \green{e}\\
+$~~$$\wedge$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n v})"
+}
+
+\BlankLine
+\BlankLine
+
+The post-conditions are a little more complex since it has to consider the new environment with the new visited / explored nodes, the new state of the stack and the updates in $\mathcal{S}$:
+
+\isa{
+\bblue{definition} post\_dfs \where\\ 
+"\blue{post\_dfs} \green{v prev\_e e} $\equiv$\\
+$~~~~$wf\_env \green{e}\\
+$~~\wedge$ ($\forall$ \green{x}. reachable \green{v} \green{x} $\longrightarrow$ \green{x} $\in$ visited e)\\
+$~~\wedge$ sub\_env \green{prev\_e} \green{e}\\
+$~~\wedge$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n} \green{v})\\
+$~~\wedge$ ($\exists$ \green{ns}. stack \green{prev\_e} = ns \@ (stack \green{e}))\\
+$~~\wedge$ ($\forall$ \green{m n}.
+$\stackprec{m}{n}{prec\_e}$ $\longrightarrow$ \\ $~~~~~~$ ($\forall$ \green{u} $\in$ $\mathcal{S}$ \green{prev\_e m}.
+reachable \green{u} \green{v} $\wedge$ reachable \green{v} \green{n} $\longrightarrow$ $\mathcal{S}$ \green{e m} = $\mathcal{S}$ \green{e n}))\\
+$~~\wedge$ ((\green{v} $\in$ explored \green{e} $\wedge$ stack \green{e} = stack \green{prev\_e}) $\lor$ \green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack e)))"
+}
 

### 2022-05-27 LaTeX/report.tex
comments on report

@@
 
+% sm: better: \def\NN{\ensuremath{\mathcal{N}}} etc.
 \def\NN{$\mathcal{N}~$}
@@
         \centering
-        \includegraphics[height = .1\textwidth]{/Users/vincent/Documents/MinesNancy/logoartem.png}
+        \includegraphics[height = .1\textwidth]{img/logoartem.png}
         \hspace{1cm}
@@
 
+\ednote{sm}{The introduction should also give an (informal) overview of what this report is about.
+Alternatively, split it into a section on the context of the work and one that introduces the subject.
+Avoid jumping directly into formal definitions without an introduction for the mere mortal.}
+
 \pagebreak
+
+\ednote{sm}{Again, the question whether or not to break pages at sections is decided by the \LaTeX{} style. I would not impose a page break here.}
 
@@
 \subsubsection{Directed graphs}\label{sec:directedgraphs}
+
+\ednote{sm}{The first definition should introduce directed graphs. In particular, it is not formally clear what ``\VV'', ``\GG'', and ``reachability'' refer to in the following definition.
+
+Side note: Your macros \VV, \GG, \EE introduce trailing blank space, which they should not. (Also, they do not work within math environments, which is mildly annoying.)
+You may want to use the xspace style for intelligently appending a space after a macro.}
 
@@
 \begin{definition}[SCC]
-    Let \GG = (\VV, \EE) be an directed graph.
+    Let \GG = (\VV, \EE) be a directed graph.
 $\mathcal{C} \subseteq \mathcal{V}$ is a strongly connected component (SCC) of \GG if:
@@
 
+\ednote{sm}{This section should contain some text in order to explain the figures, beyond the captions. Also, Fig.~1 is more about connected graphs than SCCs.}
+
 \pagebreak
@@
 \end{remark}
+
+\ednote{sm}{I don't quite understand this remark: $\mathcal{S}$ approximates mutual reachability, but doesn't capture it initially.
+Shouldn't the equivalence relation rather be $x \in \mathcal{S}(y) \land y \in \mathcal{S}(x)$?}
 
@@
 \end{definition}
+
+\ednote{sm}{This definition is really about graphs and would fit better in section 2.
+Also, what you really want is $\textsc{Post}(v) = \{w : (v,w) \in \mathcal{E}\}$.
+The definition only provides an inclusion, for example $\textsc{Post}(v) = \emptyset$ satisfies it.}
 
@@
 \ednote{sm}{%
+  Give some more context than just the reference for the algorithm.
   The algorithm should also explain how $\mathcal{S}$ is initialized.
+  And the result is really such that each element of \texttt{SCCs} is a (maximal) strongly connected component, not a set of components, and it is the set of SCCs in the subgraph of \GG reachable from $v_0$.
 }
@@
 
+\ednote{sm}{I do not find this proof convincing.
+The \texttt{while} loop is only executed at a certain point during the algorithm, and its condition is for two sets $\mathcal{S}$ to be different, not equal, as in the condition in the proof?}
+
 \begin{remark}\label{rem:rootSCC}
@@
     The disjointness of all on-stack partial SCCs is given by lemma \ref{lemma:disjointness}.
-Nodes from $\textsc{Visited} \setminus \textsc{Explored}$ are in \texttt{R} because they are being processed.
+Nodes from $\textsc{Visited} \setminus \textsc{Explored}$ are in \texttt{R}\edmargin{sm}{not true: they have a (unique) representative in \texttt{R} but need not themselves be in \texttt{R}} because they are being processed.
 So, $\textsc{Live} \subseteq \texttt{R}$.\\
@@
 \end{proof}
+
+\ednote{sm}{Should be strengthened to ``the set of maximal SCCs reachable from $v_0$.}
+
+\ednote{sm}{A figure illustrating the execution of the algorithm would be nice, also in the talk presenting the project.}
 
@@
 
+\ednote{sm}{I don't really understand the prose above.
+A recursive function definition has to be based on some underlying inductive definition, which is simply not available for graphs.
+You say this, but the beginning of the paragraph seems to say otherwise.
+I'd just say that Isabelle provides a construction for inductive predicate definitions, which is appropriate here, and then explain that the two clauses represent the reflexive case and the extension of reachability by prepending an edge.}
+
 \isa{
@@
 }
+
+\ednote{sm}{I'm afraid that the blue color used for \texttt{reachable} here can be confusing, although Isabelle indeed displays the definition in this way. I'd make it black throughout.}
 
@@
 Mathematical writing: $\forall x, \forall y, \forall z, (x \Rightarrow^* y \wedge y \Rightarrow z) \Longrightarrow x \Rightarrow^* z$
+
+\ednote{sm}{Note that this is the ``mirror'' of clause \texttt{reachable\_succ} (appending an edge).}
 
@@
 
+\ednote{sm}{``Telling Isabelle'' sounds like you are introducing an axiom. Rather, these facts are deduced using Isabelle from the above definitions.}
+
 \subsubsection{Equivalence relation and graph partition}
@@
 In Isabelle, a first recursive definition was written as follows:
+
+\ednote{sm}{Note that this is different from the \textsc{Unite} function introduced earlier, and perhaps rename one of the two.
+Also, I'd introduce this function together with the \texttt{dfs} and \texttt{dfss} functions where it is used.
+Finally, the heading of the subsection doesn't seem to fit the text.}
 
@@
     $~~~~~~~~~~$\green{cc} = $\bigcup$ \{$\mathcal{S}$ \green{e x} $|$ \green{x}. \green{x} $\in$ set \green{pfx} $\cup$ \{hd \green{sfx}\}\}\\
-    $~~~~~~$\bblue{in}  \green{e}\env{$\mathcal{S}$ := $\lambda$\green{x}. \bblue{if} \green{x} $\in$ \green{xx} \bblue{then} \green{cc} \bblue{else} $\mathcal{S}$ \green{e x}, stack := \green{sfx}}"
+    $~~~~~~$\bblue{in}  \green{e}\env{$\mathcal{S}$ := $\lambda$\green{x}. \bblue{if} \green{x} $\in$ \green{cc} \bblue{then} \green{cc} \bblue{else} $\mathcal{S}$ \green{e x}, stack := \green{sfx}}"
 }
@@
 Then, \texttt{cc} -- which stands for \textit{connected component} -- contains all the nodes which are equivalent to \texttt{w} in the sub-graph currently explored.
-The function \texttt{takeWhile} applied to a boolean function \texttt{P}\footnote{\texttt{P :: \generic{a} $\Rightarrow$ bool}} seen as a property and a list \texttt{xs} returns the elements of \texttt{xs} which satisfy \texttt{P} and stops to the first element not satisfying P.
+The function \texttt{takeWhile} applied to a boolean function \texttt{P}\footnote{\texttt{P :: \generic{a} $\Rightarrow$ bool}} seen as a property and a list \texttt{xs} returns the elements of \texttt{xs} which satisfy \texttt{P} and stops at the first element not satisfying \texttt{P}.
 The function \texttt{dropWhile} is the opposite of \texttt{takeWhile}.
@@
 
+\ednote{sm}{The two are equivalent within the context of the algorithm, not in general.
+And the equivalence has not been proved.
+In fact, the non-recursive (why ``imperative''?) definition is intended to be simpler for the proof because it avoids introducing separate pre- and post-conditions for the function and proving such a ``contract''.}
+
 \subsubsection{Ordering relation}\label{sec:orderingrelation}
-In the proof, a precedence relation\footnote{In fact, a total order is being defined on stacks.} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack.
+In the proof, a precedence relation\footnote{In fact, a total order is being defined on stacks.}\edmargin{sm}{partial?
+don't have $x \preceq y \lor y \preceq x$} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack.
 Let $x$ and $y$ be two nodes and $R$ be a stack.
@@
 
-All the different properties (\textit{i.e.} lemmas) which follow this definition in the Isabelle implementation are detailed in the natural mathematical writing in \nameref{appendix}.
+All the different properties (\textit{i.e.} lemmas) which follow this definition in the Isabelle implementation are detailed in the natural mathematical writing in the \nameref{appendix}.
 The right part of the notation represents the orders of priority for each operand since $\preceq$ is an infix operator.
@@
 The main function is called \texttt{dfs} and takes its name after the Depth First Search algorithm because the algorithm \ref{alg:seqsetbased} roughly consists in a deep traversal of a graph.
-The second function is called \texttt{dfss} and represents the \textit{while} loop of the algorithm \ref{alg:seqsetbased}.
+The second function is called \texttt{dfss} and represents the \textit{while} loop\edmargin{sm}{\textbf{foreach} loop?!} of the algorithm \ref{alg:seqsetbased}.
 The two functions are mutually recursive because they recursively call each other.
-In particular, \texttt{dfss} will call either itself or \texttt{dfs}, depending on the case.
+In particular, \texttt{dfss} will call either\edmargin{sm}{both itself and \texttt{dfs}} itself or \texttt{dfs}, depending on the case.
 
@@
 
+\ednote{sm}{\texttt{force} finishes the proof of pattern completeness, the proof of termination remains open, and it would actually show that these are well-defined functions.
+Also, you should say that \texttt{force} is more aggressive \emph{than \texttt{auto}}.}
+
 \pagebreak
@@
 \subsubsection{General scheme}
-As the algorithm is composed of two mutually recursive functions, the correctness of the algorithm is proved by induction on the environment structure (cf \ref{envdef}).
-Since both \texttt{dfs} and \texttt{dfss} are quite complex, the proof is split into several parts.
+As the algorithm is composed of two mutually recursive functions, the correctness of the algorithm is proved by induction on the environment structure (cf \ref{envdef}).\edmargin{sm}{mutual induction on the functions?
+There is no induction principle for environments.} Since both \texttt{dfs} and \texttt{dfss} are quite complex, the proof is split into several parts.
 The idea is to prove for each function that its execution given some pre-conditions on the input environment implies some post-conditions on the output environment.
 Then, it has to be made for the mutually recursive calls as well, so that given the same pre-conditions on one function, the pre-conditions on the other function are also satisfied.
-Finally, it has to be proved that if the pre-conditions are satisfied for one function, and if the pre-conditions implie the post-conditions on the other function, then the post-conditions are also satisfied for the first function.
+Finally, it has to be proved that if the pre-conditions are satisfied for one function, and if the pre-conditions imply the post-conditions on the other function, then the post-conditions are also satisfied for the first function.
 
@@
 \begin{itemize}
-    \item[$\bullet$] First, the stack is a set of distinct elements.
+    \item[$\bullet$] First, the stack is a list of distinct elements.
     \item[$\bullet$] All elements of the stack are visited.
@@
     \item[$\bullet$] The union of the sets of equivalent nodes in the stack is equal to the set of visited nodes minus the set of explored nodes.
-    \item[$\bullet$] A node in the stack can reach all nodes before it in the stack (\textit{i.e.} pushed after)
+    \item[$\bullet$] A node in the stack can reach all nodes before it in the stack (\textit{i.e.} pushed later).
     \item[$\bullet$] $\mathcal{S}$ represents a set of strongly connected components (not maximal).
-    \item[$\bullet$] For all explored nodes, the sub-graph induced by their successors is totally explored
+    \item[$\bullet$] For all explored nodes, the sub-graph induced by their successors is totally explored.
     \item[$\bullet$] \texttt{sccs} is a set of maximal SCCs
@@
 
+\ednote{sm}{Actually, there is a bit of redundancy here.
+For example, the second and fourth conjunct follow from the fifth and eighth.
+This is not a bad thing per se since it may help automatic proof, but could be discussed.}
+
 \BlankLine
@@
 It is also useful to induce a notion of monotonicity on the environments during the execution of the algorithm.
-This is define as follows through the definition of an ordering relation on environments:
+This is defined as follows through the definition of an ordering relation on environments:
 
@@
 }
+
+\ednote{sm}{Add a bit of explanation, as for the basic invariant above, in particular for the last conjunct.}
 
@@
 $~~\wedge$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n} \green{v})\\
-$~~\wedge$ ($\exists$ \green{ns}. stack \green{prev\_e} = ns \@ (stack \green{e}))\\
+$~~\wedge$ ($\exists$ \green{ns}. stack \green{prev\_e} = ns @ (stack \green{e}))\\
 $~~\wedge$ ($\forall$ \green{m n}.
@@
 }
+
+\ednote{sm}{Add some explanations, and continue \dots}
+
+\ednote{sm}{Don't forget to write a conclusion, explaining what has been done, what is missing / could be improved, and what your experience has been.}
 
@@
 
+\ednote{sm}{Add a few more references, such as about Isabelle, and possibly other SCC algorithms (Dijkstra, Tarjan). For each reference, indicate where it was published, not just the year.}
+
 \end{document}

### 2022-05-27 LaTeX/report.tex
Consequent updates on the report, advances in the proof

@@
 %__________________________________________
+\newcommand\blankpage{%
+    \null
+    \thispagestyle{empty}%
+    \addtocounter{page}{-1}%
+    \newpage}
+
+%__________________________________________
 
 % sm: better: \def\NN{\ensuremath{\mathcal{N}}} etc.
-\def\NN{$\mathcal{N}~$}
-\def\GG{$\mathcal{G}~$}
-\def\VV{$\mathcal{V}~$}
-\def\EE{$\mathcal{E}~$}
+\def\NN{\ensuremath{\mathcal{N}}}
+\def\GG{\ensuremath{\mathcal{G}}}
+\def\VV{\ensuremath{\mathcal{V}}}
+\def\EE{\ensuremath{\mathcal{E}}}
 
@@
     \end{center}
-    \vspace{1cm}
+    \vspace{3 cm}
     \begin{center}
-        \large{Département Informatique --- Parcours Recherche \\ Tuteur : Stephan Merz}
+        Vincent Trélat
+        \BlankLine
+        \BlankLine
+        Supervized by Stephan Merz
     \end{center}
-    \vspace{1 cm}
-    \begin{center}
-        Vincent TRÉLAT
-    \end{center}
-    \vspace{2 cm}
+    \vspace{3 cm}
     \begin{center}
@@
     \end{center}
-    \vspace{2 cm}
+    \vspace{3 cm}
     \begin{center}
@@
 \restoregeometry
-\pagebreak
+
+\blankpage
+
+\hfill
+\begin{minipage}{.45\textwidth}
+    \textit{
+    My warmest and most sincere thanks to Stephan Merz, who has been a great help in the development the whole project and who reviewed this paper with great care.
+    }
+\end{minipage}
 
@@
 \tableofcontents
+
 \pagebreak
 
-\section{Introduction}
+\section{Preamble}
 \subsection{Academic context}
@@
 Formal methods are a field of computer science related to mathematical logic and reasoning.
-The whole purpose of the discipline is to ensure by a logical proof that a given algorithm is not only correct on its domain of definition, but also to find -- or define -- that domain.
+The whole purpose of the discipline is to give precise, mathematical definitions to computer science concepts.
 Formal methods find applications in a variety of fields, both concrete, such as the railway industry or self-driving cars, and abstract, such as computational architecture.
-\BlankLine
-Although a formal proof lies first on paper, the real formalisation starts when proofs are mechanised in a proof assistant.
-
-\ednote{sm}{%
-  First, a stylistic remark: do not end lines in a \LaTeX\ document using \texttt{\textbackslash\textbackslash} (except in \texttt{tabular} environments or similar).
-Use a blank line to start a new paragraph: the style will define the appearance of a paragraph break.
-
-  More importantly, formal methods should not be equated with theorem proving.
-They are really about giving precise, mathematical definitions to computer science concepts.
 Although the purpose of giving such definitions is to enable formal verification, many techniques besides theorem proving, such as model-based testing, run-time monitoring, model checking etc.\ are used.
-}
 
@@
 \begin{quote}
-    Isabelle is a generic proof assistant. It allows mathematical formulas to be expressed in a formal language and provides tools for proving those formulas in a logical calculus.
+    ``Isabelle is a generic proof assistant. It allows mathematical formulas to be expressed in a formal language and provides tools for proving those formulas in a logical calculus.''
 \end{quote}
@@
 \end{flushright}
-Isabelle is a really powerful proof assistant coming with a higher order logic (HOL) proving environment.
+Isabelle \cite{nipkow_isabellehol_2002} is a really powerful proof assistant coming with a higher order logic (HOL) proving environment.
 Isabelle proofs are written in the Isar (``intelligible semi-automated reasoning'') language that is designed to make proofs readable and comprehensible for a mathematically inclined reader, with minimal overhead introduced by the formalism.
@@
 
-\ednote{sm}{The introduction should also give an (informal) overview of what this report is about.
-Alternatively, split it into a section on the context of the work and one that introduces the subject.
-Avoid jumping directly into formal definitions without an introduction for the mere mortal.}
+\section{Introduction}
 
-\pagebreak
+The objective of this project is to mechanize a proof of correctness of a set-based algorithm inspired by Tarjan's algorithm \cite{TarjanDFS} for computing the strongly connected components of a graph.
+A similar work has been done on Tarjan's algorithm \cite{TarjanMerz}.
+The algorithm was first published in Vincent Bloemen's thesis \cite{bloemen_strong_2019} who furthermore gives and explains a few invariants, and was later reused in \cite{bloemen_multi-core_2016} with the aim of working on a parallel version of the algorithm.
 
-\ednote{sm}{Again, the question whether or not to break pages at sections is decided by the \LaTeX{} style. I would not impose a page break here.}
+In this report, a few arguments are given for the understanding of the formal proof.
+Some important invariants are explained and the main lemmas are briefly detailed.
+Some of the proofs will be explicitly given in slightly less rigorous mathematical terms for the sake of better understandability.
 
 \section{Formalisation}
-\subsection{Strongly connected components}
-\subsubsection{Directed graphs}\label{sec:directedgraphs}
-
-\ednote{sm}{The first definition should introduce directed graphs. In particular, it is not formally clear what ``\VV'', ``\GG'', and ``reachability'' refer to in the following definition.
-
-Side note: Your macros \VV, \GG, \EE introduce trailing blank space, which they should not. (Also, they do not work within math environments, which is mildly annoying.)
-You may want to use the xspace style for intelligently appending a space after a macro.}
+\subsection{Graphs and reachability}
+\begin{definition}[Directed graph]
+    A directed graph \GG\xspace is the data of a set of nodes \VV\xspace and a set of oriented edges \EE.
+\end{definition}
 
@@
 
+\begin{definition}[Successors set for a node]
+    Let $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ and $v \in \mathcal{V}$. The set of successors of $v \in \VV$ is \textsc{Post}($v$) such that:
+    \begin{equation*}
+        \textsc{Post}(v) = \{w \in \VV~|~(v,w) \in \mathcal{E}\}
+    \end{equation*}
+\end{definition}
+
+\subsection{Strongly connected components}
+\subsubsection{Directed graphs}\label{sec:directedgraphs}
+
 \begin{definition}[SCC]
@@
 \BlankLine
-$\mathcal{C}$ is maximal, or $\mathcal{C}$ is a maximal SCC of \GG if there is no other SCC containing $\mathcal{C}$, \textit{i.e.} if:
+$\mathcal{C}$ is maximal, or $\mathcal{C}$ is a maximal SCC of \GG \xspace if there is no other SCC containing $\mathcal{C}$, \textit{i.e.} if:
 \begin{equation*}
@@
 \begin{definition}(Strong connectedness)
-Let \GG = (\VV, \EE) be a directed graph. \GG is strongly connected if \VV is a SCC.
+Let \GG = (\VV, \EE) be a directed graph. \GG \xspace is strongly connected if \VV \xspace is a SCC.
 \end{definition}
@@
 \subsubsection{Examples}
+Let us give some visual examples of a strongly connected component in a directed graph.
+
 \begin{figure}[!h]
@@
         \ctikzfig{example2}
-        \subcaption{Strongly connected graph}
+        \subcaption{Strongly connected component}\label{fig:example1_a}
     \end{subfigure}
@@
         \ctikzfig{example3}
-        \subcaption{Not strongly connected graph}
+        \subcaption{Not strongly connected component}\label{fig:example1_b}
     \end{subfigure}
-    \caption{Basic example of what is a small SCC}
+    \caption{Basic example of what is a small SCC}\label{fig:example1}
 \end{figure}
+
+In \textsc{Figure} \ref{fig:example1}, two small directed graphs are shown.
+The first one (\textsc{Figure} \ref{fig:example1_a}) is strongly connected, but the second one (\textsc{Figure} \ref{fig:example1_b}) is not because the node $x$ is not reachable from $w$ for instance.
+
 \begin{figure}[!h]
     \ctikzfig{exampleSCC}
-    \caption{Example of a graph where each colored set of node is a -- maximal -- SCC\label{fig:exampleSCC}}
+    \caption{Example of a graph where each colored set of nodes is a -- maximal -- SCC\label{fig:exampleSCC}}
 \end{figure}
+
+Let us give another example on a larger graph.
+\textsc{Figure} \ref{fig:exampleSCC} shows a directed graph on which each colored set of nodes is a -- maximal -- SCC.
+Therefore, one can informally understand that SCCs roughly describe the cycles in a graph.
 
@@
     \ctikzfig{exampleSCCreduced}
-    \caption{Reduced visualization of the graph represented in figure \ref{fig:exampleSCC}}
+    \caption{Reduced visualization of the graph represented in \textsc{Figure} \ref{fig:exampleSCC}}\label{fig:exampleSCCreduced}
 \end{figure}
 
-\ednote{sm}{This section should contain some text in order to explain the figures, beyond the captions. Also, Fig.~1 is more about connected graphs than SCCs.}
-
-\pagebreak
+One may want to have a more general view of and consider only the distinct components of a graph.
+Thus, the graph show on \textsc{Figure} \ref{fig:exampleSCC} can be reduced to a graph in which the previously colored nodes are replaced by a single node containing all the nodes of the same SCCs, \textit{i.e.} all the equivalent nodes, as shown in \textsc{Figure} \ref{fig:exampleSCCreduced}.
 
@@
 \begin{remark}
-    In the following, the same notation $\mathcal{S}$ will be used to denote both the function defined above and the induced equivalence relation\footnote{For the relation $(x, y) \mapsto x \Rightarrow^* y ~\land~ y \Rightarrow^* x $} since $\mathcal{S}$ associates to each node its class of equivalence.
+    In the following, the same notation $\mathcal{S}$ will be used to denote both the function defined above and the induced equivalence relation\footnote{For the relation $(x, y) \mapsto x \in \mathcal{S}(y) ~\land~ y \in \mathcal{S}(x) $} since $\mathcal{S}$ associates to each node its class of equivalence.
 \end{remark}
-
-\ednote{sm}{I don't quite understand this remark: $\mathcal{S}$ approximates mutual reachability, but doesn't capture it initially.
-Shouldn't the equivalence relation rather be $x \in \mathcal{S}(y) \land y \in \mathcal{S}(x)$?}
 
@@
 
-\begin{definition}[Successors set for a node]
-    Let $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ and $v \in \mathcal{V}$. The set of successors of $v$ in \GG is \textsc{Post}($v$) such that:
-    \begin{equation*}
-        \forall w\in \textsc{Post}(v), (v,w)\in \mathcal{E}
-    \end{equation*}
-\end{definition}
+\subsection{The algorithm}
+This section gives a pseudo-code of the set-based algorithm for which we will write a formal proof. See \cite{bloemen_strong_2019} for the original paper.
+\BlankLine
+The algorithm only takes as input a directed graph $\GG = (\VV, \EE)$ and a starting node $v_0$ as the root of its exploration.
+It returns a partition \texttt{SCCs} of \VV \xspace being the set of SCCs in the subgraph of \GG\xspace reachable from $v_0$, where each element of \texttt{SCCs} is a maximal strongly connected components of \GG.
+The equivalence relation $\mathcal{S}$ is initialized so that at the beginning, each node is its own SCC.
+That is, $\mathcal{S}(v) = \{v\}$ for all $v \in \VV$, or seen as a set of disjoint sets: $\mathcal{S} = \bigcup \{ \{v\} \mid v \in \VV \}$.\\
+\BlankLine
 
-\ednote{sm}{This definition is really about graphs and would fit better in section 2.
-Also, what you really want is $\textsc{Post}(v) = \{w : (v,w) \in \mathcal{E}\}$.
-The definition only provides an inclusion, for example $\textsc{Post}(v) = \emptyset$ satisfies it.}
-
-\subsection{The algorithm}
-See \cite{articleSCC} for the original paper.\\
+\noindent
 \begin{algorithm}[H]\label{alg:seqsetbased}
     \SetAlgoLined
-    \KwData{A graph \GG = (\VV, \EE), a starting node $v_0$\;}
-    \KwResult{A partition \texttt{SCCs} of \VV where each element of \texttt{SCCs} is a maximal set of strongly connected components of \GG\;}
+    \KwData{\GG = (\VV, \EE), $v_0$\;}
+    \KwResult{\texttt{SCCs}\;}
     Initialize an empty set \texttt{EXPLORED}\;
@@
     Initialize an empty stack \texttt{R}\;
+    Initialize $\mathcal{S} \colon v \in \VV \mapsto \{v\}$\;
     setBased($v_0$)\;
@@
 \end{algorithm}
-\ednote{sm}{%
-  Give some more context than just the reference for the algorithm.
-  The algorithm should also explain how $\mathcal{S}$ is initialized.
-  And the result is really such that each element of \texttt{SCCs} is a (maximal) strongly connected component, not a set of components, and it is the set of SCCs in the subgraph of \GG reachable from $v_0$.
-}
-
-\pagebreak
 
@@
 \begin{proof}
-    Let $x, y \in \texttt{R}$, $x \neq y$. We can assume w.l.o.g that $\stackprec{y}{x}{R}$.
-    If $y \notin \mathcal{S}(x)$, then $\mathcal{S}(x)$ is disjoint from $\mathcal{S}(y)$.
+    Let us consider the stacj $\texttt{R}$ at some step of the algorithm. Let $x = \text{hd}(R)$, \textit{i.e.} the first element of the stack.
     \BlankLine
-    Otherwise, let us assume that $y \in \mathcal{S}(x)$, \textit{i.e.} $\mathcal{S}(x) = \mathcal{S}(y)$.
-Then, the \textit{while}\footnote{See \nameref{alg:seqsetbased}} loop will unstack all nodes of \texttt{R} until the successor of $x$ in \texttt{R} is reached and add them with \textsc{Unite} to $\mathcal{S}(x)$, so that the only representative of $\mathcal{S}(x)$ on \texttt{R} is $x$.
+
+    Then let $y \in \texttt{POST}(x)$. If $y$ is already explored, then we can skip it.
+    \BlankLine
+    Otherwise, if $y$ is not in the visited set, then we explore it by calling the function \texttt{setBased} on $y$ and since $y$ is not visited, $\mathcal{S}(y) = \{y\}$.
+    \BlankLine
+    Finally, if $y$ is in the visited set, it means that either $y$ in on the stack or it has a representative $z \in \mathcal{S}(y)$ on the stack.
+However, one can understand that the \textit{while} loop below is executed until the equivalence relation is fully updated.
+In particular, the invariant is maintained, and there is only one representative of $\mathcal{S}(y)$ on the stack, being the earliest visited node in $\mathcal{S}(y)$ the traversal.
+
 \end{proof}
-
-\ednote{sm}{I do not find this proof convincing.
-The \texttt{while} loop is only executed at a certain point during the algorithm, and its condition is for two sets $\mathcal{S}$ to be different, not equal, as in the condition in the proof?}
 
@@
     The disjointness of all on-stack partial SCCs is given by lemma \ref{lemma:disjointness}.
-Nodes from $\textsc{Visited} \setminus \textsc{Explored}$ are in \texttt{R}\edmargin{sm}{not true: they have a (unique) representative in \texttt{R} but need not themselves be in \texttt{R}} because they are being processed.
-So, $\textsc{Live} \subseteq \texttt{R}$.\\
-    By L.6-7 of Algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.\\
-    L.9-10 ensure that no explored node is pushed in \texttt{R}.\\
+Nodes from $\textsc{Visited} \setminus \textsc{Explored}$ have a (unique) representative in \texttt{R} because they are being processed.
+So, $\textsc{Live} \subseteq \texttt{R}$.
+    \BlankLine
+    By L.6-7 of Algorithm \ref{alg:seqsetbased}, $\textsc{Visited} \subseteq \texttt{R}$.
+    L.9-10 ensure that no explored node is pushed in \texttt{R}.
     L.24-25 keep the invariant by unstacking explored nodes from \texttt{R}, so $\texttt{R} \cap \textsc{Explored} = \varnothing$.
@@
 \begin{theorem}
-    The sequential algorithm \ref{alg:seqsetbased} is correct, \textit{i.e.} it returns a set of maximal SCCs.
+    The sequential algorithm \ref{alg:seqsetbased} is correct, \textit{i.e.} it returns the set of maximal SCCs reachable from $v_0$.
 \end{theorem}
@@
 
-\ednote{sm}{Should be strengthened to ``the set of maximal SCCs reachable from $v_0$.}
-
-\ednote{sm}{A figure illustrating the execution of the algorithm would be nice, also in the talk presenting the project.}
-
-\pagebreak
+An example of the execution of the algorithm is given in \nameref{appendix}, in the section \ref{appendix:exec_ex}.
 
@@
 \subsubsection{Ordering relation}\label{sec:orderingrelation}
-In the proof, a precedence relation\footnote{In fact, a total order is being defined on stacks.}\edmargin{sm}{partial?
-don't have $x \preceq y \lor y \preceq x$} noted {\color{isa_dark_blue}$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack.
+In the proof, a precedence relation\footnote{In fact, a partial order is being defined on stacks.}noted \blue{$\bullet~\preceq~\bullet~\text{in}~\bullet$} will be needed on the stack.
 Let $x$ and $y$ be two nodes and $R$ be a stack.
@@
 The main function is called \texttt{dfs} and takes its name after the Depth First Search algorithm because the algorithm \ref{alg:seqsetbased} roughly consists in a deep traversal of a graph.
-The second function is called \texttt{dfss} and represents the \textit{while} loop\edmargin{sm}{\textbf{foreach} loop?!} of the algorithm \ref{alg:seqsetbased}.
+The second function is called \texttt{dfss} and represents the \textit{foreach} loop of the algorithm \ref{alg:seqsetbased}.
 The two functions are mutually recursive because they recursively call each other.
-In particular, \texttt{dfss} will call either\edmargin{sm}{both itself and \texttt{dfs}} itself or \texttt{dfs}, depending on the case.
+In particular, \texttt{dfss} will call both \texttt{dfs} and itself, depending on the case.
 
@@
 The two last keywords require explanations as well : \texttt{pat\_completeness} stands for \textit{pattern completeness} and ensures that there is no missing patterns.
-The keyword \texttt{force} is used\footnote{\texttt{force} is more aggressive in instantiation and seems to find the right instance.} to help Isabelle know -- by proving it -- that both \texttt{dfs} and \texttt{dfss} are actually functions and that those functions are well defined with respect to the usual logical and mathematical meaning.
+The keyword \texttt{force} is used\footnote{\texttt{force} is more aggressive in instantiation than \texttt{auto} and seems to find the right instance.} to help Isabelle know -- by proving it -- that both \texttt{dfs} and \texttt{dfss} are actually functions and that those functions are well defined with respect to the usual logical and mathematical meaning.
 
@@
 Also, you should say that \texttt{force} is more aggressive \emph{than \texttt{auto}}.}
-
-\pagebreak
 
@@
 These properties are natural and most of them are easy to prove.
-
-\ednote{sm}{Actually, there is a bit of redundancy here.
+Actually, there is a bit of redundancy here.
 For example, the second and fourth conjunct follow from the fifth and eighth.
-This is not a bad thing per se since it may help automatic proof, but could be discussed.}
+This is not a bad thing per se since it may help automatic proof, but could be discussed.
 
@@
 
-\ednote{sm}{Add a bit of explanation, as for the basic invariant above, in particular for the last conjunct.}
+In particular, the last conjunct expresses the fact that the equivalence relation on the stack is monotonic.
 
@@
 \ednote{sm}{Don't forget to write a conclusion, explaining what has been done, what is missing / could be improved, and what your experience has been.}
-
-\pagebreak
 
@@
 
-% continue with precedes_append_right_iff
+\subsection{An example of execution of algorithm \ref{alg:seqsetbased}.}\label{appendix:exec_ex}
+\begin{figure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg1}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg2}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg3}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg4}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg5}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg6}
+    \end{subfigure}
+\end{figure}
+\begin{figure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg7}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg8}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg9}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg10}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg11}
+    \end{subfigure}
+    \begin{subfigure}[t]{0.5\textwidth}
+        \ctikzfig{slides_seqalg12}
+    \end{subfigure}
+\end{figure}
 
@@
 
-\begin{thebibliography}{99}
-
-\bibitem{tarjanMerz} R. Chen, C. Cohen, J.-J. Lévy, S. Merz, L. Théry, \emph{Formal Proofs of Tarjan’s Strongly Connected Components Algorithm in Why3, Coq and Isabelle}, 2019
-
-\bibitem{thesisBloemen} V. Bloemen, A. Laarman, J. van de Pol, \emph{Multi-Core On-The-Fly SCC Decomposition}, 2016
-
-\bibitem{articleSCC} V. Bloemen, \emph{Strong Connectivity and Shortest Paths for Checking Models}, 2019
-
-\end{thebibliography}
-
-\ednote{sm}{Add a few more references, such as about Isabelle, and possibly other SCC algorithms (Dijkstra, Tarjan). For each reference, indicate where it was published, not just the year.}
+% bibliography
+\bibliographystyle{plain}
+\bibliography{references}
 

### 2022-05-28 LaTeX/report.tex
Updates on the report

@@
     $~~~$sccs :: "\generic{v} set set"\\
-    $~~~$stack :: "\generic{v} list"\\
+    $~~~$stack :: "\generic{v} list"
 }
+
+\BlankLine
+\BlankLine
+\BlankLine
 
@@
     {\locale} graph =\\
-    $~~~$\fixes{} {\color{isa_dark_blue} vertices} :: "\generic{v} set" \and{} {\color{isa_dark_blue} successors} :: "\generic{v} $\Rightarrow$ \generic{v} set"\\
-    $~~~$\assumes{} vfin: "finite {\color{isa_dark_blue}vertices}"\\
-    $~~~$\and{} sclosed: "$\forall {\text{{\color{isa_dark_green}x}}} \in$ {\color{isa_dark_blue}vertices}.
-{\color{isa_dark_blue}successors} {\color{isa_dark_green}x} $\subseteq$ {\color{isa_dark_blue}vertices}"\\
+    $~~~$\fixes{} vertices :: "\generic{v} set" \and{} successors :: "\generic{v} $\Rightarrow$ \generic{v} set"\\
+    $~~~$\assumes{} vfin: "finite vertices"\\
+    $~~~$\and{} sclosed: "$\forall$ \green{x} $\in$ vertices. successors \green{x} $\subseteq$ vertices"
 }
+
+\BlankLine
+\BlankLine
+\BlankLine
 
@@
     \abbreviation{} edge \where\\
-    $~~~$"{\color{isa_dark_blue}edge} {\color{isa_dark_green}x y} $\equiv$ {\color{isa_dark_green}y} $\in$ {\color{isa_dark_blue}successors} {\color{isa_dark_green}x}"
+    $~~~$"edge \green{x y} $\equiv$ \green{y} $\in$ successors \green{x}"
 }
-\BlankLine
+
 \BlankLine
@@
 If both definitions are valid, the inductive one is kept for the following reasons.
-Although a recursive definition allows one to do some rewriting in the middle of terms, a recursive definition expresses both the positive and negative information\footnote{In this case, the positive information designates the fact of being reachable and the negative information designates the fact of not being reachable.} whereas the inductive one only expresses the positive information directly.
+Although a recursive definition has to be based on some underlying inductive definition -- which is simply not available for graphs -- it expresses both the positive and negative information\footnote{In this case, the positive information designates the fact of being reachable and the negative information designates the fact of not being reachable.} whereas the inductive one only expresses the positive information directly.
 Therefore, with an inductive definition, the negative information has to be proved.
@@
 Thus, the choice of the inductive definition is not a choice of simplicity but of necessity.
-Lastly, Isabelle will generate a simple inductive rule for the proofs split into the reflexive case, which stands in the definition, and the transitive case, which has to be proved.
-
-\ednote{sm}{I don't really understand the prose above.
-A recursive function definition has to be based on some underlying inductive definition, which is simply not available for graphs.
-You say this, but the beginning of the paragraph seems to say otherwise.
-I'd just say that Isabelle provides a construction for inductive predicate definitions, which is appropriate here, and then explain that the two clauses represent the reflexive case and the extension of reachability by prepending an edge.}
+Isabelle provides a construction for inductive predicate definitions, which is appropriate here because the two clauses represent the reflexive case and the extension of reachability by prepending an edge.
+This will be particularly useful in the proofs.
 
@@
     \inductive{} reachable \where\\
-    $~~~$reachable\_refl[iff]: "{\color{isa_dark_blue}reachable} {\color{isa_dark_green}x x}"\\
-    $~{\text{|}}$ reachable\_succ[elim]: "$\ldb$edge {\color{isa_dark_green}x y};  {\color{isa_dark_blue}reachable} {\color{isa_dark_green}y z}$\rdb \Longrightarrow$ {\color{isa_dark_blue}reachable} {\color{isa_dark_green}x z}"\\
+    $~~~$reachable\_refl[iff]: "reachable \green{x x}"\\
+    $~\mid$ reachable\_succ[elim]: "$\ldb$edge \green{x y}; reachable \green{y z}$\rdb \Longrightarrow$ reachable \green{x z}"\\
 }
-
-\ednote{sm}{I'm afraid that the blue color used for \texttt{reachable} here can be confusing, although Isabelle indeed displays the definition in this way. I'd make it black throughout.}
 
@@
 
-\ednote{sm}{Note that this is the ``mirror'' of clause \texttt{reachable\_succ} (appending an edge).}
+\begin{remark}
+    Note that this is the ``mirror'' of clause \texttt{reachable\_succ} (appending an edge).
+\end{remark}
 
@@
 
-Once again, there are some lemmas to prove, such as telling Isabelle when an element can be added to a SCC, or that two vertices that are reachable from each other are in the same SCC, or that two SCCs having a common element are identical.\\
-
-\ednote{sm}{``Telling Isabelle'' sounds like you are introducing an axiom. Rather, these facts are deduced using Isabelle from the above definitions.}
-
-\subsubsection{Equivalence relation and graph partition}
-In the algorithm \ref{alg:seqsetbased}, the SCCs are progressively tracked in \texttt{sccs} and the equivalence relation $\mathcal{S}$ is updated with the \textsc{Unite} function.
-In Isabelle, a first recursive definition was written as follows:
-
-\ednote{sm}{Note that this is different from the \textsc{Unite} function introduced earlier, and perhaps rename one of the two.
-Also, I'd introduce this function together with the \texttt{dfs} and \texttt{dfss} functions where it is used.
-Finally, the heading of the subsection doesn't seem to fit the text.}
-
-\isa{
-    {\color{isa_blue}function} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
-    "\blue{unite} \green{v w e} =\\
-    $~~~$({\color{isa_blue}if} ($\mathcal{S}$ \green{e v} = $\mathcal{S}$ \green{e w}) \bblue{then} \green{e}\\
-    $~~~$\bblue{else let} \green{r} = hd(stack \green{e});\\
-    $~~~~~~~~~~~~$\green{r'}= hd(tl(stack \green{e}));\\
-    $~~~~~~~~~~~~$\green{joined} = $\mathcal{S}$ \green{e r} $\cup$ $\mathcal{S}$ \green{e r};\\
-    $~~~~~~~~~~~~$\green{e'}= \green{e}\env{\\
-    $~~~~~~~~~~~~~~~$stack := tl(stack \green{e}),\\
-    $~~~~~~~~~~~~~~~\mathcal{S}$ := ($\lambda$ n. \bblue{if} n $\in$ \green{joined} \bblue{then} \green{joined} \bblue{else} $\mathcal{S}$ \green{e n})\\
-    $~~~~~~~~~~~~$}\\
-    $~~~$\bblue{in} \blue{unite} \green{v w e'})"\\
-    $~~~$\bblue{by} pat\_completeness auto
-}
-
-\BlankLine
-
-However, this definition makes the proofs too difficult due to the recursion. An imperative version was therefore written:
-
-\isa{
-    \bblue{definition} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
-    $~~~$"\blue{unite} \green{v w e} $\equiv$\\
-    $~~~~~~$\bblue{let} \green{pfx} = takeWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
-    $~~~~~~~~~~$\green{sfx} = dropWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
-    $~~~~~~~~~~$\green{cc} = $\bigcup$ \{$\mathcal{S}$ \green{e x} $|$ \green{x}. \green{x} $\in$ set \green{pfx} $\cup$ \{hd \green{sfx}\}\}\\
-    $~~~~~~$\bblue{in}  \green{e}\env{$\mathcal{S}$ := $\lambda$\green{x}. \bblue{if} \green{x} $\in$ \green{cc} \bblue{then} \green{cc} \bblue{else} $\mathcal{S}$ \green{e x}, stack := \green{sfx}}"
-}
-
-\BlankLine
-\BlankLine
-
-The idea of this definition is to create a partition of \texttt{stack e = pfx @ sfx} such that \texttt{pfx} contains the nodes which are to be merged into \texttt{$\mathcal{S}$ e w} and \texttt{sfx} contains the root of \texttt{$\mathcal{S}$ e w} followed by the rest of the stack.
-Then, \texttt{cc} -- which stands for \textit{connected component} -- contains all the nodes which are equivalent to \texttt{w} in the sub-graph currently explored.
-The function \texttt{takeWhile} applied to a boolean function \texttt{P}\footnote{\texttt{P :: \generic{a} $\Rightarrow$ bool}} seen as a property and a list \texttt{xs} returns the elements of \texttt{xs} which satisfy \texttt{P} and stops at the first element not satisfying \texttt{P}.
-The function \texttt{dropWhile} is the opposite of \texttt{takeWhile}.
-Both the imperative and recursive versions of \texttt{unite} are equivalent.
-
-\ednote{sm}{The two are equivalent within the context of the algorithm, not in general.
-And the equivalence has not been proved.
-In fact, the non-recursive (why ``imperative''?) definition is intended to be simpler for the proof because it avoids introducing separate pre- and post-conditions for the function and proving such a ``contract''.}
+Once again, there are some lemmas to prove which are deduced using Isabelle from the above definitions, such as giving conditions on when an element can be added to a SCC, or that two vertices that are reachable from each other are in the same SCC, or that two SCCs having a common element are identical, etc.
 
@@
 
+In the algorithm \ref{alg:seqsetbased}, the SCCs are progressively tracked in \texttt{sccs} and the equivalence relation $\mathcal{S}$ is updated with the \textsc{Unite} function.
+Note that the function written in Isabelle is different from the \textsc{Unite} function introduced earlier.
+From now on, \textsc{Unite} designates the following function, which was first written as a recursive function as follows:
+
+\isa{
+    {\color{isa_blue}function} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
+    "\blue{unite} \green{v w e} =\\
+    $~~~$({\color{isa_blue}if} ($\mathcal{S}$ \green{e v} = $\mathcal{S}$ \green{e w}) \bblue{then} \green{e}\\
+    $~~~$\bblue{else let} \green{r} = hd(stack \green{e});\\
+    $~~~~~~~~~~~~$\green{r'}= hd(tl(stack \green{e}));\\
+    $~~~~~~~~~~~~$\green{joined} = $\mathcal{S}$ \green{e r} $\cup$ $\mathcal{S}$ \green{e r};\\
+    $~~~~~~~~~~~~$\green{e'}= \green{e}\env{\\
+    $~~~~~~~~~~~~~~~$stack := tl(stack \green{e}),\\
+    $~~~~~~~~~~~~~~~\mathcal{S}$ := ($\lambda$ n. \bblue{if} n $\in$ \green{joined} \bblue{then} \green{joined} \bblue{else} $\mathcal{S}$ \green{e n})\\
+    $~~~~~~~~~~~~$}\\
+    $~~~$\bblue{in} \blue{unite} \green{v w e'})"\\
+    $~~~$\bblue{by} pat\_completeness auto
+}
+
+\BlankLine
+
+However, this definition makes the proofs too difficult due to the recursion. A non-recursive version was therefore written:
+
+\isa{
+    \bblue{definition} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
+    $~~~$"\blue{unite} \green{v w e} $\equiv$\\
+    $~~~~~~$\bblue{let} \green{pfx} = takeWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
+    $~~~~~~~~~~$\green{sfx} = dropWhile ($\lambda$\green{x}. \green{w} $\notin$ $\mathcal{S}$ \green{e x}) (stack \green{e});\\
+    $~~~~~~~~~~$\green{cc} = $\bigcup$ \{$\mathcal{S}$ \green{e x} $|$ \green{x}. \green{x} $\in$ set \green{pfx} $\cup$ \{hd \green{sfx}\}\}\\
+    $~~~~~~$\bblue{in}  \green{e}\env{$\mathcal{S}$ := $\lambda$\green{x}. \bblue{if} \green{x} $\in$ \green{cc} \bblue{then} \green{cc} \bblue{else} $\mathcal{S}$ \green{e x}, stack := \green{sfx}}"
+}
+
+\BlankLine
+\BlankLine
+\BlankLine
+
+The idea of this definition is to create a partition of \texttt{stack e = pfx @ sfx} such that \texttt{pfx} contains the nodes which are to be merged into \texttt{$\mathcal{S}$ e w} and \texttt{sfx} contains the root of \texttt{$\mathcal{S}$ e w} followed by the rest of the stack.
+Then, \texttt{cc} -- which stands for \textit{connected component} -- contains all the nodes which are equivalent to \texttt{w} in the sub-graph currently explored.
+The function \texttt{takeWhile} applied to a boolean function \texttt{P}\footnote{\texttt{P :: \generic{a} $\Rightarrow$ bool}} seen as a property and a list \texttt{xs} returns the elements of \texttt{xs} which satisfy \texttt{P} and stops at the first element not satisfying \texttt{P}.
+The function \texttt{dropWhile} is the opposite of \texttt{takeWhile}.
+Both the recursive and non-recursive versions of \texttt{unite} are equivalent -- it should be proved though -- in the context of this report.
+In fact, the non-recursive definition is intended to be simpler for the proof because it avoids introducing separate pre- and post-conditions for the function and proving such a ``contract''.
+
+\BlankLine
+
 Now that the environment is set up, the actual algorithm -- seen as a function -- can be implemented.
@@
 In particular, \texttt{dfss} will call both \texttt{dfs} and itself, depending on the case.
+Their implementation is as follows:
+
+\BlankLine
 
@@
 The two last keywords require explanations as well : \texttt{pat\_completeness} stands for \textit{pattern completeness} and ensures that there is no missing patterns.
-The keyword \texttt{force} is used\footnote{\texttt{force} is more aggressive in instantiation than \texttt{auto} and seems to find the right instance.} to help Isabelle know -- by proving it -- that both \texttt{dfs} and \texttt{dfss} are actually functions and that those functions are well defined with respect to the usual logical and mathematical meaning.
+The method \texttt{force} finishes the proof of pattern completeness, the proof of termination remains open, and it would actually show that these are well-defined functions.
+{\texttt{force} is more aggressive in instantiation than \texttt{auto} and seems to find the right instance.
 
-\ednote{sm}{\texttt{force} finishes the proof of pattern completeness, the proof of termination remains open, and it would actually show that these are well-defined functions.
-Also, you should say that \texttt{force} is more aggressive \emph{than \texttt{auto}}.}
-
-\subsection{Formal proof}
 \subsubsection{General scheme}
-As the algorithm is composed of two mutually recursive functions, the correctness of the algorithm is proved by induction on the environment structure (cf \ref{envdef}).\edmargin{sm}{mutual induction on the functions?
-There is no induction principle for environments.} Since both \texttt{dfs} and \texttt{dfss} are quite complex, the proof is split into several parts.
+As the algorithm is composed of two mutually recursive functions, the correctness of the algorithm is proved by mutual induction on the functions with the help of the environment structure (cf \ref{envdef}).
+Since both \texttt{dfs} and \texttt{dfss} are quite complex, the proof is split into several parts.
 The idea is to prove for each function that its execution given some pre-conditions on the input environment implies some post-conditions on the output environment.
@@
 \bblue{definition} pre\_dfs \where\\
-$~~$"\blue{pre\_dfs} \green{v e} $\equiv$\\
+$~~$"pre\_dfs \green{v e} $\equiv$\\
 $~~~~$wf\_env \green{e}\\
@@
 \bblue{definition} post\_dfs \where\\ 
-"\blue{post\_dfs} \green{v prev\_e e} $\equiv$\\
+"post\_dfs \green{v prev\_e e} $\equiv$\\
 $~~~~$wf\_env \green{e}\\
@@
 $~~\wedge$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n} \green{v})\\
-$~~\wedge$ ($\exists$ \green{ns}. stack \green{prev\_e} = ns @ (stack \green{e}))\\
+$~~\wedge$ ($\exists$ \green{ns}. stack \green{prev\_e} = \green{ns} @ (stack \green{e}))\\
 $~~\wedge$ ($\forall$ \green{m n}.
-$\stackprec{m}{n}{prec\_e}$ $\longrightarrow$ \\ $~~~~~~$ ($\forall$ \green{u} $\in$ $\mathcal{S}$ \green{prev\_e m}.
+$\stackprec{\green{m}}{\green{n}}{\green{prec\_e}}$ $\longrightarrow$ \\ $~~~~~~$ ($\forall$ \green{u} $\in$ $\mathcal{S}$ \green{prev\_e m}.
 reachable \green{u} \green{v} $\wedge$ reachable \green{v} \green{n} $\longrightarrow$ $\mathcal{S}$ \green{e m} = $\mathcal{S}$ \green{e n}))\\
-$~~\wedge$ ((\green{v} $\in$ explored \green{e} $\wedge$ stack \green{e} = stack \green{prev\_e}) $\lor$ \green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack e)))"
+$~~\wedge$ ((\green{v} $\in$ explored \green{e} $\wedge$ stack \green{e} = stack \green{prev\_e}) $\lor$ \\
+$~~~~~~$ (\green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack \green{e}))) $\wedge$ \\
+$~~~~~~$ ($\exists$ \green{n} $\in$ set (stack \green{prev\_e}). $\mathcal{S}$ \green{e v} = $\mathcal{S}$ \green{e n}))"
 }
 
-\ednote{sm}{Add some explanations, and continue \dots}
+\BlankLine
+\BlankLine
+\BlankLine
+
+Let us give some explanations:
+\begin{itemize}
+    \item[$\bullet$] The environment must be well-formed.
+    \item[$\bullet$] The sub-graph induced by the nodes reachable from the node on which the function is called must be totally visited.
+    \item[$\bullet$] The previous environment must be a sub-environment of the new one (\textit{i.e.} there is a monotonic ordering on the environments).
+    \item[$\bullet$] The condition of reachability on the stack from the pre-condition must remain satisfied.
+    \item[$\bullet$] The new stack is a suffix of the previous one (\textit{i.e.} it expressively represents the structure of FIFO stack).
+    \item[$\bullet$] This one is a little more difficult to understand : it expresses the fact that if two nodes $m$ and $n$ such that $m \preceq n$ in the stack before the call on \texttt{dfs}, and if for all node $u$ on the stack, $u \Rightarrow^* v$ and $v \Rightarrow^* n$, then the execution of \texttt{dfs} has updated $\mathcal{S}$ such that $\mathcal{S}(m) = \mathcal{S}(n)$ in the environment returned by the function.
+This should be equated with the fourth conjunct : $v$ is reachable from all nodes on stack, and the fact that partials SCCs have a unique representative on stack.
+    \item[$\bullet$] The last one is easier to understand: either $v$ is explored and the stack considered before and after the execution of the function has not changed, or the head of the stack is the representative of $v$ and $v$ had a representative in the previous stack.
+\end{itemize}
+
+\subsubsection{\texttt{dfss} pre- and post-conditions}
+The pre-conditions on \texttt{dfss} are also rather simple, except the sixth conjunct which expresses the fact that before the execution of the function, if a node $n$ on the stack is reachable from $v$, then either $v \in \mathcal{S}(n)$, or $n$ is reachable from a successor of $v$.
+Once again, this condition has to be equated with the fact that all nodes on stack can reach $v$ (fifth conjunct in the following definition).
+
+\isa{
+\bblue{definition} pre\_dfss \where{}\\
+"pre\_dfss \green{v vs e} $\equiv$\\
+$~~~~$wf\_env \green{e}\\
+$~~\land$ \green{v} $\in$ visited \green{e}\\
+$~~\land$ \green{vs} $\subseteq$ successors \green{v}\\
+$~~\land$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n v})\\
+$~~\land$ (stack \green{e} $\neq$ [])\\
+$~~\land$ ($\forall$ \green{n}$\in$set(stack \green{e}).reachable \green{v n} $\longrightarrow$ \green{v}$\in$$\mathcal{S}$ \green{e n} $\lor$ ($\exists$ \green{m} $\in$ \green{vs}.reachable \green{m n}))\\
+$~~\land$ (\green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack \green{e})))"
+}
+
+\BlankLine
+\BlankLine
+\BlankLine
+
+The post-conditions on \texttt{dfss} are less complicated than the post-conditions of \texttt{dfs} and can easily be understood with the explanations of the other invariants.
+
+\isa{
+\bblue{definition} post\_dfss \where\\
+"post\_dfss \green{v vs prev\_e e} $\equiv$\\
+$~~~~$wf\_env \green{e}\\
+$~~\land$ ($\forall$ \green{w} $\in$ \green{vs}. $\forall$ \green{x}. reachable \green{w x} $\longrightarrow$ \green{x} $\in$ visited \green{e})\\
+$~~\land$ sub\_env \green{prev\_e e}\\
+$~~\land$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{n v})\\
+$~~\land$ (stack \green{e} $\neq$ [])\\
+$~~\land$ ($\forall$ \green{n} $\in$ set (stack \green{e}). reachable \green{v n} $\longrightarrow$ \green{v} $\in$ $\mathcal{S}$ \green{e n})\\
+$~~\land$ ($\exists$ \green{ns}. stack \green{prev\_e} = \green{ns} @ (stack \green{e}))\\
+$~~\land$ (\green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack \green{e})))"
+}
+
+\subsection{Formal proof}
+\subsubsection{\texttt{pre\_dfs} implies \texttt{pre\_dfss}}
+
+\subsubsection{\texttt{pre\_dfss} implies \texttt{pre\_dfs}}
+This lemma assumes the pre-conditions on \texttt{dfss} on a node $v$, a set (of successors) $vs$ and an environment $e$.
+It assumes that $w$ a successor of $v$ that is not visited, and it shows that the pre-conditions on \texttt{dfs} on $w$ the same environment $e$ are satisfied.
+The proof is quite as all conditions of \texttt{pre\_dfs} can be deduced from the conditions of \texttt{pre\_dfss}.
+
+\isa{
+\lm{} pre\_dfss\_pre\_dfs:\\
+$~~~$\fixes{} \blue{w}\\
+$~~~$\assumes{} "pre\_dfss \blue{v vs e}" and "\blue{w} $\notin$ visited \blue{e}" and "\blue{w} $\in$ \blue{vs}"\\
+$~~~$\shows{} "pre\_dfs \blue{w e}"
+}
+
+\subsubsection{\texttt{pre\_dfs} implies \texttt{post\_dfs}}
+
+\subsubsection{Partial correctness}
+
+\section{Conclusion}
 
@@
 \subsection{An example of execution of algorithm \ref{alg:seqsetbased}.}\label{appendix:exec_ex}
+This sequence of figures is meant to be read from left to right.
+Brown nodes are nodes which have not been visited yet.
+Pink nodes are nodes which are being visited, \textit{i.e.} which are on stack.
+Other colors are for reported SCCs.\\
+
 \begin{figure}

### 2022-05-29 LaTeX/report.tex
Partial_correctness and report updates

@@
 
-\theoremstyle{definition}
-\newtheorem{isabelle}{Isabelle}
+\newcommand{\isabelle}[1]{
+    \noindent\begin{minipage}{1.1\linewidth}
+        \isa{
+            #1
+        }
+    \end{minipage}
+}
 
@@
 
-\isa{
+\isabelle{
     {\color{isa_blue}record} \generic{v} env =\\
@@
 
-\isa{
+\isabelle{
     {\locale} graph =\\
@@
 
-\isa{
+\BlankLine
+
+\isabelle{
     \abbreviation{} edge \where\\
@@
 \noindent
-Regarding the reachability binary relation, a choice has to be made since there are several ways to define it.
-In particular, there are two possible keywords, {\color{isa_blue}\texttt{inductive}} and {\color{isa_blue}\texttt{fun}}, respectively for an inductive or recursive definition.
-If both definitions are valid, the inductive one is kept for the following reasons.
+
+Isabelle allows the definition of recursive functions.
+Such a definition must guarantee that any recursive call takes an argument that decreases (according to some measure function) with respect to the original argument.
+This is typically the case for functions that follow the definition of an algebraic type (integers, lists, trees, etc.).
+In addition, Isabelle allows inductive definitions of relations where two values are related if it is possible to justify this relation by applying a finite number of times the definition clauses of the relation.
+In our case, this applies to the definition of reachability where two nodes are connected if they are identical or if the first node has a successor from which the second node is already known to be reachable.
+
+\BlankLine
+\BlankLine
+
 Although a recursive definition has to be based on some underlying inductive definition -- which is simply not available for graphs -- it expresses both the positive and negative information\footnote{In this case, the positive information designates the fact of being reachable and the negative information designates the fact of not being reachable.} whereas the inductive one only expresses the positive information directly.
 Therefore, with an inductive definition, the negative information has to be proved.
-One would be right to argue that it would be more convenient to be able to tell without proving it that two nodes are not reachable from each other, but this does not interest us for the following.
+One would be right to argue that it would be more convenient to be able to tell without proving it that two nodes are not reachable from each other, but this does not interest us for the following and this is actually more difficult than proving the reachability.
 Another important point is that there is no datatype for a recursive definition, especially in this case with the transitive closure of the $\Rightarrow^*$ relation.
-Thus, the choice of the inductive definition is not a choice of simplicity but of necessity.
+Thus, the inductive definition is not a choice but is necessary.
 Isabelle provides a construction for inductive predicate definitions, which is appropriate here because the two clauses represent the reflexive case and the extension of reachability by prepending an edge.
@@
 
-\isa{
+\BlankLine
+\BlankLine
+
+\isabelle{
     \inductive{} reachable \where\\
@@
 
-\isa{
+\isabelle{
 \lm{} succ\_reachable:\\
@@
 
-\isa{
+\isabelle{
 \lm{} reachable\_trans:\\
@@
 
-\isa{
+\isabelle{
 {\color{isa_blue}definition} is\_subscc \where\\
@@
 
-\isa{
+\isabelle{
 {\color{isa_blue}definition} is\_scc \where\\
@@
 It is defined in Isabelle as follows:\\
-\isa{
+
+\isabelle{
     {\color{isa_blue}definition} precedes ("\_ $\preceq$ \_ in \_" [100,100,100] 39) \where\\
@@
 
-\isa{
+\BlankLine
+
+\isabelle{
     {\color{isa_blue}function} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
@@
 
-\isa{
+\BlankLine
+
+\isabelle{
     \bblue{definition} unite :: "\generic{v} $\Rightarrow$ \generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \where{}\\
@@
 
-\isa{
+\isabelle{
     {\color{isa_blue}{function}} dfs :: "\generic{v} $\Rightarrow$ \generic{v} env $\Rightarrow$ \generic{v} env" \and{}\\
@@
 
-\isa{
+\isabelle{
     \bblue{definition} wf\_env \where{}\\
@@
 
-\isa{
+\isabelle{
 \bblue{definition} sub\_env \where\\
@@
 
-\isa{
+\isabelle{
 \bblue{definition} post\_dfs \where\\ 
@@
 
-\isa{
+\isabelle{
 \bblue{definition} pre\_dfss \where{}\\
@@
 
-\isa{
+\isabelle{
 \bblue{definition} post\_dfss \where\\
@@
 \subsubsection{\texttt{pre\_dfs} implies \texttt{pre\_dfss}}
+This lemma assumes that the pre-conditions on \texttt{dfs} are satisfied and shows that it implies that the pre-conditions on \texttt{dfss} are satisfied on the environment on which \texttt{dfs} is called, \textit{i.e.} the environment for which $v$ was pushed on the stack and added to the set of visited nodes.
+
+\BlankLine
+
+\isabelle{
+\bblue{lemma} pre\_dfs\_pre\_dfss:\\
+$~~~$\assumes{} "pre\_dfs \blue{v e}"\\
+$~~~$\shows{} "pre\_dfss \blue{v} (successors \blue{v}) \\ $~~~~~~~~~~~~$\blue{e}\env{visited := visited \blue{e} $\cup$ \{\blue{v}\}, stack := \blue{v} \# stack \blue{e}}"
+}
+
+\BlankLine
+
+The proof is quite straightforward, except for the reachability of elements in the stack.
+Let $e'$ be the environment on which \texttt{dfss} will be called.
+Then, we want to show the following proposition:
+
+\begin{equation*}
+    \forall x, y,~x \preceq y \text{ in stack } e' \Longrightarrow y \Rightarrow^* x
+\end{equation*}
+
+\begin{proof}
+    Let $x$ and $y$ be two nodes on the stack of $e'$ such that $x \preceq y \text{ in stack } e'$.
+    \begin{itemize}
+        \item[if $x = v$:] then we have to show that $y \Rightarrow^* v$.
+If $y = v$, it is trivial, otherwise, $y$ is in the stack of $e$ and $v$ is reachable from $y$ from the pre-condition of \texttt{dfs} (fourth conjunct in the definition of \texttt{dfs}).
+        \item[if $x \neq v$:] then $x$ is in the stack of $e$ and $x \preceq y \text{ in stack } e$. From the pre-condition of \texttt{dfs}, $y \Rightarrow^* x$.
+    \end{itemize}
+
+\end{proof}
 
 \subsubsection{\texttt{pre\_dfss} implies \texttt{pre\_dfs}}
-This lemma assumes the pre-conditions on \texttt{dfss} on a node $v$, a set (of successors) $vs$ and an environment $e$.
-It assumes that $w$ a successor of $v$ that is not visited, and it shows that the pre-conditions on \texttt{dfs} on $w$ the same environment $e$ are satisfied.
+This lemma fixes a node $w$ and assumes the pre-conditions on \texttt{dfss} on a node $v$, a set (of successors) $vs$ and an environment $e$.
+It assumes that $w$ is a successor of $v$ that is not visited, and it shows that the pre-conditions on \texttt{dfs} on $w$ the same environment $e$ are satisfied.
 The proof is quite as all conditions of \texttt{pre\_dfs} can be deduced from the conditions of \texttt{pre\_dfss}.
 
-\isa{
+\isabelle{
 \lm{} pre\_dfss\_pre\_dfs:\\
@@
 
+\pagebreak
+
 \section{Appendix}\label{appendix}
@@
 Pink nodes are nodes which are being visited, \textit{i.e.} which are on stack.
-Other colors are for reported SCCs.\\
+Other colors are for reported SCCs.
 

### 2022-05-31 LaTeX/report.tex
Updates

@@
 \usepackage{listings}
+\usepackage{mathtools}
 \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
@@
 
-\subsubsection{\texttt{pre\_dfs} implies \texttt{post\_dfs}}
+\subsubsection{\texttt{pre\_dfs} implies \texttt{post\_dfs}}\label{lemma:pre_post_dfs}
 
 \subsubsection{Partial correctness}
+This lemma shows two things:
+\BlankLine
+\noindent
+\texttt{dfs\_dfss\_dom (Inl(v,e)) $\land$ pre\_dfs v e} $\Longrightarrow$ \texttt{post\_dfs v e (dfs v e)}
+\BlankLine
+\noindent
+\texttt{dfs\_dfss\_dom (Inr(v,e)) $\land$ pre\_dfss v vs e} $\Longrightarrow$ \texttt{post\_dfss v vs e (dfss v vs e)}
+
+\BlankLine
+\noindent
+where \texttt{dfs\_dfss\_dom (Inl(v,e))} -- respectively \texttt{Inr(v, e)} -- is the assumption that \texttt{v} and \texttt{e} are in the domain of definition of \texttt{dfs} (respectively \texttt{dfss}).
+\BlankLine
+The first proposition is a consequence of the lemma \nameref{lemma:pre_post_dfs}.
+\BlankLine
+For the second proposition, we need to write the proof according to the definition of \texttt{dfss}. That is, we need to show the following proposition:
+
+\BlankLine
+
+With these assumptions:
+\begin{itemize}
+    \item $\texttt{dfs\_dfss\_dom} (\texttt{Inr}(v,vs,e))$
+    \item $\texttt{pre\_dfss}~v~vs~e$
+    \item $vs \neq \varnothing$ and \\$\forall w \in vs,  w \notin \texttt{explored } e~\land~w \notin \texttt{visited } e~\land~\texttt{pre\_dfs}~w~e \Longrightarrow \texttt{post\_dfs}~w~e~(dfs~w~e)$
+    \item let $e' =
+    \begin{dcases}
+		e & w \in \texttt{explored}~e \\
+		\texttt{dfs}~v~e & w \in \texttt{visited}~e \\
+        \texttt{unite}~v~w~e & \text{otherwise}
+	\end{dcases}$\\
+    $vs \neq \varnothing$ and \\$\forall w \in vs, \texttt{pre\_dfss}~v~(vs - \{w\})~e' \Longrightarrow \texttt{post\_dfss}~v~(vs - \{w\})~e'~(\texttt{dfss}~v~(vs~-~\{w\})~e')$
+\end{itemize}
+\noindent
+We need to show the post-condition of \texttt{dfss}:
+\begin{equation*}
+    \texttt{post\_dfss}~v~vs~e~(\texttt{dfss}~v~vs~e)s
+\end{equation*}
+
+If the set of successors $vs$ is empty, then the post-condition of \texttt{dfss} is satisfied.
+Otherwise, the proof is split according to the cases in the definition of $e'$.
+In the first case, $w$ is in the set of explored nodes of $e$.
+Hence, $e' = e$.
+The proof is rather straightforward thanks to the assumption on \texttt{pre\_dfss}.
+The following case, \textit{i.e.} $w \in \texttt{visited }e$ is much more complicated.
+In this case, $e' = \texttt{dfs}~v~e$.
 

### 2022-06-01 LaTeX/report.tex
Updates

@@
 Note that the function written in Isabelle is different from the \textsc{Unite} function introduced earlier.
-From now on, \textsc{Unite} designates the following function, which was first written as a recursive function as follows:
+From now on, \texttt{unite} designates the following function, which was first written as a recursive function as follows:
 
@@
 \BlankLine
-
+\label{unite_def}
 \isabelle{
@@
 In particular, the last conjunct expresses the fact that the equivalence relation on the stack is monotonic.
+This is a useful property as it explicitly gives a monotonic behaviour on the environment structure -- hence its name -- during the execution of the algorithm, even though we do not have an inductive rule on environments.
+Besides, this property is transitive, which will be used in the proofs.
 
@@
     \item[$\bullet$] The new stack is a suffix of the previous one (\textit{i.e.} it expressively represents the structure of FIFO stack).
-    \item[$\bullet$] This one is a little more difficult to understand : it expresses the fact that if two nodes $m$ and $n$ such that $m \preceq n$ in the stack before the call on \texttt{dfs}, and if for all node $u$ on the stack, $u \Rightarrow^* v$ and $v \Rightarrow^* n$, then the execution of \texttt{dfs} has updated $\mathcal{S}$ such that $\mathcal{S}(m) = \mathcal{S}(n)$ in the environment returned by the function.
-This should be equated with the fourth conjunct : $v$ is reachable from all nodes on stack, and the fact that partials SCCs have a unique representative on stack.
     \item[$\bullet$] The last one is easier to understand: either $v$ is explored and the stack considered before and after the execution of the function has not changed, or the head of the stack is the representative of $v$ and $v$ had a representative in the previous stack.
@@
 Once again, this condition has to be equated with the fact that all nodes on stack can reach $v$ (fifth conjunct in the following definition).
+The definition is as follows:
 
@@
 $~~\land$ (stack \green{e} $\neq$ [])\\
-$~~\land$ ($\forall$ \green{n}$\in$set(stack \green{e}).reachable \green{v n} $\longrightarrow$ \green{v}$\in$$\mathcal{S}$ \green{e n} $\lor$ ($\exists$ \green{m} $\in$ \green{vs}.reachable \green{m n}))\\
 $~~\land$ (\green{v} $\in$ $\mathcal{S}$ \green{e} (hd (stack \green{e})))"
@@
 
-The post-conditions on \texttt{dfss} are less complicated than the post-conditions of \texttt{dfs} and can easily be understood with the explanations of the other invariants.
+The post-conditions on \texttt{dfss} are less complicated than the post-conditions of \texttt{dfs} and can easily be understood with the explanations of the other invariants:
 
@@
 It assumes that $w$ is a successor of $v$ that is not visited, and it shows that the pre-conditions on \texttt{dfs} on $w$ the same environment $e$ are satisfied.
-The proof is quite as all conditions of \texttt{pre\_dfs} can be deduced from the conditions of \texttt{pre\_dfss}.
 
@@
 
+\BlankLine
+\BlankLine
+
+The proof of the lemma is very simple since all conditions of \texttt{pre\_dfs} can be deduced from the conditions of \texttt{pre\_dfss}.
+
 \subsubsection{\texttt{pre\_dfs} implies \texttt{post\_dfs}}\label{lemma:pre_post_dfs}
+This lemma is probably the most important lemma -- and the most difficult to show -- in the proof of the algorithm.
+It assumes that the pre-condition on \texttt{dfs} are satisfied and shows that the post-condition on \texttt{dfs} are satisfied.
+It is stated as follows:
+
+\isabelle{
+\lm{} pre\_dfs\_implies\_post\_dfs:\\
+$~~~$\fixes{} \blue{v e}\\
+$~~~$\green{defines} "\blue{e1} $\equiv$ \blue{e}\env{visited := visited \blue{e} $\cup$ \{v\}, stack := (v \# stack \blue{e})}"\\
+$~~~$\green{defines} "\blue{e'} $\equiv$ dfss \blue{v} (successors \blue{v}) \blue{e1}"\\
+$~~~$\assumes{} 1: "pre\_dfs \blue{v e}"\\
+$~~~~~~~$\and{} 2: "dfs\_dfss\_dom (Inl(\blue{v}, \blue{e}))"\\
+$~~~~~~~$\and{} 3: "post\_dfss \blue{v} (successors \blue{v}) \blue{e1 e'}"\\
+$~~~$\shows{} "post\_dfs \blue{v e} (dfs \blue{v e})"
+}
+
+\BlankLine
+\BlankLine
+
+Note that the definitions of $e_1$ and $e'$ follow from the definition of \texttt{dfs} which relies -- in an \textit{if} statement -- on the fact that $v = \text{hd } (\text{stack } e)$ or not, therefore the proof also considers this split of cases.
+
+\BlankLine
+
+The first case assumes that $v = \text{hd}(\text{stack }e')$.
+Then with the second assumption ($v$ and $e$ are in the domain of definition of \texttt{dfs}), we can show that \texttt{dfs} returns the following environment:
+\BlankLine
+\noindent
+$\texttt{dfs}~v~e = e'(\!|\texttt{sccs}:=\texttt{sccs}~e' \cup \{\mathcal{S}~e'~v\},\\
+~~~~~~~~~~~~~~~~~~\texttt{explored}:=\texttt{explored}~e' \cup (\mathcal{S}~e'~v),\\
+~~~~~~~~~~~~~~~~~~\texttt{stack}:=\texttt{tl}(\texttt{stack}~e')|\!)"$
+\BlankLine
+That is, it returns an environment $e_2 := \text{dfs}~v~e$ obtained from $e'$ in which a new SCC, namely $\mathcal{S}~e'~v$, is added.
+The different conjuncts in the definition of \texttt{post\_dfs} are then shown one by one.
+Some useful things to note -- and prove -- are the following:
+
+\begin{itemize}
+    \item $\text{stack }e_2 = \text{tl}(\text{stack }e')$, and since $v$ is the head of the stack of $e'$, we can show that $\text{stack }e_1 = \text{stack }e'$.
+The main idea is to use, from the assumption 3, the fact that:
+    \begin{equation*}
+        \exists~ns,~\text{stack } e_1 = ns~@~(\text{stack }e')
+    \end{equation*}
+
+    \item The equivalence relation is the same for $e'$ and $e_2$: $\forall x, \mathcal{S}~e'~x = \mathcal{S}~e_2~x$
+    \item It is easy to show that $e$ is a sub-environment of $e_1$ (i.e.
+$\texttt{sub\_env}~e~e_1$).
+We can also show that $e_1$ is a sub-environment of $e_2$, which is much more difficult.
+However, the proofs in Isabelle are quite simple to read and understand.
+\textit{In fine}, by transitivity of the relation \texttt{sub\_env}, $e$ is a sub-environment of $e_2$.
+    \item Showing that $\mathcal{S}~e'~v$ is a maximal SCC in the environment $e_2$ is a tricky part.
+The proof uses a contradiction-based reasoning and the fact that the equivalence relation is the same for $e'$ and $e_2$ to find a node being in a bigger SCC containing $\mathcal{S}~e'~v$ but not in $\mathcal{S}~e'~v$, and exploit the reachability from and to $v$ and the assumption 3.
+\end{itemize}
+
+In the second case, $v$ is not the head of the stack of $e'$, therefore $\texttt{dfs}~v~e = e'$ and the proof is much easier thanks to the post-condition on \texttt{dfss}, \textit{i.e.} assumption 3.
 
@@
     \item $\texttt{pre\_dfss}~v~vs~e$
-    \item $vs \neq \varnothing$ and \\$\forall w \in vs,  w \notin \texttt{explored } e~\land~w \notin \texttt{visited } e~\land~\texttt{pre\_dfs}~w~e \Longrightarrow \texttt{post\_dfs}~w~e~(dfs~w~e)$
+    \item $vs \neq \varnothing$ and \\$\forall w \in vs,  w \notin \texttt{explored } e~\land~w \notin \texttt{visited } e~\land~\texttt{pre\_dfs}~w~e \Longrightarrow \texttt{post\_dfs}~w~e~(\texttt{dfs}~w~e)$
     \item let $e' =
@@
 The proof is rather straightforward thanks to the assumption on \texttt{pre\_dfss}.
+
+\BlankLine
+
 The following case, \textit{i.e.} $w \in \texttt{visited }e$ is much more complicated.
 In this case, $e' = \texttt{dfs}~v~e$.
+Altgouh the proof remains straightforward, the different properties are more difficult to establish.
+The main idea is to use the fact that $\texttt{post\_dfs}~w~e~e'$ is satisfied to show the pre-condition : $\texttt{pre\_dfss}~v~vs-\{w\}~e$.
+From the assumptions, we can obtain the post-condition : $\texttt{post\_dfss}~v~vs-\{w\}~e~e'$.
+There still remains some work to do inorder to show that this implies the final post-condition, \textit{i.e.} that $\texttt{post\_dfss}~v~vs-\{w\}~e~(\texttt{dfss}~v~vs-\{w\}~e)$ is satisfied.
+
+\BlankLine
+
+In the last case, the function \texttt{unite} is called to compute the new environment $e'$.
+The proof uses the local variables defined in the definition of \texttt{unite}, \textit{i.e.} \texttt{pfx}, \texttt{sfx} and \texttt{cc} (see \ref{unite_def}).
+In particular, it is important to note that $\texttt{stack e} = \texttt{pfx } @ \texttt{ sfx}$ and $w$ is represented by a node in the stack being the head of \texttt{sfx}.
+A visual representation of the stack is given in \textsc{Figure \ref{fig:partial_correctness_unite}}.
+Morally, it is easy to understand that because of the cycle drawn on \textsc{Figure \ref{fig:partial_correctness_unite}}, $\mathcal{S}$ should be updated in the new environment so that all components of $\mathcal{S}$ for each node of \texttt{sfx} should be merged together.
+
+\begin{figure}[!h]
+    \ctikzfig{partial_correctness_unite}
+    \caption{Representation of \texttt{stack e} in the case $w \in \texttt{ visited } e$.
+The black arrows represent reachability in both directions because the definition of a representative on the stack states that both nodes are in the same component of $\mathcal{S}$ for the environment $e$.}\label{fig:partial_correctness_unite}
+\end{figure}
+
+\BlankLine
+
+The approach is similar to the previous case : we prove the pre-condition on \texttt{dfss} and use the hypotheses to show the post-condition of \texttt{dfss}.
+The rest consists in proving the post-condition of \texttt{dfss} on $v$, $vs - \{w\}$ and the new environment $e'$ and the environment return by \texttt{dfss} implies the final post-condition.
 
@@
 
-\end{document}
+\end{document}x

### 2022-06-01 LaTeX/report.tex
Updates

@@
 In this case, $e' = \texttt{dfs}~v~e$.
-Altgouh the proof remains straightforward, the different properties are more difficult to establish.
+Although the proof remains straightforward, the different properties are more difficult to establish.
 The main idea is to use the fact that $\texttt{post\_dfs}~w~e~e'$ is satisfied to show the pre-condition : $\texttt{pre\_dfss}~v~vs-\{w\}~e$.
@@
 \section{Conclusion}
-
-\ednote{sm}{Don't forget to write a conclusion, explaining what has been done, what is missing / could be improved, and what your experience has been.}
+\subsection{What has been done ?}
+The proof of correctness is almost done.
+The model and the invariants we have come up with are quite comprehensive and coherent now.
+We regularly test newer invariants with TLC, a model checker for specifications written in TLA+.
+Thus, we also have a TLA+ version of this algorithm, although it is not really the same implementation.
+However, as we juste use it to verify a few properties on specific examples, it is just meant to give us a better feeling of confidence or find counter examples.
+I've been working on this project for almost a year, and yet it's only been a month or two that I really feel like I understand the algorithm.
+
+\subsection{What is missing ?}
+In the proof of correctness, some intermediate proofs are still skipped. It is mainly due to some former invariants that were deleted because they were wrong.
+
+\BlankLine
+
+Another important missing piece is the proof of termination of the functions.
+In order to achieve this, we need to define a measure and find a quantity that strictly decreases for that measure.
+In \texttt{sub\_env}, the inclusions are not strict.
+An idea would be to consider the complementary set of visited between two calls of \texttt{dfs} as a set getting strictly smaller.
+
+\subsection{My experience}
+Before starting the project, I had no knowledge of formal methods.
+I had to learn it quickly in one month and as I went along.
+I liked it right away.
+As I am very attached to rigour, there is a very addictive side to wanting to demonstrate everything.
+This is why I decided to pursue my career in the field of formal methods.
+\BlankLine
+It was also my first experience in the world of research.
+It was an opportunity to confirm my desire to do basic research, and I also found a domain.
+I was also able to meet several of the researchers of the \href{https://www.loria.fr/fr/la-recherche/les-equipes/mosel-veridis/}{MOSEL-VERIDIS} team at \href{https://www.loria.fr/en/}{LORIA} in Nancy, as well as \href{https://cs.au.dk/~jaco/}{Jaco Van de Pol} who is the supervised the thesis of Vincent Bloemen \cite{bloemen_strong_2019}.
 

### 2022-06-02 LaTeX/report.tex
Updates

@@
     $~~~~~${\color{isa_blue}{in if}} {\color{isa_dark_green}v} = hd(stack {\color{isa_dark_green}e'})\\
-    $~~~~~~~~~~${\color{isa_blue}{then}} {\color{isa_dark_green}e'}\env{sccs:=sccs {\color{isa_dark_green}e'} $\cup$ {$\mathcal{S}$ {\color{isa_dark_green}e' v}}, explored:=explored {\color{isa_dark_green}e'} $\cup$ ($\mathcal{S}$ {\color{isa_dark_green}e' v}), stack:=tl(stack {\color{isa_dark_green}e'})}\\
+    $~~~~~~~~~~${\color{isa_blue}{then}} {\color{isa_dark_green}e'}\env{sccs:=sccs {\color{isa_dark_green}e'} $\cup$ \{$\mathcal{S}$ {\color{isa_dark_green}e' v}\}, explored:=explored {\color{isa_dark_green}e'} $\cup$ ($\mathcal{S}$ {\color{isa_dark_green}e' v}), stack:=tl(stack {\color{isa_dark_green}e'})}\\
     $~~~~~~~~~~${\color{isa_blue}else} {\color{isa_dark_green}e'})"\\

### 2022-08-11 Isabelle/document/root.tex

\documentclass[11pt,a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{isabelle,isabellesym}

% further packages required for unusual symbols (see also
% isabellesym.sty), use only when needed

%\usepackage{amssymb}
  %for \<leadsto>, \<box>, \<diamond>, \<sqsupset>, \<mho>, \<Join>,
  %\<lhd>, \<lesssim>, \<greatersim>, \<lessapprox>, \<greaterapprox>,
  %\<triangleq>, \<yen>, \<lozenge>

%\usepackage{eurosym}
  %for \<euro>

%\usepackage[only,bigsqcap,bigparallel,fatsemi,interleave,sslash]{stmaryrd}
  %for \<Sqinter>, \<Parallel>, \<Zsemi>, \<Parallel>, \<sslash>

%\usepackage{eufrak}
  %for \<AA> ... \<ZZ>, \<aa> ... \<zz> (also included in amssymb)

%\usepackage{textcomp}
  %for \<onequarter>, \<onehalf>, \<threequarters>, \<degree>, \<cent>,
  %\<currency>

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

% for uniform font size
%\renewcommand{\isastyle}{\isastyleminor}

\begin{document}

\title{Correctness of an algorithm for computing strongly connected components of a graph}
\author{Stephan Merz and Vincent Trélat}
\maketitle

\tableofcontents

% sane default for proof documents
\parindent 0pt\parskip 0.5ex

% generated text of all theories
\input{session}

% optional bibliography
%\bibliographystyle{abbrv}
%\bibliography{root}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2022-08-16 Isabelle/document/root.tex
refactored proofs and explanations

@@
 
-\title{Correctness of an algorithm for computing strongly connected components of a graph}
+\title{A Set-based Algorithm for Computing Strongly Connected Components of a Graph}
 \author{Stephan Merz and Vincent Trélat}
 \maketitle
+
+\begin{abstract}
+  We prove the correctness of a sequential algorithm for computing strongly
+  connected components (SCCs) of a graph due to Vincent Bloemen.
+\end{abstract}
 
@@
 
-% optional bibliography
+% optional bibliography ... doesn't appear to work
 %\bibliographystyle{abbrv}
 %\bibliography{root}
+
+\begin{thebibliography}{9}
+\bibitem{bloemen:strong}
+Vincent Bloemen.
+\newblock Strong Connectivity and Shortest Paths for Checking Models.
+\newblock University of Twente, Enschede, The Netherlands, 2019.
+  \url{https://ris.utwente.nl/ws/portalfiles/portal/122499728/thesis.pdf}
+
+\bibitem{dijkstra:finding}
+Edsger W. Dijkstra.
+\newblock Finding the Maximum Strong Components in a Directed Graph.
+\newblock In: Selected Writings in Computing -- A Personal Perspective,
+  pp.\ 22-30.
+  Texts and Monographs in Computer Science. Springer, 1982.
+
+\bibitem{munro:efficient}
+J. Ian Munro.
+\newblock Efficient Determination of the Transitive Closure of a Directed Graph.
+\newblock \emph{Information Processing Letters} 1(2):56-58, 1971.
+
+\bibitem{tarjan:depth-first}
+Robert Tarjan.
+\newblock Depth first search and linear graph algorithms.
+\newblock \emph{SIAM Journal on Computing}, 1972.
+\end{thebibliography}
 

### 2022-08-17 Isabelle/document/root.tex
version of the theories submitted to the AFP

@@
 
-\title{A Set-based Algorithm for Computing Strongly Connected Components of a Graph}
+\title{Correctness of a Set-based Algorithm for Computing Strongly Connected Components of a Graph}
 \author{Stephan Merz and Vincent Trélat}
@@
 \begin{abstract}
-  We prove the correctness of a sequential algorithm for computing strongly
-  connected components (SCCs) of a graph due to Vincent Bloemen.
+  We prove the correctness of a sequential algorithm for computing 
+  maximal strongly connected components (SCCs) of a graph 
+  due to Vincent Bloemen.
 \end{abstract}
@@
 
+\bibitem{lammich:gabow}
+Peter Lammich.
+\newblock Verified Efficient Implementation of Gabow's Strongly Connected Components Algorithm.
+\newblock \emph{Archive of Formal Proofs}, 2014.
+  \url{https://isa-afp.org/entries/Gabow_SCC.html}.
+
 \bibitem{munro:efficient}

### 2022-09-01 cpp/submission.tex
initial version of CPP submission

\documentclass[sigplan,10pt,anonymous,review]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

\let\Bbbk\relax
\usepackage{isabellelst}

\title{Our Beautiful Paper}

\author{Stephan Merz}
\orcid{0000-0003-0974-1844}
\affiliation{
  \institution{University of Lorraine, CNRS, Inria, LORIA}
  \city{Nancy}
  \country{France}
}

\author{Vincent Trélat}
%\orcid{0000-0003-0974-1844}
\affiliation{
  \institution{University of Lorraine}
  \city{Nancy}
  \country{France}
}

\keywords{graph algorithm, strongly connected component, formal verification, interactive theorem proving}

\newcommand{\prog}[1]{\textit{#1}}
\renewcommand{\SS}{\mathcal{S}}

% ----------------------------------------------------------------------
\begin{document}
\begin{abstract}
  We formalize an algorithm for computing strongly connected components in a graph due to Bloemen and prove its correctness in the proof assistant Isabelle/HOL.
\end{abstract}

\maketitle

\section{Introduction}
\label{sec:introduction}

\section{Background}
\label{sec:background}

\subsection{Graphs and strongly connected components}
\label{sec:graphs}

\subsection{Isabelle/HOL}
\label{sec:isabelle}

\section{Formalization of the Algorithm in Isabelle/HOL}
\label{sec:formalization}

We represent graphs with nodes of type $\nu$ as a locale that fixes a finite set of vertices and a successor function associating to each node its set of successors.

\begin{small}
\begin{lstlisting}[language=isabelle]
locale graph =
  fixes   vertices :: 'v set
  and     successors :: 'v => 'v set
  assumes finite vertices  
  and     forall v sin vertices. successors v subseteq vertices
\end{lstlisting}
\end{small}

The data on which the algorithm operates is collected in an ``environment'' record with the following components:

\begin{small}
\begin{lstlisting}[language=isabelle]
record 'v env =
  root :: 'v
  SS :: 'v => 'v set
  visited :: 'v set
  explored :: 'v set
  sccs :: 'v set set
  stack :: 'v list
  cstack :: 'v list
  vsuccs :: 'v => 'v set
\end{lstlisting}
\end{small}

The \prog{root} component stores the node for which the algorithm was originally called.
The function $\SS$ maps every node $v$ to the set of nodes that have already been determined to be part of the same SCC as $v$. A core invariant of the algorithm will be that this mapping represents a partition of nodes into sets of equivalence classes: for all nodes $v$ and $w$, we maintain the relationship
\[
  v \in \SS~w\ \longleftrightarrow\ \SS~v = \SS~w.
\]
The sets \prog{visited} and \prog{explored} represent the sets of nodes that have already been seen, respectively fully explored, by the algorithm, and \prog{sccs} is the set of (maximal) SCCs that the algorithm has found so far. Variable \prog{stack} holds the roots of the (incomplete) SCCs in depth-first order, whereas \prog{cstack} represents the stack of calls of the main function \prog{dfs} of the algorithm. Finally, \prog{vsuccs} remembers the set of outgoing edges (represented by the successor nodes) of each node that have already been followed during the exploration. Note that the variables \prog{root} and \prog{cstack} are not actually used by the algorithm: they are ghost variables used in the correctness proof.

The algorithm will initially be called with an environment of the form $\prog{init-env}~v$ that holds $v$ as the root node, initializes $\SS$ by assigning the singleton $\{u\}$ to every node $u$, and has all other component initialized to the empty set or the empty list.

Environments are partially ordered according to the following relation.

\begin{small}
\begin{lstlisting}[language=isabelle]
definition sub-env where
sub-env e e' ==
    root e' = root e
  /\ visited e subseteq visited e'
  /\ explored e subseteq explored e'
  /\ (forall v. vsuccs e v subseteq vsuccs e' v)
  /\ (forall v. SS e v subseteq SS e' v)
  /\ (Union {SS e v | v. v sin set (stack e)})
     subseteq (Union {SS e' v | v. v sin set (stack e')})
\end{lstlisting}
\end{small}

We follow the definition of the algorithm in~\cite{bloemen:strong}, but represent it as a pair of mutually recursive functions \prog{dfs} and \prog{dfss} defined as follows.

\begin{small}
\begin{lstlisting}[language=isabelle]
function dfs  :: 'v => 'v env => 'v env
and      dfss :: 'v => 'v env => 'v env where
  dfs v e =
    let e1 = e(| visited := visited cup {v},
                 stack := v # stack e,
                 cstack := v # cstack e |);
        e' = dfss v e1
    in  if v = hd (stack e')
        then e'(| sccs := sccs e' cup {SS e' v},
                  explored := explored e' cup SS e' v,
                  stack := tl (stack e'),
                  cstack := tl (cstack e') |)
        else e'(| cstack := tl (cstack e') |)
| dfss v e =
    let vs = successors v - vsuccs e v
    in  if vs = {} then e
        else let w = SOME x. x sin vs;
                 e' = (if w sin explored e then e
                       else if w notin visited e
                            then dfs w e
                            else unite w e);
                 e'' = e'(| vsuccs := lambda x. 
                              if x = v 
                              then vsuccs e' v cup {w}
                              else vsuccs e' x|)
             in dfss v e''
\end{lstlisting}
\end{small}

Function \prog{dfs} explores a node $v$ by adding it to the set of visited nodes, pushing it on both stacks, and then calling the function \prog{dfss}. If $v$ is still at the head of the stack, it is the root node of an SCC, and therefore $\SS~v$ is added to the set of maximal SCCs, all its nodes are marked as fully explored, and $v$ is popped from both stacks. Otherwise, $v$ is only popped from the call stack. The tail-recursive function \prog{dfss} iterates over all successors $w$ of node $v$. If $w$ has already been fully explored, there is nothing to do. If $w$ is unvisited, the function \prog{dfs} is called for $w$, otherwise the algorithm detected a loop back to a node that has already been visited and whose root node $r$ must therefore be on the stack. It unites the SCCs of all nodes contained in the prefix\footnote{Note that the top of the stack is at the head of the list.} of the stack starting at $r$, using the following auxiliary function.

\begin{small}
\begin{lstlisting}[language=isabelle]
definition unite :: 'v => 'v env => 'v env where
unite w e ==
  let pfx = takeWhile (lambda x. w notin SS e x) (stack e);
      sfx = dropWhile (lambda x. w notin SS e x) (stack e);
      cc = Union { SS e x | x . x sin set pfx cup {hd sfx} }
  in  e(| SS := lambda x. if x sin cc then cc else SS e x;
          stack := sfx |)
\end{lstlisting}
\end{small}

In words, \prog{unite} computes the prefix of the stack consisting of the roots of SCCs that do not contain $w$. Since $w$ must be represented by some node on the stack, the remaining suffix is non-empty, and $\SS~(\prog{hd}~\prog{sfx})$ contains $w$. A new connected component $\prog{cc}$ is computed by taking the union of the sets $\SS~x$ for all nodes $x$ in the prefix and the head of the suffix. The mapping $\SS$ is updated so that all nodes in $\prog{cc}$ are mapped to $\prog{cc}$, and the stack is shortened to the suffix of the previous stack, implicitly making the head of that suffix the root of the component $cc$.

A technical complication in the definition of the algorithm is that the functions \prog{dfs} and \prog{dfss} need not terminate when their pre-conditions are violated. We will come back to proving termination in Sect.~\ref{sec:termination}.

\section{Proof of Partial Correctness}
\label{sec:partial-correctness}

\subsection{Main invariant}
\label{sec:invariant}

We define well-formed environments to be those satisfying the following conditions, and we will prove this predicate to hold throughout the execution of the algorithm.

\begin{small}
\begin{lstlisting}[language=isabelle]
definition wf-env where
wf-env e ==
    (forall n sin visited e. reachable (root e) n)
  /\ distinct (stack e)
  /\ distinct (cstack e)
  /\ (forall n m. n prec m in stack e --> n prec m in cstack e)
  /\ (forall n m. n prec m in stack e --> reachable m n)
  /\ explored e subseteq visited e
  /\ set (cstack e) subseteq visited e
  /\ (forall n sin explored e. forall m. 
        reachable n m --> m sin explored e)
  /\ (forall n. vsuccs e n subseteq successors n cap visited e)
  /\ (forall n. n notin visited e --> vsuccs e n = {})
  /\ (forall n sin explored e. vsuccs e n = successors n)
  /\ (forall n sin visited e - set (cstack e). 
        vsuccs e n = successors n)
  /\ (forall n m. m sin SS e n <--> SS e n = SS e m)
  /\ (forall n. n notin visited e --> SS e n = {n})
  /\ (forall n sin set (stack e). forall m sin set (stack e).
        n noteq m --> SS e n cap SS e m = {})
  /\ Union {SS e n | n. n sin set (stack e)} 
     = visited e - explored e
  /\ (forall n in set (stack e). forall m in SS e n.
        m sin set (cstack e) --> m prec n in cstack e)
  /\ (forall n m. n prec m in stack e /\ n noteq m -->
        (forall u sin SS e n. ~ reachable-avoiding u m 
                                    (unvisited e n)))
  /\ (forall n. is-subscc (SS e n))
  /\ (forall S sin sccs e. is-scc S)
  /\ Union (sccs e) = explored e
\end{lstlisting}
\end{small}

In words, every visited node is reachable from the root node, the two stacks do not contain duplicate elements, and the stack is a subsequence of the call stack. Nodes on the stack are reachable from all stack nodes below them. All explored nodes, as well as all nodes on the call stack, are visited. All nodes reachable from fully explored nodes are themselves fully explored. The visited successors of a node are both successors and visited, and no outgoing edge of an unvisited node has been followed. All outgoing edges of fully explored nodes, as well as of visited nodes that are no longer on the call stack, have been followed. The mapping $\SS$ represents an equivalence relation, as discussed above, and $\SS~n$ is the singleton $\{n\}$ for an unvisited node $n$. The nodes on the stack are roots of (non-maximal) SCCs, hence their equivalence classes are disjoint. A node has been visited but not yet fully explored iff its root node is on the stack. Moreover, root nodes of SCCs are the oldest nodes in depth-first order. No node $u$ belonging to the SCC of a node $n$ on the stack can reach a node $m$ strictly below $n$ on the stack without going through some edge from some node in $n$'s equivalence class that has not yet been followed. In other words, the current SCCs are maximal with respect to the current knowledge of the algorithm. The set of not yet followed edges for nodes in the equivalence class of $n$ are defined as
%
\begin{small}
\begin{lstlisting}[language=isabelle]
definition unvisited where
unvisited e n ==
  {(a,b) | a b. a sin SS e n 
              /\ b sin successors a - vsuccs e a}
\end{lstlisting}
\end{small}
%
Finally, every equivalence class is a (not necessarily maximal) SCC, elements of \prog{sccs} are maximal SCCs, and their union is the set of explored nodes.

%% Not sure if we need this ...
We prove a number of simple consequences of this predicate. For example, $n \in \SS~n$ holds for every node. If node $w$ is reachable from a visited node $v$, but not through any edge that has not yet been followed, then $w$ must be visited. Also, edges towards fully explored nodes do not contribute to reachability of unexplored nodes avoiding certain edges:
%
\begin{small}
\begin{lstlisting}[language=isabelle]
lemma avoiding explored:
assumes wf-env e and reachable-avoiding x y E
    and y notin explored e and w sin explored e
shows   reachable-avoiding x y (E cup {(v,w)})
\end{lstlisting}
\end{small}

\subsection{Pre- and post-conditions of functions \prog{dfs} and \prog{dfss}}
\label{sec:pre-post}

\section{Proof of Termination}
\label{sec:termination}

\section{Conclusion}
\label{sec:conclusion}

\bibliography{bib}
\bibliographystyle{abbrv}

\end{document}

### 2022-09-11 cpp/submission.tex
Added some prose in the introduction and background sections

@@
 \label{sec:introduction}
+Directed graphs are widely used in many fields of computer science and mathematics.
+The study of their properties is therefore an important topic because of their application in areas that may affect our lives.
+
+They are also heavily used in critical domains that indirectly affect us, such as model checking.
+Indeed, as model checking involves exhaustive exploration of graphs of states, it is important to have efficient -- and correct -- algorithms to perform different operations such as the computation of strongly connected components.
 
@@
 \label{sec:graphs}
+Graphs referred to in this paper are finite directed graphs.
+The proof relies on the notion of reachability in a graph, which is defined inductively through the notion of successor in a directed graph.
+
+A strongly connected component (SCC) of such a graph is a subset of its set of vertices such that two vertices of it are symmetrically reachable from each other.
+More specifically, if $\mathcal{V}$ is the set of vertices of a graph $\mathcal{G}$ and if $\Rightarrow^*$ denotes reachability as the reflexive and transitive closure of the \emph{successor} binary relation, then a subset $\mathcal{C}$ of $\mathcal{V}$ is a SCC if:
+\begin{equation*}
+  \forall u, v \in \mathcal{C} \quad u \Rightarrow^* v \wedge v \Rightarrow^* u
+\end{equation*}
+Informally, SCCs roughly denote -- or a a generalisation of -- cycles in a graph.
 
@@
 \label{sec:isabelle}
+
+Isabelle is a generic proof assistant \cite{nipkow_isabellehol_2002} allowing mathematical formulas to be expressed in a formal language.
+It provides tools for proving those formulas in a logical calculus.
+Isabelle proofs are written in the Isar (``Intelligible Semi-Automated Reasoning'') language that is designed to make proofs readable and comprehensible for a mathematically inclined reader, with minimal overhead introduced by the formalism.
 

### 2022-09-11 cpp/submission.tex
Added prose about pre- post-conditions

@@
 Graphs referred to in this paper are finite directed graphs.
-The proof relies on the notion of reachability in a graph, which is defined inductively through the notion of successor in a directed graph.
+The proof relies on the notion of reachability in a graph, which is defined inductively in Isabelle through the notion of successor in a directed graph.
 
@@
 
+The proof of correctness relies on the fact that a call of \prog{dfs} for arguments that satisfy some hypotheses (the pre-conditions) terminates and satisfies some properties (the post-conditions).
+Because of the mutual recursion, we must define pre- and post-conditions for both \prog{dfs} and \prog{dfss}.
+
+First, we define pre- and post-conditions for function \prog{dfs}.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+definition pre-dfs where 
+  pre-dfs v e ==
+    wf-env e
+  /\ v notin visited e
+  /\ reachable (root e) v
+  /\ vsuccs e v = {}
+  /\ (forall n sin set (stack e). reachable n v)
+\end{lstlisting}
+\end{small}
+
+Function \prog{dfs} should be called for a well-formed environment and a node $v$ that has not yet been visited and that is reachable from the root node, as well as from all nodes in the stack.
+No outgoing edges from node $v$ have yet been followed.
+
+\begin{small}
+  \begin{lstlisting}[language=isabelle]
+definition post-dfs where 
+  post-dfs v e e' ==
+    wf-env e'
+  /\ v sin visited e'
+  /\ sub-env e e'
+  /\ vsuccs e' v = successors v
+  /\ (forall w sin visited e. vsuccs e' w = vsuccs e w)
+  /\ (forall n sin set (stack e'). reachable n v)
+  /\ (exists ns. stack e = ns @ (stack e'))
+  /\ ((v sin explored e' /\ stack e' = stack e 
+      /\ (forall n sin set (stack e'). SS e' n = SS e n)) 
+    \/ (stack e' noteq [] /\ v sin SS e' (hd (stack e')) 
+      /\ (forall n sin set (tl(stack e')). SS e' n = SS e n)))
+  /\ cstack e' = cstack e
+  \end{lstlisting}
+  \end{small}
+
+Function \prog{dfs} maintains the invariant \prog{wf-env} and returns an environment $e'$ that extends the input environment $e$.
+Node $v$ has been visited and all its outgoing edges have been followed.
+Because the algorithm works in depth-first fashion, no new outgoing edges of nodes that had already been visited in the input environment have been followed, and the stack of $e'$ is a suffix of the one of $e$ such that $v$ is still reachable from all nodes on the stack.
+The stack may have been shortened because SCCs represented at the top of the stack may have been merged.
+The call stack is reestablished as it was in $e$.
+There are two possible outcomes of the algorithm:
+\begin{itemize}
+\item Either $v$ has been fully explored, in which case the stacks of $e$ and $e'$ are the same, and the equivalence classes of all nodes represented on the stack are unchanged.
+This corresponds to the case where $v$ is the root node of its (maximal) SCC.
+\item Alternatively, the stack of $e'$ must be non-empty and $v$ must be represented by the node at the top of the stack.
+The SCCs of the nodes lower on the stack are unchanged.
+This corresponds to the case where $v$ is not the root node of its SCC, but some SCCs at the top of the stack may have been merged.
+\end{itemize}
+
+We then define pre- and post-conditions for function \prog{dfss}.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+definition pre-dfss where 
+  pre-dfss v e ==
+    wf-env e 
+  /\ v sin visited e
+  /\ (stack e snoteq [])
+  /\ (v sin SS e (hd (stack e)))
+  /\ (forall w sin vsuccs e v.
+      w sin explored e cup SS e (hd(stack e)))
+  /\ (forall n sin set (stack e). reachable n v)
+  /\ (exists ns. cstack e = v # ns)
+\end{lstlisting}
+\end{small}
+
+The pre- and post-conditions of function \prog{dfss} correspond to the invariant of the loop over all outgoing edges from node $v$.
+The environment must be well-formed, node $v$ must be visited and represented by the top element of the (non-empty) stack, it must be reachable from all nodes on the stack, and it must be the top node on the call stack.
+All outgoing edges of node $v$ that have already been followed must either lead to completely explored nodes (that are no longer represented on the stack) or to nodes that are part of the same SCC as $v$.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+definition post-dfss where 
+  post-dfss v e e' == 
+     wf-env e'
+   /\ vsuccs e' v = successors v
+   /\ (forall w sin visited e - {v}. vsuccs e' w = vsuccs e w)
+   /\ sub-env e e'
+   /\ (forall w sin successors v.
+        w sin explored e' cup SS e' (hd (stack e')))
+   /\ (forall n sin set (stack e'). reachable n v)
+   /\ (stack e' noteq [])
+   /\ (exists ns. stack e = ns @ (stack e'))
+   /\ v sin SS e' (hd (stack e'))
+   /\ (forall n sin set (tl (stack e')). SS e' n = SS e n)
+   /\ (hd (stack e') = v <-->
+        (forall n sin set (tl (stack e')). ~ reachable v n))
+   /\ cstack e' = cstack e
+\end{lstlisting}
+\end{small}
+
+The post-condition establishes that all outgoing edges of node $v$ have been followed.
+As for function \prog{dfs}, no new outgoing edges of previously visited nodes have been followed, the new stack is a suffix of the old one, and the call stack is restored.
+In case node $v$ is still on the stack (and therefore is the root node of its SCC), no node that is lower on the stack can be reachable from $v$.
+This condition guarantees the maximality of the computed SCCs.
+
+\subsection{Pre-conditions implie pre-conditions}
+Because of the mutual recursion of functions \prog{dfs} and \prog{dfss}, function \prog{dfss} may be innerly called from a call of function \prog{dfs} and vice versa.
+We therefore need to establish that the pre-conditions of \prog{dfs} and \prog{dfss} imply the pre-conditions of the other function.
+
+\subsection{Pre-conditions imply post-conditions}
+
 \section{Proof of Termination}

### 2022-09-12 cpp/submission.tex
progress on CPP submission

--- 2022-09-11 cpp/submission.tex
@@
 
-\title{Our Beautiful Paper}
+\title{Verification in Isabelle/HOL of a Set-based Algorithm for Computing Strongly Connected Components}
 
@@
 \begin{abstract}
-  We formalize an algorithm for computing strongly connected components in a graph due to Bloemen and prove its correctness in the proof assistant Isabelle/HOL.
+  The efficient computation of strongly connected components in a directed graph is a fundamental algorithmic problem with applications in many fields of computer science.
+We report on the formal verification of a recent algorithm proposed by Bloemen in the proof assistant Isabelle/HOL.
+Our work is intended as a stepping stone towards a correctness proof of the parallel version of that algorithm, which has been implemented in the high-performance model checker LTSmin.
 \end{abstract}
@@
   dfs v e =
-    let e1 = e(| visited := visited cup {v},
+    let e1 = e(| visited := visited e cup {v},
                  stack := v # stack e,
@@
 
+Internally, Isabelle generates a single function \prog{dfs-dfss} whose argument type is the sum of the argument types for the individual functions, and defines the latter as projections of the combined function.
+Auto-generated theorems, such as computational induction, simultaneously handle both functions in order to enable mutual induction.
 A technical complication in the definition of the algorithm is that the functions \prog{dfs} and \prog{dfss} need not terminate when their pre-conditions are violated.
@@
 We prove a number of simple consequences of this predicate.
-For example, $n \in \SS~n$ holds for every node.
+For example, $n \in \SS~n$ holds for every node~$n$.
 If node $w$ is reachable from a visited node $v$, but not through any edge that has not yet been followed, then $w$ must be visited.
@@
 \begin{lstlisting}[language=isabelle]
-lemma avoiding explored:
+lemma avoiding-explored:
 assumes wf-env e and reachable-avoiding x y E
@@
 \label{sec:pre-post}
+
+The proof of correctness relies on the fact that a call of \prog{dfs} for arguments that satisfy some hypotheses (the pre-conditions) terminates and satisfies some properties (the post-conditions).
+Because of the mutual recursion, we must define pre- and post-conditions for both \prog{dfs} and \prog{dfss}.
+
+First, we define pre- and post-conditions for function \prog{dfs}.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+definition pre-dfs where 
+  pre-dfs v e ==
+    wf-env e
+  /\ v notin visited e
+  /\ reachable (root e) v
+  /\ (forall n sin set (stack e). reachable n v)
+\end{lstlisting}
+\end{small}
+
+Function \prog{dfs} should be called for a well-formed environment and a node $v$ that has not yet been visited and that is reachable from the root node, as well as from all nodes in the stack.
+% In fact, this is a consequence of wf-env e and v \notin visited e, and can be dropped.
+% No outgoing edges from node $v$ have yet been followed.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+definition post-dfs where 
+  post-dfs v e e' ==
+    wf-env e'
+  /\ sub-env e e'
+  /\ v sin visited e'
+  /\ vsuccs e' v = successors v
+  /\ (forall w sin visited e. vsuccs e' w = vsuccs e w)
+  /\ (exists ns. stack e = ns @ (stack e'))
+  /\ cstack e' = cstack e
+  /\ ((v sin explored e' /\ stack e' = stack e 
+      /\ (forall n sin set (stack e'). SS e' n = SS e n)) 
+    \/ (stack e' noteq [] /\ v sin SS e' (hd (stack e')) 
+      /\ (forall n sin set (tl (stack e')). SS e' n = SS e n)))
+\end{lstlisting}
+\end{small}
+
+Function \prog{dfs} maintains the invariant \prog{wf-env} and returns an environment $e'$ that extends the input environment $e$.
+Node $v$ has been visited and all its outgoing edges have been followed.
+Because the algorithm works in depth-first fashion, no new outgoing edges of nodes that had already been visited in the input environment have been followed, and the stack of $e'$ is a suffix of the stack of $e$.
+% Again, this follows from the pre-condition and the suffix property.
+% such that $v$ is still reachable from all nodes on the stack. 
+The stack may have been shortened because SCCs represented at the top of the stack may have been merged.
+The call stack is reestablished as it was in $e$.
+There are two possible outcomes of the algorithm:
+\begin{itemize}
+\item Either $v$ has been fully explored, in which case the stacks of $e$ and $e'$ are the same, and the partial SCCs of all nodes represented on the stack are unchanged.
+This corresponds to the case where $v$ is the root node of its (maximal) SCC.
+\item Alternatively, the stack of $e'$ must be non-empty and $v$ must be a member of the partial SCC of the node at the top of the stack.
+The SCCs represented by the stack elements other than the top are unchanged.
+This corresponds to the case where $v$ is not the root node of its SCC, and some SCCs at the top of the stack may have been merged.
+\end{itemize}
+
+We then define pre- and post-conditions for function \prog{dfss}.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+definition pre-dfss where 
+  pre-dfss v e ==
+    wf-env e 
+  /\ v sin visited e
+  /\ stack e noteq []
+  /\ v sin SS e (hd (stack e))
+  /\ (forall w sin vsuccs e v.
+      w sin explored e cup SS e (hd (stack e)))
+  /\ (forall n sin set (stack e). reachable n v)
+  /\ (exists ns. cstack e = v # ns)
+\end{lstlisting}
+\end{small}
+
+The pre-condition of function \prog{dfss} corresponds to the invariant of the loop over all outgoing edges from node~$v$.
+The environment is well-formed, node~$v$ has been visited and is represented by the top of the (non-empty) stack, it is reachable from all nodes on the stack, and it is the top node on the call stack.
+All outgoing edges of node $v$ that have already been followed either lead to completely explored nodes or to nodes that are part of the same SCC as~$v$.
+
+Function \prog{dfss} establishes the following post-condition:
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+definition post-dfss where 
+  post-dfss v e e' == 
+     wf-env e'
+   /\ sub-env e e'
+   /\ vsuccs e' v = successors v
+   /\ (forall w sin visited e - {v}. vsuccs e' w = vsuccs e w)
+   /\ (forall w sin successors v.
+        w sin explored e' cup SS e' (hd (stack e')))
+   /\ (forall n sin set (stack e'). reachable n v)
+   /\ (stack e' noteq [])
+   /\ (exists ns. stack e = ns @ (stack e'))
+   /\ cstack e' = cstack e
+   /\ v sin SS e' (hd (stack e'))
+   /\ (forall n sin set (tl (stack e')). SS e' n = SS e n)
+   /\ (hd (stack e') = v -->
+        (forall n sin set (tl (stack e')). ~ reachable v n))
+\end{lstlisting}
+\end{small}
+
+In the same way as function \prog{dfs}, function \prog{dfss} ensures that the resulting environment $e'$ is well-formed and extends the input environment $e$, that all outgoing edges of $v$, and no outgoing edges of previously visited nodes other than $v$ have been followed.
+Again, the new stack is a suffix of the old one, and the call stack is restored.
+In case node $v$ is still on the stack (and therefore is the root node of its SCC), no stack node below $v$ can be reached from $v$.
+This condition is important for ensuring that the computed SCC is maximal.
+
+\subsection{Pre-conditions imply pre-conditions}
+\label{sec:pre-pre}
+
+Given that the two functions \prog{dfs} and \prog{dfss} mutually call each other, we need to establish the pre-condition of the called function, assuming the pre-condition of the calling function.
+
+For example, the following lemma asserts that the pre-condition of \prog{dfss} holds at the call from the body of \prog{dfs}.
+Although some of the conditions of the invariant for the augmented invariant require a few lines of proof, the overall proof is straightforward.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+lemma pre-dfs-pre-dfss:
+  assumes pre-dfs v e
+  shows   pre-dfss v (e(| visited := visited e cup {v},
+                         stack := v # stack e,
+                         cstack := v # cstack e |))
+\end{lstlisting}
+\end{small}
+
+The following lemma asserts the pre-condition of the recursive call to \prog{dfs} from \prog{dfss} for a successor that has not been visited yet.
+Given that the environment of the call is unchanged, the proof of the lemma is entirely automatic.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+lemma pre-dfss-pre-dfs:
+  assumes pre-dfss v e and w sin successors v - visited e
+  shows   pre-dfs w e
+\end{lstlisting}
+\end{small}
+
+Finally, we need to show the pre-condition of \prog{dfss} at the call of \prog{dfss} at the end of the function body, i.e.\ the preservation of the loop invariant.
+We state three separate lemmas according to the branch of the conditional in the computation of the intermediate environment $e'$ in the body of \prog{dfss}.
+As an example, we show the statement of the lemma corresponding to the second branch: since this involves a call to \prog{dfs}, we also assume the post-condition of that call.
+In all three cases, the main effort goes into proving the invariant for the environment~$e''$.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+lemma pre-dfss-post-dfs-pre-dfss:
+  fixes e v w
+  defines e' = dfs w e
+  defines e'' = e'(| vsuccs := (lambda x. if x = v 
+                                then vsuccs e' v cup {w}
+                                else vsuccs e' v) |)
+  assumes pre-dfss v w 
+    and   w sin successors v - visited e
+    and   post-dfs w e e'
+  shows   pre-dfss v e''
+\end{lstlisting}
+\end{small}
+
+\subsection{Pre-conditions imply post-conditions}
+\label{sec:pre-post}
+
+The following theorem is central for the proof of partial correctness: it shows that the functions \prog{dfs} and \prog{dfss} establish their respective post-conditions, assuming that the pre-conditions hold for the input parameters.
+Because we have not proved termination of the functions, we need an additional hypothesis for each statement that, intuitively, asserts that the corresponding function will terminate for its input parameters.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+lemma pre-post:
+  [| dfs-dfss-dom (Inl (v,e)); pre-dfs v e |] 
+     ==> post-dfs v e (dfs v e)
+  [| dfs-dfss-dom (Inr (v,e)); pre-dfss v e |]
+     ==> post-dfss v e (dfss v e)
+\end{lstlisting}
+\end{small}
+
+Since the two functions are mutually recursive, the two statements must be proved by simultaneous computational induction, based on the corresponding proof rule generated by Isabelle.
+The proofs of these implications require substantially more interactions than the proofs of the lemmas of Sect.~\ref{sec:pre-pre}.
+
+As an example, in the first branch of the conditional in the body of function \prog{dfs} we need to prove that all nodes~$y$ reachable from an explored node $x$ are themselves explored.
+The interesting case occurs when $x \in \SS~e'~v$, where $e'$ denotes the environment computed by the call to \prog{dfss}.
+Because $\SS~e'~v$ is a partial SCC, node~$x$ is reachable from node~$v$.
+We first show that $y \in \prog{visited}~e'$.
+Otherwise, given that $x \in \prog{visited}~e'$, by induction we obtain nodes $n \in \prog{visited}~e'$ and $m \in \prog{successors}~n - \prog{vsuccs}~e'~n$ (i.e., $m$ is an unvisited successor of~$n$) such that $n$ is reachable from~$x$ and $y$ is reachable from~$m$.
+But then, we have $n \notin \prog{explored}~e'$ since all successors of explored nodes have been visited, hence there must be a node $n' \in \prog{set}~(\prog{stack}~e')$ such that $n \in \SS~e'~n'$.
+Using the last conjunct of the post-condition \prog{post-dfss}, it follows that $n' = v$, for otherwise $v$ could reach a node deeper in the stack.
+But then, we have $n \in \SS~e'~v \subseteq \prog{explored}~e'$, and according to another conjunct of the invariant that we already proved earlier, we must have $\prog{successors}~n = \prog{vsuccs}~e'~n$, reaching a contradiction.
+
+We now prove $y \in \prog{explored}~e''$, where $e''$ denotes the environment resulting from the call to \prog{dfs}.
+If $y \in \prog{explored}~e'$, this is clearly true.
+Otherwise, we obtain some node $n \in \prog{set}~(\prog{stack}~e')$ such that $y \in \SS~e'~n$.
+If $n=v$, the assertion follows by the definition of $e''$.
+Otherwise, $n$ is a stack node below $v$ that is reachable from~$v$ via nodes~$x$ and~$y$, contradicting the last conjunct of the post-condition \prog{post-dfss}.
+
+\textbf{Perhaps explain a complicated reasoning step that makes use of lemma avoiding-explored?}
 

### 2022-09-13 cpp/submission.tex
added background, finished partial correctness

@@
 
-\subsection{Graphs and strongly connected components}
-\label{sec:graphs}
-Graphs referred to in this paper are finite directed graphs.
-The proof relies on the notion of reachability in a graph, which is defined inductively in Isabelle through the notion of successor in a directed graph.
-
-A strongly connected component (SCC) of such a graph is a subset of its set of vertices such that two vertices of it are symmetrically reachable from each other.
-More specifically, if $\mathcal{V}$ is the set of vertices of a graph $\mathcal{G}$ and if $\Rightarrow^*$ denotes reachability as the reflexive and transitive closure of the \emph{successor} binary relation, then a subset $\mathcal{C}$ of $\mathcal{V}$ is a SCC if:
-\begin{equation*}
-  \forall u, v \in \mathcal{C} \quad u \Rightarrow^* v \wedge v \Rightarrow^* u
-\end{equation*}
-Informally, SCCs roughly denote -- or a a generalisation of -- cycles in a graph.
-
 \subsection{Isabelle/HOL}
@@
 
-Isabelle is a generic proof assistant \cite{nipkow_isabellehol_2002} allowing mathematical formulas to be expressed in a formal language.
-It provides tools for proving those formulas in a logical calculus.
-Isabelle proofs are written in the Isar (``Intelligible Semi-Automated Reasoning'') language that is designed to make proofs readable and comprehensible for a mathematically inclined reader, with minimal overhead introduced by the formalism.
+Isabelle~\cite{paulson:isabelle} is a logical framework in which different object logics can be encoded.
+Isabelle/HOL~\cite{nipkow_isabellehol_2002} is the instance of Isabelle for higher-order logic, which we use in our work and which we will refer to in the following simply as Isabelle.
+It provides an expressive formal language for representing formal mathematical theories as well as algorithms and their correctness proofs, and it comes with an extensive set of tools for defining \emph{locales} (algebraic structures parameterized by operators and assumptions), (mutually) \emph{recursive functions} etc.
 
-\section{Formalization of the Algorithm in Isabelle/HOL}
-\label{sec:formalization}
+Isabelle belongs to the LCF family of proof assistants whose soundness is ensured by a comparatively small trusted kernel that serves to certify proofs.
+From the user's point of view, Isabelle proofs are written in the Isar (``Intelligible Semi-Automated Reasoning'') language that is designed to make proofs readable and comprehensible for a mathematically inclined reader, with minimal overhead introduced by the formalism.
+Extensive automation is provided by built-in proof procedures based on rewriting and first-order reasoning, as well as the \emph{Sledgehammer}~\cite{blanchette:sledgehammer} add-on that selects potentially relevant lemmas from the context (including the standard proof library), calls external automatic proof backends such as SMT solvers as well as automatic proof tools for first-order and higher-order logic, and in case of success reconstructs their proofs in the trusted kernel.
 
-We represent graphs with nodes of type $\nu$ as a locale that fixes a finite set of vertices and a successor function associating to each node its set of successors.
+\subsection{Graphs and strongly connected components}
+\label{sec:graphs}
+
+We represent finite directed graphs with nodes of type $\nu$ as an Isabelle locale that fixes a finite set of vertices and a successor function associating to each node its set of successor nodes.
 
@@
 \end{small}
+
+Reachability in graphs is defined inductively as follows.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+inductive reachable where
+  reachable-refl: reachable x x
+| reachable-succ: [| y sin successors x; reachable y z |]
+                  ==> reachable x z
+\end{lstlisting}
+\end{small}
+
+Elementary lemmas, such as transitivity of the reachability relation, are easily proved by induction, followed by an application of Isabelle's automatic proof methods.
+For our later proofs, we also need the following variant of reachability between nodes avoiding a set $E$ of edges.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+inductive reachable-avoiding where
+  ra-refl: reachable-avoiding x x E
+| ra-succ: [| y sin successors x; (x,y) notin E; 
+              reachable-avoiding y z E |] 
+           ==> reachable-avoiding x z E
+\end{lstlisting}
+\end{small}
+
+Again, this relation is transitive, it implies reachability, is anti-monotonic in the set $E$, and reachability agrees with reachability avoiding the empty set of edges.
+More interestingly, we show the following lemma about adding an edge to the set $E$.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+lemma ra-add-edge:
+  assumes reachable-avoiding x y E
+  shows   reachable-avoiding x y (E cup {(v,w)})
+        \/ reachable-avoiding x v (E cup {(v,w)})
+           /\ reachable-avoiding w y (E cup {(v,w)})
+\end{lstlisting}
+\end{small}
+
+A partial SCC is a set $S$ of nodes all of which are reachable from each other.
+In particular, a cycle in a graph constitutes a partial SCC.
+A maximal SCC is a non-empty partial SCC none of whose supersets is a partial SCC.
+In the following, we write SCC to denote maximal SCCs.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+definition is-subscc where
+  is-subscc S == forall x sin S. forall y sin S. reachable x y
+
+definition is-scc where
+  is-scc S == S noteq {} /\ is-subscc S
+              /\ (forall S'. S subseteq S' /\ is-subscc S'
+                       --> S' = S)
+\end{lstlisting}
+\end{small}
+
+Sledgehammer finds a one-line proof showing that two SCCs are either disjoint or identical:
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+lemma scc-partition:
+  assumes is-scc S and is-scc S' and x sin S cap S'
+  shows   S = S'
+\end{lstlisting}
+\end{small}
+
+\section{Formalization of Bloemen's Algorithm in Isabelle/HOL}
+\label{sec:formalization}
 
@@
 
-\subsection{Pre-conditions imply pre-conditions}
+\subsection{Lemmas establishing the pre-conditions}
 \label{sec:pre-pre}
+
+First, it is easy to see (and Isabelle proves automatically) that the initial environment satisfies the pre-condition of function \prog{dfs}.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+lemma init-env-pre-dfs: pre-dfs v (init-env v)
+\end{lstlisting}
+\end{small}
 
@@
 For example, the following lemma asserts that the pre-condition of \prog{dfss} holds at the call from the body of \prog{dfs}.
-Although some of the conditions of the invariant for the augmented invariant require a few lines of proof, the overall proof is straightforward.
+Although some of the conditions of the invariant for the augmented environment require a few lines of proof, the overall proof is straightforward.
 
@@
 The following lemma asserts the pre-condition of the recursive call to \prog{dfs} from \prog{dfss} for a successor that has not been visited yet.
-Given that the environment of the call is unchanged, the proof of the lemma is entirely automatic.
+Its proof is entirely automatic, helped by the fact that the environment of the call to \prog{dfs} is unchanged.
 
@@
 We state three separate lemmas according to the branch of the conditional in the computation of the intermediate environment $e'$ in the body of \prog{dfss}.
-As an example, we show the statement of the lemma corresponding to the second branch: since this involves a call to \prog{dfs}, we also assume the post-condition of that call.
-In all three cases, the main effort goes into proving the invariant for the environment~$e''$.
+As an example, we show the statement of the lemma corresponding to the third branch.
 
@@
 \begin{lstlisting}[language=isabelle]
-lemma pre-dfss-post-dfs-pre-dfss:
+lemma pre-dfss-unite-pre-dfss:
   fixes e v w
-  defines e' = dfs w e
+  defines e' = unite w e
   defines e'' = e'(| vsuccs := (lambda x. if x = v 
@@
                                 else vsuccs e' v) |)
-  assumes pre-dfss v w 
-    and   w sin successors v - visited e
-    and   post-dfs w e e'
+  assumes pre-dfss v e
+    and   w sin successors v - vsuccs e
+    and   w sin visited e
+    and   w notin explored e
   shows   pre-dfss v e''
@@
 \end{small}
+
+Some steps in the proof of this lemma require a substantial number of interactions.
+We started by proving several auxiliary lemmas about function \prog{unite}, including the fact that the intermediate environment $e'$ is well-formed.
+Then, we have to show that this invariant also holds for $e''$.
+For example, given two nodes $x \neq y$ such that $x$ precedes $y$ in the stack of $e''$ (which is the same as the stack of $e'$) and a node $u \in \SS~e''~x = \SS~e'~x$, we must show that $y$ can be reachable from $u$ only by following some edge in $\prog{unvisited}~e''~x$.
+Therefore, assume that $y$ is reachable from $u$ avoiding all edges in $\prog{unvisited}~e''~x$.
+The proof proceeds by a case analysis depending on whether $x$ is the head of $\prog{stack}~e'$ or not.
+
+In case it is, the definitions of \prog{unvisited} and of $e''$ imply that either $\prog{unvisited}~e'~x = \prog{unvisited}~e''~x$ or $\prog{unvisited}~e'~x = \prog{unvisited}~e''~x \cup \{(v,w)\}$.
+In the first case, we can conclude directly by appealing to the well-formedness of $e'$.
+In the second case, lemma \prog{ra-add-edge} tells us that either $y$ is reachable from $u$ avoiding edges in $\prog{unvisited}~e'~x$ or that $y$ is reachable from $w$ avoiding the same set of edges.
+In the first case, we again conclude appealing to the well-formedness of $e'$.
+In the second case, we know that $w \in \SS~e'~x$ (since $x$ is the head of $\prog{stack}~e'$) and therefore find another contradiction to the well-formedness of $e'$.
+
+In case $x$ is not the head of $\prog{stack}~e'$, both nodes $x$ and $y$ must already be elements of $\prog{stack}~e$, with $x$ still preceding $y$.
+Also, the pre-condition $\prog{pre-dfss}~v~e$ implies that $v$ is a member of the partial SCC represented by the head of the stack, and using the well-formedness of $e$ it follows that $v \notin \SS~e'~x$ and therefore $\prog{unvisited}~e''~x = \prog{unvisited}~e'~x = \prog{unvisited}~e~x$.
+We can now conclude by appealing to the well-formedness of the environment $e$.
 
@@
 Since the two functions are mutually recursive, the two statements must be proved by simultaneous computational induction, based on the corresponding proof rule generated by Isabelle.
-The proofs of these implications require substantially more interactions than the proofs of the lemmas of Sect.~\ref{sec:pre-pre}.
+The proofs of these implications again involve a certain number of steps requiring detailed interaction.
 
@@
 
-\textbf{Perhaps explain a complicated reasoning step that makes use of lemma avoiding-explored?}
+From the lemmas proved so far, we can infer the partial correctness of the function \prog{dfs} when called with the initial environment.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+theorem partial-correctness:
+  assumes dfs-dfss-dom (Inl (v, init-env v)
+  shows   sccs (dfs v (init-env v)) = 
+          {S . is-scc S /\ (forall n sin S. reachable v n)}
+\end{lstlisting}
+\end{small}
+
+In words, assuming that function \prog{dfs} terminates for the input parameters, it computes the set of the maximal SCCs in the sub-graph reachable from node~$v$.
+In particular, if $v$ is a root node of the graph, the maximal SCCs of the graph will be computed.
 

### 2022-09-13 cpp/submission.tex
termination proof

@@
 Indeed, as model checking involves exhaustive exploration of graphs of states, it is important to have efficient -- and correct -- algorithms to perform different operations such as the computation of strongly connected components.
+
+\paragraph{Outline of the paper.}
+
+Section~\ref{sec:background} introduces necessary background for our work, in particular the proof assistant Isabelle/HOL and our representation of finite directed graphs and strongly connected components.
+Our encoding of Bloemen's algorithm in Isabelle is described in Section~\ref{sec:formalization}.
+The presentation of the correctness proof appears in Sections~\ref{sec:partial-correctness}, introducing the main invariants and the pre- and post-conditions used in the proof of partial correctness, and~\ref{sec:termination}, which explains the proof of termination.
+An assessment of the proof effort appears in Section~\ref{sec:effort}, and Section~\ref{sec:conclusion} contains a summary of the work and some perspectives for future work.
 
@@
 
+Informally, the algorithm terminates because (i) whenever function \prog{dfs} is called for arguments~$v$ and~$e$, the node~$v$ will be added to the set of \prog{visited} nodes (and therefore the complement of that set w.r.t.\ the finite set of vertices decreases) and (ii) every call of function $\prog{dfss} v e$ either terminates immediately or adds a new node $w$ to $\prog{vsuccs}~e~v$.
+Observe that this is true only if the calls respect the pre-conditions: for example, (i) does not hold if~$v$ was already visited at the call of \prog{dfs}.
+We now explain how we make this argument formal and prove termination of the algorithm, assuming the pre-conditions.
+
+As a first step, we define the following relation on argument tuples of the two functions.\footnote{Remember that \prog{Inl} and \prog{Inr} inject arguments of function \prog{dfs}, respectively \prog{dfss}, into the sum type representing the pair of functions.}
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+definition dfs-dfss-term where dfs-dfss-term ==
+  { (Inr (v,e1), Inl (v,e)) | v e e1.
+    v sin vertices - visited e 
+    /\ visited e1 = visited e cup {v} }
+cup { (Inl (w,e), Inr (v,e)) | v w e. v sin vertices }
+cup { (Inr (v,e''), Inr (v,e)) | v e e''.
+    v sin vertices /\ sub-env e e''
+    /\ (exists w sin vertices. w sin vsuccs e'' v - vsuccs e v) }
+\end{lstlisting}
+\end{small}
+
+This relation over-approximates the pairs of argument tuples for which recursive calls occur.
+For example, the first set in the above definition represents calls of \prog{dfss} from \prog{dfs} where $v$ was not visited at the call of \prog{dfs} but is added to the set of visited nodes at the call of \prog{dfss}.
+Observe that the second set, which corresponds to the call of \prog{dfs} from \prog{dfss}, requires that the call environment remains unchanged and in particular does not impose an immediate increase of either set \prog{visited} or \prog{vsuccs}.
+However, the use of different injection functions implies that one cannot construct a cycle using these argument tuples.
+
+Crucially, we prove that the relation \prog{dfs-dfss-term} is well-founded.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+lemma wf-term: wf dfs-dfss-term
+\end{lstlisting}
+\end{small}
+
+In the statement of the lemma, \prog{wf} represents the well-foundedness predicate, which is predefined in the Isabelle standard library.
+The lemma is proved by embedding the argument tuples into triples $(a,b,c)$ where $a$ is the set of unvisited vertices, $b$ is the set of edges between vertices that have not been followed, and $c$ is $0$ if the triple represents an argument of \prog{dfs} and $1$ if it is an argument of \prog{dfss}.
+These triples can be ordered by the strict subset order for the two first component and the predecessor relation on natural numbers for the third one, and Isabelle determines that the product of these orders is well-founded.
+Finally, \prog{dfs-dfss-term} can be embedded as an inverse image in this well-founded relation, and is therefore itself well-founded.
+
+We now prove that the previously defined pre-conditions imply termination of our two functions.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+theorem dfs-dfss-termination:
+  [| v sin vertices; pre-dfs v e |] 
+  ==> dfs-dfss-dom (Inl (v,e))
+  [| v sin vertices; pre-dfss v e |] 
+  ==> dfs-dfss-dom (Inr (v,e))
+\end{lstlisting}
+\end{small}
+
+The proof proceeds by well-founded induction along the relation \prog{dfs-dfss-term} and then shows that the arguments of recursive calls are indeed related.
+It makes use of the introduction rule for the predicate \prog{dfs-dfss-dom} that was generated by Isabelle for the definition of the two mutually recursive functions.
+
+Putting everything together, we can now prove our main theorem asserting the total correctness of the algorithm.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+theorem correctness:
+  assumes v sin vertices
+  shows   sccs (dfs v (init-env v)) = 
+          {S . is-scc S /\ (forall n sin S. reachable v n)}
+\end{lstlisting}
+\end{small}
+
+\section{Proof Effort}
+\label{sec:effort}
+
+\textbf{Summarize the overall effort of the proof in number of lines (definitions / proofs) for each main category.}
+
 \section{Conclusion}
 \label{sec:conclusion}
+
+\begin{itemize}
+\item successfully proved correctness of an efficient sequential SCC algorithm
+\item compare to Tarjan proof (Chen et al.): both are DFS algorithms, the algorithm considered here collapses loops as soon as they are found, minimizing stack size
+\item compare to Lammich's work: he exploits the refinement framework for deriving executable code, we work on a higher-level representation of the algorithm as elementary recursive functions, hence invariants are simpler
+\item objective: stepping stone towards a proof of the parallel version of the algorithm from Bloemen's thesis -- understand which invariants can be relaxed.
+\end{itemize}
 

### 2022-09-15 cpp/submission.tex
Added some examples in the introduction

@@
 
-They are also heavily used in critical domains that indirectly affect us, such as model checking.
-Indeed, as model checking involves exhaustive exploration of graphs of states, it is important to have efficient -- and correct -- algorithms to perform different operations such as the computation of strongly connected components.
+Graphs are also heavily used in domains to which we are indirectly linked, such as model checking and formal verification, which find applications in the safety of critical systems such as nuclear, aerospace, aeronautics or railways.
+
+Since model checking involves exhaustive exploration of graphs of states, it is important to have efficient -- and correct -- algorithms to perform different operations such as the computation of strongly connected components.
 

### 2022-09-19 cpp/submission.tex
a few changes here and there

@@
 
-\subsection{Isabelle/HOL}
+\subsection{Isabelle}
 \label{sec:isabelle}
 
-Isabelle~\cite{paulson:isabelle} is a logical framework in which different object logics can be encoded.
+Isabelle~\cite{paulson:isabelle} is a generic logical framework in which different object logics can be encoded.
 Isabelle/HOL~\cite{nipkow_isabellehol_2002} is the instance of Isabelle for higher-order logic, which we use in our work and which we will refer to in the following simply as Isabelle.
-It provides an expressive formal language for representing formal mathematical theories as well as algorithms and their correctness proofs, and it comes with an extensive set of tools for defining \emph{locales} (algebraic structures parameterized by operators and assumptions), (mutually) \emph{recursive functions} etc.
+It provides an expressive formal language for representing formal mathematical theories, including the formalization of algorithms and their correctness proofs, and it comes with an extensive set of tools for defining \emph{locales} (algebraic structures parameterized by operators and assumptions), \emph{inductive definitions} of sets and relations, (mutually) \emph{recursive functions} etc.
 
 Isabelle belongs to the LCF family of proof assistants whose soundness is ensured by a comparatively small trusted kernel that serves to certify proofs.
+In particular, theorems are represented by an inductive data type whose constructors correspond to the application of elementary axioms and rules provided by the definition of the object logic.
 From the user's point of view, Isabelle proofs are written in the Isar (``Intelligible Semi-Automated Reasoning'') language that is designed to make proofs readable and comprehensible for a mathematically inclined reader, with minimal overhead introduced by the formalism.
-Extensive automation is provided by built-in proof procedures based on rewriting and first-order reasoning, as well as the \emph{Sledgehammer}~\cite{blanchette:sledgehammer} add-on that selects potentially relevant lemmas from the context (including the standard proof library), calls external automatic proof backends such as SMT solvers as well as automatic proof tools for first-order and higher-order logic, and in case of success reconstructs their proofs in the trusted kernel.
+Extensive automation is provided by built-in proof procedures based on rewriting and first-order reasoning, as well as the \emph{Sledgehammer}~\cite{blanchette:sledgehammer} add-on that selects potentially relevant lemmas from the context (including the standard proof library), calls external automatic proof backends such as SMT solvers as well as automatic proof tools for first-order and higher-order logic, and in case of success calls back-end provers whose proofs can be reconstructed in the trusted kernel.
+
+Isabelle comes with an extensive standard library of definitions and theorems.
+Among the pre-defined types, we have the types $\alpha~\prog{set}$ of (possibly infinite) sets of elements of type $\alpha$ and $\alpha~\prog{list}$ of (finite) lists over $\alpha$.
+The latter is constructed by the empty list $[]$ and the \emph{cons} function, written $x\ \#\  xs$, that prepends an element $x$ to a list $xs$.
+The \prog{hd} and \prog{tl} functions access the head and the tail of a non-empty list.
+The concatenation of two lists $xs$ and $ys$ is denoted as $xs\ @\ ys$.
+In addition, we define the predicate
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+  x prec y in xs
+\end{lstlisting}
+\end{small}
+%
+that holds if the elements $x$ and $y$ both occur in the list $xs$, and $x$ has an occurrence to the left of an occurrence of $y$ in $xs$.
+For \prog{distinct} lists (i.e., lists without duplicate elements) $xs$, this predicate defines a partial order on the elements of $xs$.
 
@@
 
-We represent finite directed graphs with nodes of type $\nu$ as an Isabelle locale that fixes a finite set of vertices and a successor function associating to each node its set of successor nodes.
+We represent finite directed graphs with nodes of type $\nu$ as an Isabelle locale that fixes a finite set of vertices and a successor function associating to each node a set of successor nodes.
 
@@
 
-Elementary lemmas, such as transitivity of the reachability relation, are easily proved by induction, followed by an application of Isabelle's automatic proof methods.
+Elementary lemmas, such as the transitivity of the reachability relation, are easily proved by induction, followed by an application of Isabelle's automatic proof methods.
 For our later proofs, we also need the following variant of reachability between nodes avoiding a set $E$ of edges.
@@
 A maximal SCC is a non-empty partial SCC none of whose supersets is a partial SCC.
-In the following, we write SCC to denote maximal SCCs.
+In the following, when we write SCC, we refer to a maximal SCC.
 
@@
 
-The data on which the algorithm operates is collected in an ``environment'' record with the following components:
+We represent the data on which the algorithm operates as an ``environment'' record with the following components:
 
@@
 The function $\SS$ maps every node $v$ to the set of nodes that have already been determined to be part of the same SCC as $v$.
+%At the beginning of the execution, every node $v$ is mapped to the singleton component $\{v\}$.
 A core invariant of the algorithm will be that this mapping represents a partition of nodes into sets of equivalence classes: for all nodes $v$ and $w$, we maintain the relationship
@@
 The sets \prog{visited} and \prog{explored} represent the sets of nodes that have already been seen, respectively fully explored, by the algorithm, and \prog{sccs} is the set of (maximal) SCCs that the algorithm has found so far.
-Variable \prog{stack} holds the roots of the (incomplete) SCCs in depth-first order, whereas \prog{cstack} represents the stack of calls of the main function \prog{dfs} of the algorithm.
+Variable \prog{stack} holds the roots of the (partial) SCCs of visited, but not fully explored nodes, in depth-first order, whereas \prog{cstack} represents the call stack of the main function \prog{dfs} of the algorithm.
 Finally, \prog{vsuccs} remembers the set of outgoing edges (represented by the successor nodes) of each node that have already been followed during the exploration.
@@
 
-The algorithm will initially be called with an environment of the form $\prog{init-env}~v$ that holds $v$ as the root node, initializes $\SS$ by assigning the singleton $\{u\}$ to every node $u$, and has all other component initialized to the empty set or the empty list.
+The algorithm is initially called with an environment of the form $\prog{init-env}~v$ that holds $v$ as the root node, initializes $\SS$ by assigning the singleton $\{u\}$ to every node $u$, and has all other components initialized to the empty set or the empty list.
 
@@
 Function \prog{dfs} explores a node $v$ by adding it to the set of visited nodes, pushing it on both stacks, and then calling the function \prog{dfss}.
-If $v$ is still at the head of the stack, it is the root node of an SCC, and therefore $\SS~v$ is added to the set of maximal SCCs, all its nodes are marked as fully explored, and $v$ is popped from both stacks.
-Otherwise, $v$ is only popped from the call stack.
+If $v$ is still at the head of the stack when the call to \prog{dfss} finishes, the SCC to which $v$ belongs is added to the set of maximal SCCs, all its nodes are marked as fully explored, and $v$ is popped from both stacks.
+Otherwise, $v$ is not the root node of the SCC to which it belongs and is therefore no longer contained in the stack, but it is popped from the call stack.
 The tail-recursive function \prog{dfss} iterates over all successors $w$ of node $v$.
@@
 Since $w$ must be represented by some node on the stack, the remaining suffix is non-empty, and $\SS~(\prog{hd}~\prog{sfx})$ contains $w$.
-A new connected component $\prog{cc}$ is computed by taking the union of the sets $\SS~x$ for all nodes $x$ in the prefix and the head of the suffix.
+A new connected component $\prog{cc}$ is computed by taking the union of the partial SCCs represented by the nodes in the prefix, as well as the head of the suffix.
 The mapping $\SS$ is updated so that all nodes in $\prog{cc}$ are mapped to $\prog{cc}$, and the stack is shortened to the suffix of the previous stack, implicitly making the head of that suffix the root of the component $cc$.
@@
 A technical complication in the definition of the algorithm is that the functions \prog{dfs} and \prog{dfss} need not terminate when their pre-conditions are violated.
-We will come back to proving termination in Sect.~\ref{sec:termination}.
+We will come back to proving termination in Section~\ref{sec:termination}.
 
@@
 
-We define well-formed environments to be those satisfying the following conditions, and we will prove this predicate to hold throughout the execution of the algorithm.
+We define \emph{well-formed} environments to be those satisfying the following conditions, and we will prove this predicate to hold throughout the execution of the algorithm.
 
@@
 
-In words, every visited node is reachable from the root node, the two stacks do not contain duplicate elements, and the stack is a subsequence of the call stack.
+In words, every visited node is reachable from the root node, the two stacks contain every node at most once, and the stack of SCC roots is a subsequence of the call stack.
 Nodes on the stack are reachable from all stack nodes below them.
@@
 All nodes reachable from fully explored nodes are themselves fully explored.
-The visited successors of a node are both successors and visited, and no outgoing edge of an unvisited node has been followed.
+The visited successors of a node are both successors and visited, and an unvisited node has no visited successors.
 All outgoing edges of fully explored nodes, as well as of visited nodes that are no longer on the call stack, have been followed.
-The mapping $\SS$ represents an equivalence relation, as discussed above, and $\SS~n$ is the singleton $\{n\}$ for an unvisited node $n$.
-The nodes on the stack are roots of (non-maximal) SCCs, hence their equivalence classes are disjoint.
-A node has been visited but not yet fully explored iff its root node is on the stack.
+The mapping $\SS$ represents an equivalence relation, as discussed above, and it maps an unvisited node $n$ to the singleton $\{n\}$.
+The nodes on the stack are roots of (partial) SCCs, hence their equivalence classes are disjoint.
+A node has been visited but not yet fully explored iff the root node of its SCC is on the stack.
 Moreover, root nodes of SCCs are the oldest nodes in depth-first order.
-No node $u$ belonging to the SCC of a node $n$ on the stack can reach a node $m$ strictly below $n$ on the stack without going through some edge from some node in $n$'s equivalence class that has not yet been followed.
+No node $u$ belonging to the SCC of a node $n$ on the stack can reach a node $m$ strictly below $n$ on the stack without going through an edge from some node in $n$'s equivalence class that has not yet been followed.
 In other words, the current SCCs are maximal with respect to the current knowledge of the algorithm.
-The set of not yet followed edges for nodes in the equivalence class of $n$ are defined as
+The set of not yet followed edges for nodes in the equivalence class of $n$ is defined as
 %
@@
 %
-Finally, every equivalence class is a (not necessarily maximal) SCC, elements of \prog{sccs} are maximal SCCs, and their union is the set of explored nodes.
+Finally, every equivalence class is a (partial) SCC, elements of \prog{sccs} are maximal SCCs, and the union of their nodes is the set of explored nodes.
 
@@
 We prove a number of simple consequences of this predicate.
-For example, $n \in \SS~n$ holds for every node~$n$.
+For example, $n \in \SS~e~n$ holds for every node~$n$ if $e$ is a well-formed environment.
 If node $w$ is reachable from a visited node $v$, but not through any edge that has not yet been followed, then $w$ must be visited.
-Also, edges towards fully explored nodes do not contribute to reachability of unexplored nodes avoiding certain edges:
+Also, edges towards fully explored nodes do not contribute to reachability of unexplored nodes avoiding some set of edges:
 %
@@
 
-\subsection{Pre- and post-conditions of functions \prog{dfs} and \prog{dfss}}
+\subsection{Pre- and post-conditions of the functions}
 \label{sec:pre-post}
 
-The proof of correctness relies on the fact that a call of \prog{dfs} for arguments that satisfy some hypotheses (the pre-conditions) terminates and satisfies some properties (the post-conditions).
-Because of the mutual recursion, we must define pre- and post-conditions for both \prog{dfs} and \prog{dfss}.
+The proof of partial correctness of the algorithm relies on the fact that the environment produced by a call of \prog{dfs} for arguments that satisfy certain hypotheses (the pre-conditions) satisfies certain other properties (the post-conditions).
+Because of the mutual recursion between the two functions, we must define pre- and post-conditions for both \prog{dfs} and \prog{dfss}.
 
@@
 
-Function \prog{dfs} maintains the invariant \prog{wf-env} and returns an environment $e'$ that extends the input environment $e$.
+Function \prog{dfs} produces a well-formed environment $e'$ that extends the input environment $e$.
 Node $v$ has been visited and all its outgoing edges have been followed.
-Because the algorithm works in depth-first fashion, no new outgoing edges of nodes that had already been visited in the input environment have been followed, and the stack of $e'$ is a suffix of the stack of $e$.
+Because the algorithm works in depth-first fashion, no new outgoing edges of nodes that had already been visited in the input environment have been followed, and the stack of SCC roots in $e'$ is a suffix of the corresponding stack of $e$.
 % Again, this follows from the pre-condition and the suffix property.
 % such that $v$ is still reachable from all nodes on the stack. 
-The stack may have been shortened because SCCs represented at the top of the stack may have been merged.
+In particular, the stack may have been shortened because SCCs represented at the top of the stack may have been merged.
 The call stack is reestablished as it was in $e$.
@@
 \item Alternatively, the stack of $e'$ must be non-empty and $v$ must be a member of the partial SCC of the node at the top of the stack.
-The SCCs represented by the stack elements other than the top are unchanged.
-This corresponds to the case where $v$ is not the root node of its SCC, and some SCCs at the top of the stack may have been merged.
+The partial SCCs represented by the stack elements other than the top are unchanged.
+This corresponds to the case where $v$ is not the root node of its SCC, and some partial SCCs at the top of the stack may have been merged.
 \end{itemize}
@@
 
-In the same way as function \prog{dfs}, function \prog{dfss} ensures that the resulting environment $e'$ is well-formed and extends the input environment $e$, that all outgoing edges of $v$, and no outgoing edges of previously visited nodes other than $v$ have been followed.
-Again, the new stack is a suffix of the old one, and the call stack is restored.
-In case node $v$ is still on the stack (and therefore is the root node of its SCC), no stack node below $v$ can be reached from $v$.
-This condition is important for ensuring that the computed SCC is maximal.
+In the same way as function \prog{dfs}, function \prog{dfss} ensures that the resulting environment $e'$ is well-formed and extends the input environment $e$, that all outgoing edges of $v$ have been followed, and that the sets of visited successors of previously visited nodes other than $v$ are unchanged.
+The new stack is non-empty, it is a suffix of the old stack, and the call stack is restored.
+The partial SCCs corresponding to nodes in the stack other than the top stack element are left untouched.
+In case node $v$ is still on the stack (and is therefore the root node of its partial SCC), no stack node below $v$ can be reached from $v$.
+This condition is important for ensuring that the computed SCC is in fact maximal.
 
@@
 For example, the following lemma asserts that the pre-condition of \prog{dfss} holds at the call from the body of \prog{dfs}.
-Although some of the conditions of the invariant for the augmented environment require a few lines of proof, the overall proof is straightforward.
+Although some of the well-formedness conditions for the augmented environment require a few lines of proof, the overall proof is straightforward.
 
@@
 lemma pre-dfss-pre-dfs:
-  assumes pre-dfss v e and w sin successors v - visited e
+  assumes pre-dfss v e
+  and     w sin successors v - visited e
   shows   pre-dfs w e
@@
 Some steps in the proof of this lemma require a substantial number of interactions.
-We started by proving several auxiliary lemmas about function \prog{unite}, including the fact that the intermediate environment $e'$ is well-formed.
+We start by proving several auxiliary lemmas about function \prog{unite}, including the fact that the intermediate environment $e'$ is well-formed.
 Then, we have to show that this invariant also holds for $e''$.
@@
 
-The following theorem is central for the proof of partial correctness: it shows that the functions \prog{dfs} and \prog{dfss} establish their respective post-conditions, assuming that the pre-conditions hold for the input parameters.
-Because we have not proved termination of the functions, we need an additional hypothesis for each statement that, intuitively, asserts that the corresponding function will terminate for its input parameters.
+The following lemma is central for the proof of partial correctness: it shows that the functions \prog{dfs} and \prog{dfss} establish their respective post-conditions, assuming that the pre-conditions hold for the input parameters.
+Because we have not yet proved termination of the functions, we need an additional hypothesis for each statement that, intuitively, asserts that the corresponding function will terminate for its input parameters.
 
@@
 
-Since the two functions are mutually recursive, the two statements must be proved by simultaneous computational induction, based on the corresponding proof rule generated by Isabelle.
+Since the two functions are mutually recursive, the two statements must be proved by simultaneous computational induction, using the corresponding proof rule generated by Isabelle.
 The proofs of these implications again involve a certain number of steps requiring detailed interaction.
@@
 
-In words, assuming that function \prog{dfs} terminates for the input parameters, it computes the set of the maximal SCCs in the sub-graph reachable from node~$v$.
-In particular, if $v$ is a root node of the graph, the maximal SCCs of the graph will be computed.
+In words, assuming that function \prog{dfs} terminates for the input parameters, it computes the set of all maximal SCCs in the sub-graph reachable from node~$v$.
+In particular, if $v$ is a root node of the graph, all maximal SCCs of the graph will be computed.
 
@@
 
-Informally, the algorithm terminates because (i) whenever function \prog{dfs} is called for arguments~$v$ and~$e$, the node~$v$ will be added to the set of \prog{visited} nodes (and therefore the complement of that set w.r.t.\ the finite set of vertices decreases) and (ii) every call of function $\prog{dfss} v e$ either terminates immediately or adds a new node $w$ to $\prog{vsuccs}~e~v$.
-Observe that this is true only if the calls respect the pre-conditions: for example, (i) does not hold if~$v$ was already visited at the call of \prog{dfs}.
+Informally, the algorithm terminates because (i) whenever function \prog{dfs} is called for arguments~$v$ and~$e$, the node~$v$ will be added to the set of \prog{visited} nodes (and therefore the complement of that set w.r.t.\ the finite set of vertices decreases) and (ii) every call of function $\prog{dfss}~v~e$ either terminates immediately or adds a new node $w$ to $\prog{vsuccs}~e~v$.
+Observe that these conditions can be established only if the calls respect the pre-conditions: for example, (i) does not hold if~$v$ was already visited at the call of \prog{dfs}.
 We now explain how we make this argument formal and prove termination of the algorithm, assuming the pre-conditions.
 
-As a first step, we define the following relation on argument tuples of the two functions.\footnote{Remember that \prog{Inl} and \prog{Inr} inject arguments of function \prog{dfs}, respectively \prog{dfss}, into the sum type representing the pair of functions.}
+As a first step, we define the following relation on argument tuples of the two functions.\footnote{Remember that \prog{Inl} and \prog{Inr} inject arguments of function \prog{dfs}, respectively \prog{dfss}, into the sum type representing arguments of the pair of functions.}
 
@@
 The lemma is proved by embedding the argument tuples into triples $(a,b,c)$ where $a$ is the set of unvisited vertices, $b$ is the set of edges between vertices that have not been followed, and $c$ is $0$ if the triple represents an argument of \prog{dfs} and $1$ if it is an argument of \prog{dfss}.
-These triples can be ordered by the strict subset order for the two first component and the predecessor relation on natural numbers for the third one, and Isabelle determines that the product of these orders is well-founded.
+These triples can be ordered by the strict subset order for the two first component and the predecessor relation on natural numbers for the third one, and Isabelle proves automatically that the product of these orders is well-founded.
 Finally, \prog{dfs-dfss-term} can be embedded as an inverse image in this well-founded relation, and is therefore itself well-founded.
 
-We now prove that the previously defined pre-conditions imply termination of our two functions.
+We now show that the previously defined pre-conditions imply termination of our two functions.
 
@@
 The proof proceeds by well-founded induction along the relation \prog{dfs-dfss-term} and then shows that the arguments of recursive calls are indeed related.
-It makes use of the introduction rule for the predicate \prog{dfs-dfss-dom} that was generated by Isabelle for the definition of the two mutually recursive functions.
+It makes use of the introduction rule for the predicate \prog{dfs-dfss-dom} that was generated by Isabelle when the two mutually recursive functions were defined.
 
@@
 
-\section{Proof Effort}
+\section{Discussion of the Proof}
+\label{sec:discussion}
+
+\subsection{Proof effort}
 \label{sec:effort}
@@
 \textbf{Summarize the overall effort of the proof in number of lines (definitions / proofs) for each main category.}
+
+\subsection{Related work}
+\label{sec:related}
+
+We are not the first who prove the correctness of an algorithm for computing the SCCs of a graph using a proof assistant.
+Peter Lammich~\cite{lammich:gabow} used Isabelle/HOL for proving the correctness of Gabow's SCC algorithm.
+Like Bloemen's algorithm, Gabow applies depth-first search and augments partial SCCs whenever a back loop to an already visited vertex of the graph is found.
+Lammich's proof focuses on efficient code generation, based on the use of his refinement framework~\cite{lammich:refinement} formalized in Isabelle.
+In this framework, algorithms are represented as monadic functions, and concrete data structures are introduced, for example for representing partial SCCs.
+Our proof is based on a more abstract representation of the algorithm and its data structures, which leads to higher-level invariants and reduces proof effort.
+Indeed, our main objective is not to obtain an efficient sequential implementation of Bloemen's algorithm but rather to pave the way for a verification of the parallel version of that algorithm and to understand how our invariants can be relaxed while maintaining correctness.
+
+Chen et al.~\cite{chen:tarjan} present a comparison of the proofs of Tarjan's algorithm for computing the SCCs of a graph in the proof assistants Coq, Isabelle, and Why3.
+The algorithm is represented at a similar level of abstraction as in our proof, leading to similar proof effort.
+However, although that algorithm is also based on depth-first search, its algorithmis structure is quite different because no partial SCCs are computed along the way, and a maximal SCC is popped from the stack only when a node is determined to be an SCC root at the end of the function call.
 

### 2022-09-20 cpp/submission.tex
small corrections to paper and a simplification in Isabelle

@@
 Peter Lammich~\cite{lammich:gabow} used Isabelle/HOL for proving the correctness of Gabow's SCC algorithm.
-Like Bloemen's algorithm, Gabow applies depth-first search and augments partial SCCs whenever a back loop to an already visited vertex of the graph is found.
+Like Bloemen's algorithm, Gabow's algorithm applies depth-first search and augments partial SCCs whenever a back loop to an already visited vertex of the graph is found.
 Lammich's proof focuses on efficient code generation, based on the use of his refinement framework~\cite{lammich:refinement} formalized in Isabelle.
@@
 Our proof is based on a more abstract representation of the algorithm and its data structures, which leads to higher-level invariants and reduces proof effort.
-Indeed, our main objective is not to obtain an efficient sequential implementation of Bloemen's algorithm but rather to pave the way for a verification of the parallel version of that algorithm and to understand how our invariants can be relaxed while maintaining correctness.
+Indeed, our main objective is not to obtain an efficient sequential implementation of Bloemen's algorithm but rather to pave the way for a verification of the parallel version of that algorithm.
 
@@
 The algorithm is represented at a similar level of abstraction as in our proof, leading to similar proof effort.
-However, although that algorithm is also based on depth-first search, its algorithmis structure is quite different because no partial SCCs are computed along the way, and a maximal SCC is popped from the stack only when a node is determined to be an SCC root at the end of the function call.
+However, although Tarjan's algorithm is also based on depth-first search, its algorithmic structure is quite different because no partial SCCs are computed along the way, and a maximal SCC is popped from the stack only when a node is determined to be an SCC root at the end of the function call.
+The primary focus of the presentation in~\cite{chen:tarjan} is on a comparison between the styles of proof and the overall effort required in the three proof assistants.
 

### 2022-09-20 cpp/submission.tex
Summary of proof effort in a table

@@
 \renewcommand{\SS}{\mathcal{S}}
+\usepackage{array}
+\newcolumntype{M}[1]{>{\centering\arraybackslash}m{#1}}
 
@@
 The lemma is proved by embedding the argument tuples into triples $(a,b,c)$ where $a$ is the set of unvisited vertices, $b$ is the set of edges between vertices that have not been followed, and $c$ is $0$ if the triple represents an argument of \prog{dfs} and $1$ if it is an argument of \prog{dfss}.
-These triples can be ordered by the strict subset order for the two first component and the predecessor relation on natural numbers for the third one, and Isabelle proves automatically that the product of these orders is well-founded.
+These triples can be ordered by the strict subset order for the two first components and the predecessor relation on natural numbers for the third one, and Isabelle proves automatically that the product of these orders is well-founded.
 Finally, \prog{dfs-dfss-term} can be embedded as an inverse image in this well-founded relation, and is therefore itself well-founded.
@@
 
-\textbf{Summarize the overall effort of the proof in number of lines (definitions / proofs) for each main category.}
+We summarize the overall effort of the proof in number of lines of the Isabelle theory (definitions / proofs) in a table for each main category.
+
+\begin{center}
+  \small
+  \begin{tabular}{|M{.125\textwidth}|M{.075\textwidth}|m{.21\textwidth}|}
+    \hline
+    Category & Lines & Description\\
+    \hline
+    Lists ordering  & 102-240 & We use a relation of precedence in lists that determines a partial order on the elements that appear in a list, provided those elements have unique occurrences.
+In particular, stacks are represented as lists.\\
+    \hline
+    Graphs & 251-424 & We prove a few lemmas on basic concepts in graphs. In particular, we define reachability and SCCs and prove some auxiliary lemmas on them.\\
+    \hline
+    SCC algorithm & 444-492 & We formalize the algorithm by defining functions \prog{unite}, \prog{dfs} and \prog{dfss}.\\
+    \hline
+    Invariants & 500-886 & We formalize the main invariants, i.e.
+\prog{wf-env} and the pre- and post-conditions for both functions \prog{dfs} and \prog{dfss} and prove some direct consequences of these invariants.\\
+    \hline
+    Partial correctness & 897-3087 & We prove partial correctness by proving that the pre-conditions imply the post-conditions.\\
+    \hline
+    Termination & 3140-3413 & We prove termination using the fact that at each call, either a new vertex is visited or a new successor is added to the set of visited successors of some vertex.\\
+    \hline
+    Total correctness & 3419-3423 & We prove total correctness using termination and partial correctness.\\
+    \hline
+  \end{tabular}
+\end{center}
 

### 2022-09-21 cpp/submission.tex
Added some lines in the conclusion

@@
 
-We summarize the overall effort of the proof in number of lines of the Isabelle theory (definitions / proofs) in a table for each main category.
+We summarize the overall effort of the proof in number of lines (definitions / proofs) for each main category.
 
@@
 
-\begin{itemize}
-\item successfully proved correctness of an efficient sequential SCC algorithm
-\item compare to Tarjan proof (Chen et al.): both are DFS algorithms, the algorithm considered here collapses loops as soon as they are found, minimizing stack size
-\item compare to Lammich's work: he exploits the refinement framework for deriving executable code, we work on a higher-level representation of the algorithm as elementary recursive functions, hence invariants are simpler
-\item objective: stepping stone towards a proof of the parallel version of the algorithm from Bloemen's thesis -- understand which invariants can be relaxed.
-\end{itemize}
+As a result of this whole year of work, we successfully proved correctness of an efficient sequential set-based SCC algorithm.
+In comparison to Tarjan's algorithm whose formal proof was written by Chen et.
+al \cite{chen:tarjan}, this algorithm collapses loops as soon as they are found, thus minimizing the stack size.
+Our formalization and  proof in Isabelle are quite similar to theirs as they are based on higher-level representations of the algorithms as elementary recursive functions.
+Hence, our invariants are quite simple compared to those expressed in Lammich's work \cite{lammich:gabow} since he exploits the refinement framework for deriving executable code, resulting in implementable-like representations that make expressing invariants more difficult from a higher-level point of view.
+
+% \begin{itemize}
+% \item successfully proved correctness of an efficient sequential SCC algorithm
+% \item compare to Tarjan proof (Chen et al.): both are DFS algorithms, the algorithm considered here collapses loops as soon as they are found, minimizing stack size
+% \item compare to Lammich's work: he exploits the refinement framework for deriving executable code, we work on a higher-level representation of the algorithm as elementary recursive functions, hence invariants are simpler
+% \item objective: stepping stone towards a proof of the parallel version of the algorithm from Bloemen's thesis -- understand which invariants can be relaxed.
+% \end{itemize}
 

### 2022-09-21 cpp/submission.tex
Added some lines in the conclusion
almost final version of CPP submission

--- 2022-09-20 cpp/submission.tex
@@
 \label{sec:introduction}
-Directed graphs are widely used in many fields of computer science and mathematics.
-The study of their properties is therefore an important topic because of their application in areas that may affect our lives.
 
-Graphs are also heavily used in domains to which we are indirectly linked, such as model checking and formal verification, which find applications in the safety of critical systems such as nuclear, aerospace, aeronautics or railways.
+Directed graphs are widely used in many fields of computer science and mathematics, and they are the source of many interesting algorithmic problems.
+A fundamental problem is the computation of strongly connected components (SCCs) of a graph, that is, maximal sets~$S$ of vertices such that all vertices in~$S$ are reachable from every other vertex in~$S$ \cite[Chap.~4.2]{sedgewick:algorithms}.
+Among many other applications, computing SCCs underlies algorithmic techniques for system verification, such as model checking, that are applied to ensure the correctness of safety-critical systems in domains including nuclear energy, aerospace, aeronautics, medical applications, and the railway industry.
+An error in an algorithm for computing SCCs or in its implementation could therefore jeopardize the verdict produced by a model checker.
+Recognizing the importance of this problem, a seminal result about the formal verification of model checkers was the work by Esparza et al.~\cite{esparza:cava} on the formal proof of an executable model checker for LTL, later extended to also include techniques for partial order reduction~\cite{brunner:partial-order}.
 
-Since model checking involves exhaustive exploration of graphs of states, it is important to have efficient -- and correct -- algorithms to perform different operations such as the computation of strongly connected components.
+Several different algorithms exist for efficiently computing SCCs of a graph.
+Perhaps the most famous one is Tarjan's algorithm~\cite{tarjan:depth-first}.
+It performs a depth-first traversal of the nodes of the graph.
+When it encounters an edge that leads to an already visited node that is part of the current DFS path, it detects a cycle, and it tracks (via additional information stored for the encountered nodes) which nodes belong to the same SCC.
+Dijkstra~\cite{dijkstra:finding} suggested a variant of a DFS traversal where the stack contains roots of (partial) SCCs.
+When a cycle is detected, the corresponding partial SCCs are collapsed, shortening the stack.
+Both algorithms are efficient because they operate in time $O(|V|+|E|)$ where $V$ and $E$ denote the sets of nodes and edges of the graph, since every node and every edge is visited only once during the traversal, but they differ in their algorithmic structure and in the auxiliary information that must be stored.
+
+However, both these algorithms and their numerous variants are fundamentally sequential because their correctness crucially relies on the graph being traversed in DFS order.
+Bloemen et al.~\cite{bloemen:strong,bloemen:multi-core} introduced a variant of Dijkstra's algorithm for computing SCCs that can be implemented on a shared-memory multi-core processor, based on efficient data structures for (i) implementing a union-find structure representing partial SCCs and (ii) a cyclic list for iterating over states, that minimize locking.
+This algorithm has been implemented in the high-performance model checker LTSmin.
+The correctness proofs in Bloemen's thesis are based on ordinary paper-and-pencil reasoning and therefore do not correspond to the same exacting standards as the machine-checked verification of model checking algorithms mentioned above.
+The problem is particularly delicate given that a multi-core implementation of the algorithm need not strictly adhere to a DFS traversal order, and any core may revisit nodes that have already been visited by other cores.
+
+The present work makes a first contribution towards a machine-checked proof of the SCC algorithm by Bloemen et al.\ by establishing the correctness of the sequential version of the algorithm.
+Although this is arguably less challenging than the proof of the parallel version, we consider it to be a necessary first step that establishes a baseline against which the parallel algorithm can be compared.
+In particular, we exhibit inductive invariants that ensure the correctness of the sequential algorithm, and it then becomes meaningful to consider which invariants can be relaxed in a parallel execution.
+We are interested in the correctness of the high-level algorithm and therefore do not consider the concrete representation of the data structures on which it relies, or the generation of efficiently executable code implementing the sequential algorithm.
 
@@
 Our encoding of Bloemen's algorithm in Isabelle is described in Section~\ref{sec:formalization}.
-The presentation of the correctness proof appears in Sections~\ref{sec:partial-correctness}, introducing the main invariants and the pre- and post-conditions used in the proof of partial correctness, and~\ref{sec:termination}, which explains the proof of termination.
-An assessment of the proof effort appears in Section~\ref{sec:effort}, and Section~\ref{sec:conclusion} contains a summary of the work and some perspectives for future work.
+The presentation of the correctness proof appears in Sections~\ref{sec:partial-correctness}, introducing the main invariants and the pre- and post-conditions used in the proof of partial correctness, whereas Section~\ref{sec:termination} explains the proof of termination.
+An assessment of the proof effort appears in Section~\ref{sec:effort}, related work is discussed in Section~\ref{sec:related}, and Section~\ref{sec:conclusion} contains a summary of the work and some perspectives for future work.
 
@@
 Isabelle~\cite{paulson:isabelle} is a generic logical framework in which different object logics can be encoded.
-Isabelle/HOL~\cite{nipkow_isabellehol_2002} is the instance of Isabelle for higher-order logic, which we use in our work and which we will refer to in the following simply as Isabelle.
-It provides an expressive formal language for representing formal mathematical theories, including the formalization of algorithms and their correctness proofs, and it comes with an extensive set of tools for defining \emph{locales} (algebraic structures parameterized by operators and assumptions), \emph{inductive definitions} of sets and relations, (mutually) \emph{recursive functions} etc.
+Isabelle/HOL~\cite{nipkow:isabelle} is the instance of Isabelle for higher-order logic, which we use in our work and which we will refer to in the following simply as Isabelle.
+It provides an expressive formal language for representing formal mathematical theories, including the formalization of algorithms and their correctness proofs, and it comes with an extensive standard library of operator definitions and lemmas, and with tools for defining \emph{locales} (algebraic structures parameterized by operators and assumptions), \emph{inductive definitions} of sets and relations, (mutually) \emph{recursive functions} etc.
 
 Isabelle belongs to the LCF family of proof assistants whose soundness is ensured by a comparatively small trusted kernel that serves to certify proofs.
-In particular, theorems are represented by an inductive data type whose constructors correspond to the application of elementary axioms and rules provided by the definition of the object logic.
+In particular, theorems are represented by an inductive data type whose constructors correspond to the instantiation of elementary axioms and the application of rules, typically provided by the definition of the object logic.
 From the user's point of view, Isabelle proofs are written in the Isar (``Intelligible Semi-Automated Reasoning'') language that is designed to make proofs readable and comprehensible for a mathematically inclined reader, with minimal overhead introduced by the formalism.
-Extensive automation is provided by built-in proof procedures based on rewriting and first-order reasoning, as well as the \emph{Sledgehammer}~\cite{blanchette:sledgehammer} add-on that selects potentially relevant lemmas from the context (including the standard proof library), calls external automatic proof backends such as SMT solvers as well as automatic proof tools for first-order and higher-order logic, and in case of success calls back-end provers whose proofs can be reconstructed in the trusted kernel.
+Extensive automation is provided by built-in proof procedures based on rewriting and first-order reasoning, as well as by the \emph{Sledgehammer}~\cite{blanchette:sledgehammer} add-on that selects potentially relevant lemmas from the context (including the standard proof library), calls external automatic proof backends such as SMT solvers or automatic proof tools for first-order and higher-order logic, and in case of success applies back-end provers whose proofs can be reconstructed in the trusted kernel.
 
-Isabelle comes with an extensive standard library of definitions and theorems.
-Among the pre-defined types, we have the types $\alpha~\prog{set}$ of (possibly infinite) sets of elements of type $\alpha$ and $\alpha~\prog{list}$ of (finite) lists over $\alpha$.
+As part of the standard library, we have the pre-defined types $\alpha~\prog{set}$ of (possibly infinite) sets of elements of type~$\alpha$, and $\alpha~\prog{list}$ of (finite) lists over~$\alpha$.
 The latter is constructed by the empty list $[]$ and the \emph{cons} function, written $x\ \#\  xs$, that prepends an element $x$ to a list $xs$.
@@
 that holds if the elements $x$ and $y$ both occur in the list $xs$, and $x$ has an occurrence to the left of an occurrence of $y$ in $xs$.
-For \prog{distinct} lists (i.e., lists without duplicate elements) $xs$, this predicate defines a partial order on the elements of $xs$.
+For a \prog{distinct} list~$xs$ (i.e., a list without duplicate elements), this predicate defines a partial order on the elements of~$xs$.
 
@@
 Again, this relation is transitive, it implies reachability, is anti-monotonic in the set $E$, and reachability agrees with reachability avoiding the empty set of edges.
-More interestingly, we show the following lemma about adding an edge to the set $E$.
+More interestingly, we show the following lemma about adding an edge to the set~$E$ of forbidden edges.
 
@@
 
+Strongly connected components are of particular importance for stating the correctness of our algorithm.
 A partial SCC is a set $S$ of nodes all of which are reachable from each other.
 In particular, a cycle in a graph constitutes a partial SCC.
-A maximal SCC is a non-empty partial SCC none of whose supersets is a partial SCC.
+A maximal SCC is a non-empty partial SCC such that none of its supersets is a partial SCC.
 In the following, when we write SCC, we refer to a maximal SCC.
@@
 The \prog{root} component stores the node for which the algorithm was originally called.
-The function $\SS$ maps every node $v$ to the set of nodes that have already been determined to be part of the same SCC as $v$.
+The function $\SS$ maps every node~$v$ to the set of nodes that have already been determined to be part of the same SCC as~$v$.
 %At the beginning of the execution, every node $v$ is mapped to the singleton component $\{v\}$.
-A core invariant of the algorithm will be that this mapping represents a partition of nodes into sets of equivalence classes: for all nodes $v$ and $w$, we maintain the relationship
+A core invariant of the algorithm will be that this mapping represents a partition of nodes into sets of equivalence classes: for all nodes~$v$ and~$w$, we maintain the relationship
 \[
@@
 \]
-The sets \prog{visited} and \prog{explored} represent the sets of nodes that have already been seen, respectively fully explored, by the algorithm, and \prog{sccs} is the set of (maximal) SCCs that the algorithm has found so far.
+The sets \prog{visited} and \prog{explored} represent the sets of nodes that have already been seen, respectively fully explored, by the algorithm, and \prog{sccs} is the set of (maximal) SCCs that the algorithm has detected so far.
 Variable \prog{stack} holds the roots of the (partial) SCCs of visited, but not fully explored nodes, in depth-first order, whereas \prog{cstack} represents the call stack of the main function \prog{dfs} of the algorithm.
-Finally, \prog{vsuccs} remembers the set of outgoing edges (represented by the successor nodes) of each node that have already been followed during the exploration.
-Note that the variables \prog{root} and \prog{cstack} are not actually used by the algorithm: they are ghost variables used in the correctness proof.
+Finally, \prog{vsuccs} remembers the set of outgoing edges (represented by the target nodes) of each node that have already been followed during the exploration.
+Let us observe that the variables \prog{root} and \prog{cstack} are not actually accessed by the algorithm: they are ghost variables used in the correctness proof.
 
@@
 Function \prog{dfs} explores a node $v$ by adding it to the set of visited nodes, pushing it on both stacks, and then calling the function \prog{dfss}.
-If $v$ is still at the head of the stack when the call to \prog{dfss} finishes, the SCC to which $v$ belongs is added to the set of maximal SCCs, all its nodes are marked as fully explored, and $v$ is popped from both stacks.
+If $v$ is still at the head of the stack when the call to \prog{dfss} returns, the SCC to which $v$ belongs is added to the set of maximal SCCs, all its nodes are marked as fully explored, and $v$ is popped from both stacks.
 Otherwise, $v$ is not the root node of the SCC to which it belongs and is therefore no longer contained in the stack, but it is popped from the call stack.
@@
 If $w$ has already been fully explored, there is nothing to do.
-If $w$ is unvisited, the function \prog{dfs} is called for $w$, otherwise the algorithm detected a loop back to a node that has already been visited and whose root node $r$ must therefore be on the stack.
-It unites the SCCs of all nodes contained in the prefix\footnote{Note that the top of the stack is at the head of the list.} of the stack starting at $r$, using the following auxiliary function.
+If $w$ has not been seen before, the function \prog{dfs} is called for $w$, otherwise the algorithm detects a loop back to a node that has already been visited.
+Therefore, the root node~$r$ of the SCC to which~$w$ belongs must be on the stack.
+The partial SCCs of all nodes contained in the prefix\footnote{Note that the top of the stack is at the head of the list.} of the stack up to node~$r$ are collapsed, using the following auxiliary function.
 
@@
 
-In words, \prog{unite} computes the prefix of the stack consisting of the roots of SCCs that do not contain $w$.
+In words, \prog{unite} computes the prefix of the stack consisting of the roots of partial SCCs that do not contain $w$.
 Since $w$ must be represented by some node on the stack, the remaining suffix is non-empty, and $\SS~(\prog{hd}~\prog{sfx})$ contains $w$.
@@
 Auto-generated theorems, such as computational induction, simultaneously handle both functions in order to enable mutual induction.
-A technical complication in the definition of the algorithm is that the functions \prog{dfs} and \prog{dfss} need not terminate when their pre-conditions are violated.
+A technical complication in the definition of the algorithm is that the functions \prog{dfs} and \prog{dfss} need not terminate when their pre-conditions (defined below) are violated.
 We will come back to proving termination in Section~\ref{sec:termination}.
@@
 Nodes on the stack are reachable from all stack nodes below them.
-All explored nodes, as well as all nodes on the call stack, are visited.
+All explored nodes, as well as all nodes on the call stack (hence also the nodes on the stack of SCC roots), are visited.
 All nodes reachable from fully explored nodes are themselves fully explored.
@@
 The mapping $\SS$ represents an equivalence relation, as discussed above, and it maps an unvisited node $n$ to the singleton $\{n\}$.
-The nodes on the stack are roots of (partial) SCCs, hence their equivalence classes are disjoint.
-A node has been visited but not yet fully explored iff the root node of its SCC is on the stack.
-Moreover, root nodes of SCCs are the oldest nodes in depth-first order.
+The nodes on the stack are roots of partial SCCs, hence their equivalence classes are disjoint.
+A node has been visited but not yet fully explored iff the root node of the associated partial SCC is on the stack.
+Moreover, root nodes of partial SCCs are the oldest nodes in depth-first order.
 No node $u$ belonging to the SCC of a node $n$ on the stack can reach a node $m$ strictly below $n$ on the stack without going through an edge from some node in $n$'s equivalence class that has not yet been followed.
-In other words, the current SCCs are maximal with respect to the current knowledge of the algorithm.
-The set of not yet followed edges for nodes in the equivalence class of $n$ is defined as
+In other words, the partial SCCs are maximal with respect to the current knowledge of the algorithm.
+Given a node~$n$, the set of edges that leave some node in the equivalence class of~$n$ and have not yet been followed is defined as
 %
@@
 We prove a number of simple consequences of this predicate.
-For example, $n \in \SS~e~n$ holds for every node~$n$ if $e$ is a well-formed environment.
+For example, if $e$ is a well-formed environment then $n \in \SS~e~n$ holds for every node~$n$.
 If node $w$ is reachable from a visited node $v$, but not through any edge that has not yet been followed, then $w$ must be visited.
-Also, edges towards fully explored nodes do not contribute to reachability of unexplored nodes avoiding some set of edges:
+Also, edges towards fully explored nodes do not contribute to reachability of not fully explored nodes avoiding some set of edges:
 %
@@
 Node $v$ has been visited and all its outgoing edges have been followed.
-Because the algorithm works in depth-first fashion, no new outgoing edges of nodes that had already been visited in the input environment have been followed, and the stack of SCC roots in $e'$ is a suffix of the corresponding stack of $e$.
+Because the algorithm works in depth-first fashion, no new outgoing edges of nodes that were already visited in the input environment have been followed, and the stack of SCC roots in $e'$ is a suffix of the corresponding stack of $e$.
 % Again, this follows from the pre-condition and the suffix property.
@@
 
-We then define pre- and post-conditions for function \prog{dfss}.
+We now define pre- and post-conditions for function \prog{dfss}.
 
@@
 For example, the following lemma asserts that the pre-condition of \prog{dfss} holds at the call from the body of \prog{dfs}.
-Although some of the well-formedness conditions for the augmented environment require a few lines of proof, the overall proof is straightforward.
+Although some of the well-formedness conditions for the augmented environment require a few lines of interaction, the overall proof is straightforward.
 
@@
 Since the two functions are mutually recursive, the two statements must be proved by simultaneous computational induction, using the corresponding proof rule generated by Isabelle.
-The proofs of these implications again involve a certain number of steps requiring detailed interaction.
+The proofs of these implications again involve a certain number of steps requiring substantial interaction.
 
@@
 If $y \in \prog{explored}~e'$, this is clearly true.
-Otherwise, we obtain some node $n \in \prog{set}~(\prog{stack}~e')$ such that $y \in \SS~e'~n$.
+Otherwise, given that $y \in \prog{visited}~e'$, we obtain some node $n \in \prog{set}~(\prog{stack}~e')$ such that $y \in \SS~e'~n$.
 If $n=v$, the assertion follows by the definition of $e''$.
@@
 The lemma is proved by embedding the argument tuples into triples $(a,b,c)$ where $a$ is the set of unvisited vertices, $b$ is the set of edges between vertices that have not been followed, and $c$ is $0$ if the triple represents an argument of \prog{dfs} and $1$ if it is an argument of \prog{dfss}.
-These triples can be ordered by the strict subset order for the two first component and the predecessor relation on natural numbers for the third one, and Isabelle proves automatically that the product of these orders is well-founded.
+These triples can be ordered by the strict subset order for the two first components and the predecessor relation on natural numbers for the third one, and Isabelle proves automatically that the product of these orders is well-founded.
 Finally, \prog{dfs-dfss-term} can be embedded as an inverse image in this well-founded relation, and is therefore itself well-founded.
+
+\begin{table*}
+  \centering
+  \begin{tabular}{|l|r|r|}
+    \hline
+    Category                         &  Definitions   &  Theorems and proofs\\
+    \hline\hline
+    Precedence on lists              &        2 loc   &         119 loc\\
+    \hline
+    Graphs and reachability          &       15 loc   &          64 loc\\
+    \hline
+    Formalization of the algorithm   &       64 loc   &           6 loc\\
+    \hline
+    Invariants, pre-/post-conditions &       68 loc   &         120 loc\\
+    \hline
+    Lemmas about \prog{unite}        &                &         663 loc\\
+    \hline
+    Partial correctness of \prog{dfs} and \prog{dfss} &  &     1575 loc\\
+    \hline
+    Termination of \prog{dfs} and \prog{dfss} & 13 loc   &      236 loc\\
+    \hline
+    Total correctness                &                &           5 loc\\
+    \hline\hline
+    Sum                              &      162 loc   &        2788 loc\\
+    \hline
+  \end{tabular}
+  \caption{Quantitative information about our formalization.}
+  \label{tab:effort}
+\end{table*}
 
@@
 
-\subsection{Proof effort}
+\subsection{Effort of formalization}
 \label{sec:effort}
 
-\textbf{Summarize the overall effort of the proof in number of lines (definitions / proofs) for each main category.}
+Table~\ref{tab:effort} summarizes the overall effort of the formalization.
+We report the number of lines (definitions and theorems/proofs) for each main category, ignoring comments and empty lines.
+It should be noted that our proofs are written in the relatively verbose Isar language, which favors readability and maintainability compared to shorter, but unreadable tactic applications.
+
+We start by defining the precedence relation between elements of a list, which does not exist in Isabelle's standard library, and proving elementary lemmas about this relation, as described in Section~\ref{sec:isabelle}.
+In particular, this relation is a partial order if the list does not contain duplicates, which is the case for the stacks in our formalization.
+
+The second category concerns the formalization of graphs as a locale, the definitions of reachability and (partial and maximal) SCCs, and fundamental lemmas about these concepts, cf.\ Section~\ref{sec:graphs}.
+
+Next, we define the algorithm, including the ``environment'' data structure and the functions \prog{unite}, \prog{dfs}, and \prog{dfss}, cf.\ Section~\ref{sec:formalization}.
+There is little to prove in this category of the formalization, except for the (trivial) exhaustiveness of the patterns in the recursive definitions, and the transitivity of the relation \prog{sub-env}.
+
+We now define the main invariant \prog{wf-env} and the pre- and post-conditions of the functions \prog{dfs} and \prog{dfss}, and we prove some immediate consequences of these definitions as lemmas for later proofs.
+The main results of this category are reported in Sections~\ref{sec:invariant} and~\ref{sec:pre-post}.
+
+The following two categories represent the main proof effort of this work.
+We start by proving auxiliary lemmas about the function \prog{unite}, and in particular the fact that it preserves the main invariant.
+The partial correctness proofs establish the pre-conditions for each recursive function call, as well as the post-conditions of the functions, assuming their pre-conditions, as reported in Sections~\ref{sec:pre-pre} and~\ref{sec:pre-post}.
+
+We finish the proof by showing that the pre-conditions defined earlier ensure the termination of the functions, and finally conclude total correctness of the algorithm, as explained in Section~\ref{sec:termination}.
 
@@
 
-We are not the first who prove the correctness of an algorithm for computing the SCCs of a graph using a proof assistant.
+As already noted in the introduction, we are not the first who prove the correctness of an algorithm for computing the SCCs of a graph using a proof assistant.
 Peter Lammich~\cite{lammich:gabow} used Isabelle/HOL for proving the correctness of Gabow's SCC algorithm.
@@
 In this framework, algorithms are represented as monadic functions, and concrete data structures are introduced, for example for representing partial SCCs.
-Our proof is based on a more abstract representation of the algorithm and its data structures, which leads to higher-level invariants and reduces proof effort.
+In a separate work, Lammich and Neumann contributed significantly to the development of a fully verified executable model checker~\cite{esparza:cava} by developing a framework for verifying depth-first search algorithms~\cite{lammich:depth-first}, including Tarjan's algorithm.
+In contrast to these works, our proof is based on a more abstract representation of Bloemen's algorithm and its data structures, which leads to higher-level invariants.
 Indeed, our main objective is not to obtain an efficient sequential implementation of Bloemen's algorithm but rather to pave the way for a verification of the parallel version of that algorithm.
@@
 Chen et al.~\cite{chen:tarjan} present a comparison of the proofs of Tarjan's algorithm for computing the SCCs of a graph in the proof assistants Coq, Isabelle, and Why3.
-The algorithm is represented at a similar level of abstraction as in our proof, leading to similar proof effort.
+The algorithm is represented at a similar level of abstraction as in our proof, leading to comparable proof effort.
 However, although Tarjan's algorithm is also based on depth-first search, its algorithmic structure is quite different because no partial SCCs are computed along the way, and a maximal SCC is popped from the stack only when a node is determined to be an SCC root at the end of the function call.
@@
 
-\begin{itemize}
-\item successfully proved correctness of an efficient sequential SCC algorithm
-\item compare to Tarjan proof (Chen et al.): both are DFS algorithms, the algorithm considered here collapses loops as soon as they are found, minimizing stack size
-\item compare to Lammich's work: he exploits the refinement framework for deriving executable code, we work on a higher-level representation of the algorithm as elementary recursive functions, hence invariants are simpler
-\item objective: stepping stone towards a proof of the parallel version of the algorithm from Bloemen's thesis -- understand which invariants can be relaxed.
-\end{itemize}
+In this paper, we reported on the successful proof in Isabelle/HOL of an efficient sequential SCC algorithm proposed by Bloemen et al.~\cite{bloemen:strong}.
+In particular, we exhibited the inductive invariant \prog{wf-env} that is preserved throughout the execution of the algorithm, as well as pre- and post-conditions for the two mutually recursive functions representing the algorithm.
+Although our proof is based on arguments given in the correctness proof in Bloemen's thesis, we had to add non-trivial predicates in order to obtain an inductive invariant.
+For example, the paper-and-pencil proof does not mention the facts that the partial SCCs maintained by the algorithm are maximal w.r.t.\ the current state of knowledge, or that the root of an SCC is the oldest node in depth-first order, but these are necessary for the argument to be conclusive.
+
+As we mentioned previously, the main objective of our work is to serve as a first step towards a machine-checked proof of a parallel implementation of Bloemen's algorithm on a multi-core processor.
+We therefore formulated our correctness predicates at a high level of abstraction, without considering how the data manipulated by the algorithm would actually be represented in an implementation.
+We hope that in this way, it will be possible to consider which relaxations of our invariants are possible in a parallel execution of the algorithm.
+Preliminary experiments with encoding the parallel version in TLA$^+$ and checking several plausible invariants of the algorithm over small graphs using model checking indicate that some invariants break without corrupting the overall result of the SCC computation~\cite{vandepol:exploring}, but that slight modifications of the algorithm are incorrect.
+We believe that the invariants that we identified for the sequential algorithm, together with the detailed proof, will provide a sound basis for extending the proof to the parallel version.
+
+% As a result of this whole year of work, we successfully proved correctness of an efficient sequential set-based SCC algorithm.
+In comparison to Tarjan's algorithm, a formal proof of which was written by Chen et.
+al \cite{chen:tarjan}, this algorithm collapses loops as soon as they are found, thus minimizing the stack size.
+Our formalization and  proof in Isabelle are quite similar to theirs as they are based on higher-level representations of the algorithms as elementary recursive functions.
+Hence, our invariants are quite simple compared to those expressed in Lammich's work \cite{lammich:gabow} since he exploits the refinement framework for deriving executable code, resulting in implementable-like representations that make expressing invariants more difficult from a higher-level point of view.
+
+% \begin{itemize}
+% \item successfully proved correctness of an efficient sequential SCC algorithm
+% \item compare to Tarjan proof (Chen et al.): both are DFS algorithms, the algorithm considered here collapses loops as soon as they are found, minimizing stack size
+% \item compare to Lammich's work: he exploits the refinement framework for deriving executable code, we work on a higher-level representation of the algorithm as elementary recursive functions, hence invariants are simpler
+% \item objective: stepping stone towards a proof of the parallel version of the algorithm from Bloemen's thesis -- understand which invariants can be relaxed.
+% \end{itemize}
 

### 2022-09-22 cpp/submission.tex
submitted CPP paper

@@
 It performs a depth-first traversal of the nodes of the graph.
-When it encounters an edge that leads to an already visited node that is part of the current DFS path, it detects a cycle, and it tracks (via additional information stored for the encountered nodes) which nodes belong to the same SCC.
-Dijkstra~\cite{dijkstra:finding} suggested a variant of a DFS traversal where the stack contains roots of (partial) SCCs.
+When it encounters an edge that leads to an already visited node that is part of the current DFS path, it detects a cycle, and it tracks
+% (via additional information stored for the encountered nodes)
+which nodes belong to the same SCC.
+Dijkstra~\cite{dijkstra:finding} suggested an algorithm that is also based on a DFS traversal of the graph.
+In this algorithm, the stack contains roots of (partial) SCCs.
 When a cycle is detected, the corresponding partial SCCs are collapsed, shortening the stack.
-Both algorithms are efficient because they operate in time $O(|V|+|E|)$ where $V$ and $E$ denote the sets of nodes and edges of the graph, since every node and every edge is visited only once during the traversal, but they differ in their algorithmic structure and in the auxiliary information that must be stored.
+Both algorithms are efficient because they operate in time $O(|V|+|E|)$ where $V$ and $E$ denote the sets of nodes and edges of the graph, since every node and every edge is visited only once during the traversal, but they require different auxiliary information to be stored during execution.
 
-However, both these algorithms and their numerous variants are fundamentally sequential because their correctness crucially relies on the graph being traversed in DFS order.
-Bloemen et al.~\cite{bloemen:strong,bloemen:multi-core} introduced a variant of Dijkstra's algorithm for computing SCCs that can be implemented on a shared-memory multi-core processor, based on efficient data structures for (i) implementing a union-find structure representing partial SCCs and (ii) a cyclic list for iterating over states, that minimize locking.
-This algorithm has been implemented in the high-performance model checker LTSmin.
+Crucially, both these algorithms and their numerous variants are fundamentally sequential because their correctness relies on the graph being traversed in DFS order.
+Bloemen et al.~\cite{bloemen:strong,bloemen:multi-core} introduced a variant of Dijkstra's algorithm for computing SCCs that can be implemented on a shared-memory multi-core processor, and they designed efficient data structures for (i) implementing a union-find structure representing partial SCCs and (ii) a cyclic list for iterating over states, that minimize locking.
+Their algorithm has been implemented in the high-performance model checker LTSmin.
+They empirically demonstrate that the algorithm ensures good speed-up even for graphs that have few large SCCs, as is often the case in model checking.
 The correctness proofs in Bloemen's thesis are based on ordinary paper-and-pencil reasoning and therefore do not correspond to the same exacting standards as the machine-checked verification of model checking algorithms mentioned above.
-The problem is particularly delicate given that a multi-core implementation of the algorithm need not strictly adhere to a DFS traversal order, and any core may revisit nodes that have already been visited by other cores.
+The problem is particularly delicate given that a multi-core implementation of the algorithm need not strictly adhere to a DFS traversal order, and any core may revisit nodes that have already been visited by other cores, which might not only be an efficiency problem but also jeopardize correctness.
 
-The present work makes a first contribution towards a machine-checked proof of the SCC algorithm by Bloemen et al.\ by establishing the correctness of the sequential version of the algorithm.
+The present work makes a first contribution towards a machine-checked proof of Bloemen et al.'s SCC algorithm by establishing the correctness of the sequential version of the algorithm.
 Although this is arguably less challenging than the proof of the parallel version, we consider it to be a necessary first step that establishes a baseline against which the parallel algorithm can be compared.
@@
 \subsection{Lemmas establishing the pre-conditions}
-\label{sec:pre-pre}
+\label{sec:pre-implies-pre}
 
@@
 \subsection{Pre-conditions imply post-conditions}
-\label{sec:pre-post}
+\label{sec:pre-implies-post}
 
@@
 
-Crucially, we prove that the relation \prog{dfs-dfss-term} is well-founded.
-
-\begin{small}
-\begin{lstlisting}[language=isabelle]
-lemma wf-term: wf dfs-dfss-term
-\end{lstlisting}
-\end{small}
-
-In the statement of the lemma, \prog{wf} represents the well-foundedness predicate, which is predefined in the Isabelle standard library.
-The lemma is proved by embedding the argument tuples into triples $(a,b,c)$ where $a$ is the set of unvisited vertices, $b$ is the set of edges between vertices that have not been followed, and $c$ is $0$ if the triple represents an argument of \prog{dfs} and $1$ if it is an argument of \prog{dfss}.
-These triples can be ordered by the strict subset order for the two first components and the predecessor relation on natural numbers for the third one, and Isabelle proves automatically that the product of these orders is well-founded.
-Finally, \prog{dfs-dfss-term} can be embedded as an inverse image in this well-founded relation, and is therefore itself well-founded.
-
 \begin{table*}
@@
     \hline
-    Lemmas about \prog{unite}        &                &         663 loc\\
+    Lemmas about \prog{unite}        &          ---   &         663 loc\\
     \hline
-    Partial correctness of \prog{dfs} and \prog{dfss} &  &     1575 loc\\
+    Partial correctness of \prog{dfs} and \prog{dfss} & --- &  1575 loc\\
     \hline
@@
     \hline
-    Total correctness                &                &           5 loc\\
+    Total correctness                &          ---   &           5 loc\\
     \hline\hline
@@
 \end{table*}
+
+Crucially, we prove that the relation \prog{dfs-dfss-term} is well-founded.
+
+\begin{small}
+\begin{lstlisting}[language=isabelle]
+lemma wf-term: wf dfs-dfss-term
+\end{lstlisting}
+\end{small}
+
+In the statement of the lemma, \prog{wf} represents the well-foundedness predicate, which is predefined in the Isabelle standard library.
+The lemma is proved by embedding the argument tuples into triples $(a,b,c)$ where $a$ is the set of unvisited vertices, $b$ is the set of edges between vertices that have not been followed, and $c$ is $0$ if the triple represents an argument of \prog{dfs} and $1$ if it is an argument of \prog{dfss}.
+These triples can be ordered by the strict subset order for the two first components and the predecessor relation on natural numbers for the third one, and Isabelle proves automatically that the product of these orders is well-founded.
+Finally, \prog{dfs-dfss-term} can be embedded as an inverse image in this well-founded relation, and is therefore itself well-founded.
 
@@
 We report the number of lines (definitions and theorems/proofs) for each main category, ignoring comments and empty lines.
-It should be noted that our proofs are written in the relatively verbose Isar language, which favors readability and maintainability compared to shorter, but unreadable tactic applications.
+It should be noted that our proofs are written in the relatively verbose Isar language, which favors readability and maintainability compared to shorter, low-level tactic applications.
 
-We start by defining the precedence relation between elements of a list, which does not exist in Isabelle's standard library, and proving elementary lemmas about this relation, as described in Section~\ref{sec:isabelle}.
+We start by defining the precedence relation between elements of a list, which is not pre-defined in Isabelle's standard library, and proving elementary lemmas about this relation, as described in Section~\ref{sec:isabelle}.
 In particular, this relation is a partial order if the list does not contain duplicates, which is the case for the stacks in our formalization.
@@
 Next, we define the algorithm, including the ``environment'' data structure and the functions \prog{unite}, \prog{dfs}, and \prog{dfss}, cf.\ Section~\ref{sec:formalization}.
-There is little to prove in this category of the formalization, except for the (trivial) exhaustiveness of the patterns in the recursive definitions, and the transitivity of the relation \prog{sub-env}.
+This category of our formalization mainly consists of definitions and few proofs, except for the (trivial) exhaustiveness of the patterns in the recursive definitions, and the transitivity of the relation \prog{sub-env}.
 
@@
 We start by proving auxiliary lemmas about the function \prog{unite}, and in particular the fact that it preserves the main invariant.
-The partial correctness proofs establish the pre-conditions for each recursive function call, as well as the post-conditions of the functions, assuming their pre-conditions, as reported in Sections~\ref{sec:pre-pre} and~\ref{sec:pre-post}.
+The partial correctness proofs establish the pre-conditions for each recursive function call, as well as the post-conditions of these calls, assuming their pre-conditions, as reported in Sections~\ref{sec:pre-implies-pre} and~\ref{sec:pre-implies-post}.
 
@@
 
-As already noted in the introduction, we are not the first who prove the correctness of an algorithm for computing the SCCs of a graph using a proof assistant.
+As we already noted in the introduction, we are not the first who prove the correctness of an algorithm for computing the SCCs of a graph using a proof assistant.
 Peter Lammich~\cite{lammich:gabow} used Isabelle/HOL for proving the correctness of Gabow's SCC algorithm.
-Like Bloemen's algorithm, Gabow's algorithm applies depth-first search and augments partial SCCs whenever a back loop to an already visited vertex of the graph is found.
-Lammich's proof focuses on efficient code generation, based on the use of his refinement framework~\cite{lammich:refinement} formalized in Isabelle.
+Like Bloemen's algorithm, Gabow's algorithm is based on Dijkstra's algorithm: it applies depth-first search and combines partial SCCs whenever a back loop to an already visited vertex of the graph is found.
+Lammich's development focuses on efficient code generation, based on the use of his refinement framework~\cite{lammich:refinement} formalized in Isabelle.
 In this framework, algorithms are represented as monadic functions, and concrete data structures are introduced, for example for representing partial SCCs.
-In a separate work, Lammich and Neumann contributed significantly to the development of a fully verified executable model checker~\cite{esparza:cava} by developing a framework for verifying depth-first search algorithms~\cite{lammich:depth-first}, including Tarjan's algorithm.
+In separate work, Lammich and Neumann contributed significantly to the development of a fully verified executable model checker~\cite{esparza:cava} by developing a framework for verifying depth-first search algorithms~\cite{lammich:depth-first}, including Tarjan's algorithm.
 In contrast to these works, our proof is based on a more abstract representation of Bloemen's algorithm and its data structures, which leads to higher-level invariants.
@@
 The algorithm is represented at a similar level of abstraction as in our proof, leading to comparable proof effort.
-However, although Tarjan's algorithm is also based on depth-first search, its algorithmic structure is quite different because no partial SCCs are computed along the way, and a maximal SCC is popped from the stack only when a node is determined to be an SCC root at the end of the function call.
+However, although Tarjan's algorithm is also based on depth-first search, its algorithmic structure is quite different because no partial SCCs are computed along the way, and a maximal SCC is popped from the stack only when a node is determined to be an SCC root at the end of the DFS search starting at that node.
 The primary focus of the presentation in~\cite{chen:tarjan} is on a comparison between the styles of proof and the overall effort required in the three proof assistants.
@@
 
-In this paper, we reported on the successful proof in Isabelle/HOL of an efficient sequential SCC algorithm proposed by Bloemen et al.~\cite{bloemen:strong}.
+In this paper, we reported on the successful proof in Isabelle/HOL of an efficient sequential SCC algorithm proposed by Bloemen et al.~\cite{bloemen:strong,bloemen:multi-core}.
 In particular, we exhibited the inductive invariant \prog{wf-env} that is preserved throughout the execution of the algorithm, as well as pre- and post-conditions for the two mutually recursive functions representing the algorithm.
-Although our proof is based on arguments given in the correctness proof in Bloemen's thesis, we had to add non-trivial predicates in order to obtain an inductive invariant.
+Although our proof is based on arguments given in Bloemen's thesis, we had to add non-trivial predicates in order to obtain an inductive invariant.
 For example, the paper-and-pencil proof does not mention the facts that the partial SCCs maintained by the algorithm are maximal w.r.t.\ the current state of knowledge, or that the root of an SCC is the oldest node in depth-first order, but these are necessary for the argument to be conclusive.
 
-As we mentioned previously, the main objective of our work is to serve as a first step towards a machine-checked proof of a parallel implementation of Bloemen's algorithm on a multi-core processor.
-We therefore formulated our correctness predicates at a high level of abstraction, without considering how the data manipulated by the algorithm would actually be represented in an implementation.
-We hope that in this way, it will be possible to consider which relaxations of our invariants are possible in a parallel execution of the algorithm.
-Preliminary experiments with encoding the parallel version in TLA$^+$ and checking several plausible invariants of the algorithm over small graphs using model checking indicate that some invariants break without corrupting the overall result of the SCC computation~\cite{vandepol:exploring}, but that slight modifications of the algorithm are incorrect.
+As we mentioned previously, the main objective of our work is to serve as a first step towards a machine-checked proof of a parallel version of Bloemen's algorithm that can be implemented on a multi-core processor.
+We therefore formulated our correctness predicates at a high level of abstraction, without considering how the data manipulated by the algorithm would actually be represented in a program.
+We hope that this will help considering which relaxations of our invariants are possible in a parallel execution of the algorithm.
+Preliminary experiments with encoding the parallel version in TLA$^+$ and checking several plausible invariants of the algorithm over small graphs using model checking indicate that some invariants break, but that the overall result of the SCC computation is not corrupted~\cite{vandepol:exploring}.
+However, apparently minor modifications of the algorithm were found to be incorrect.
 We believe that the invariants that we identified for the sequential algorithm, together with the detailed proof, will provide a sound basis for extending the proof to the parallel version.
-
-% As a result of this whole year of work, we successfully proved correctness of an efficient sequential set-based SCC algorithm.
-In comparison to Tarjan's algorithm, a formal proof of which was written by Chen et.
-al \cite{chen:tarjan}, this algorithm collapses loops as soon as they are found, thus minimizing the stack size.
-Our formalization and  proof in Isabelle are quite similar to theirs as they are based on higher-level representations of the algorithms as elementary recursive functions.
-Hence, our invariants are quite simple compared to those expressed in Lammich's work \cite{lammich:gabow} since he exploits the refinement framework for deriving executable code, resulting in implementable-like representations that make expressing invariants more difficult from a higher-level point of view.
-
-% \begin{itemize}
-% \item successfully proved correctness of an efficient sequential SCC algorithm
-% \item compare to Tarjan proof (Chen et al.): both are DFS algorithms, the algorithm considered here collapses loops as soon as they are found, minimizing stack size
-% \item compare to Lammich's work: he exploits the refinement framework for deriving executable code, we work on a higher-level representation of the algorithm as elementary recursive functions, hence invariants are simpler
-% \item objective: stepping stone towards a proof of the parallel version of the algorithm from Bloemen's thesis -- understand which invariants can be relaxed.
-% \end{itemize}
 
